
\textbf{8 At Lahu Hunting Expedition}

[Tape I, Side 2 \#3]

1: Well, where shall we go hunting today?

2: Well-uh-out by Deer Hollow.

\begin{center}
* * *
\end{center}

\leftskip=0pt
3: Well then, where shall all of us in the village go hunting today anyway?

4: Well-uh-at the Metalai [stream]. From above Red Rock--I mean, from \textit{below}
Red Rock on down. We'll chase them down from above Metalai to Hognose Deer, the
place where Lawjo the Postman\footnote{This man was a friend of the Huey Tat villagers who used to be a postman in Shan State.} shot the mountain-goat that time.

5: Mm.

6: Where shall each of us go to lie in ambush\footnote{nâʔ ve 1. 'to sharpen' ['to make noise or beat the undergrowth so that the game will emerge; to flush game'] 2. 'to track down an animal' 3. 'to lie in ambush for game.' One hunting party typically flushes the game while the others lie in ambush at various places (nâʔ), waiting for a chance to chase it (g̈àʔ ve).}?

7: Me, I'll go wait at the hill near Hognose Deer River.\footnote{This unlovely name is a literal rendering of khɨ́-yɨ̀-kwì lɔ̀-qÁ.}

8: I'll go wait at goat-shoot.\footnote{hâ bɔ̂ʔ kt̀: the scene of Lawju's exploit.}

9: Down there, eh? I'm going to wait at the big stamping-ground right at Hognose
Deer River.

10: I see.

11: Well, as far as I'm concerned--I'll go wait at goat-shoot.

12: Hm. Well, you do a good piece of hunting now!

13: \{Yipee! / Yes, yes!\} Today I'm going hunting! Let's track 'em the right way!

14: Right!

15: Well, now we're off to the hunt!

16: (Two voices) Right!

\begin{center}
* * *
\end{center}

\leftskip=0pt
17: Hey, hey, hey, hey, hey, hey, hey, hey, hey!  Cho, cho, cho, cho, cho, cho,
cho! Oh, we've flushed a wild pig! We've flushed a boar, we've flushed a boar!
Come on, let's track him right! Everybody at his ambush-point, track him down right!
We've flushed a boar! Ah, here come the dogs to help us look, too!

18: Woof, woof, woof, woof, woof, woof, woof, woof, woof!

19: Here he goes, there he goes, there he goes! Cho, cho!

[A shot rings out]

20: What happened? Did they get him?

21:  Woof, woof, woof, woof, woof, woof, woof, woof, woof, woof!

22: Well, I think we hit him at least. But the bullet wasn't big [enough]! And
\textit{his} gun jammed, damn it.

23: The bullet drew a little blood. Let's go-o-o! Try to stick right on his trail,
try to stick right on his trail. Come on!

24: The gun-setting probably wasn't big [enough]--on \textit{your} gun.

25: Er--about my gun-setting--it was set for the squirrel-shooting that I did a
while back, that's why.

26: Oh, no! We won't get anything to eat then. Here we've been hunting all day
long today, but you guys set your guns so small, everything today is a goddamn
mess! We won't get a thing to eat. Try to stick to his trail at least! It's a bad
business.\footnote{chə̀ ve 'to be all right; to work; be O.K.' chə̀ tù mâ hêʔ "it's no good."}

27: we won't get a thing to eat. We won't get a thing to eat! You failure! You
ass! You rotter! You dunce!

28: Ha, ha, ha.

29: Setting his gun that way!

30: Hey, down there on the hill near the river the dogs are quiet all of a sudden!

31: Hurry, let's go look! Hurry up, let's go take a look!

32: The dogs are quiet! Hurry come look, hurry come look!

33: Down there, down below; the dogs are quiet all of a sudden!

34: Well, let's go have a look.

35: What's going on?

36: Oh! It dropped dead here! It died here!

37: Shout to them! To everybody who's still up there on the mountain.

38: Hey, come on! Get down there, I say\footnote{The quotative Pv cê may be used in impatient commands: "\emph{V}, I say!" cf. all the emphatic Pvf's built on the verb gôʔ 'say': gôʔ-ma, gôʔ-nè etc.}! The boar dropped dead here! Come here,
I tell you! Come quick!

39: Chop down some leaves, chop down some leaves! We'll divide him up right away.\footnote{The meat will be cut up and divided, each portion being wrapped up in a large banana-leaf.}

40: Divvy up, divvy up, divvy up--let's divide him up!

41: Divvy up, divvy up, divvy up!

42: Keep an eye on the dogs. They'll eat the meat.

43: O.K.

[Sounds of slicing up the boar]

44: That dog of yours might get at the meat.\footnote{pt́ à is used in warnings. cf. English 'may well \emph{V} (if you don't watch out).'}

45: Scram, scram! For god's sakes, this bunch of dogs--they're acting like they
caught him all by themselves!

46: Keep watching the dogs! They're taking up pieces [in their mouths]. Look here!

47: Scra-a-m!

48: How many of us are there?

49: Count us and see.

50: Eight people.

51: Eight people, eight shares... And take out the "chief-slices"
carefully.

52: This one here--cut me one of the shank-joints.\footnote{A bony cut, consisting of the shin, knee, and part of the lower thigh.}

53: What are you going to do with it?

54: Boil the life out of it and eat it.\footnote{cɔ̂ ve: 'to boil [sth tough] for hours until very soft' vs. cɔ́ ve 'to boil' (q.v.)}

55: Oh. Ha, ha, ha.

56: He says to cut him one of the shanks. He says he'll boil it to death and eat
it.

57: I cut it for him already. I cut it already.

58: Well, it's divided, it's divided, it's divided now.

59: There are eight people?

60: Yes.

61: Well, he's finished dividing it up. Now take \{your share / them\}, \{your
share, them\}.

62: Take 'em, take 'em, take 'em, take 'em!

63: Take the biggest share, why don't you! He took the biggest--.

64: Yes, I want this piece.

65: I want this.

66: I'll take out this share.

67: Well, I'll take this.

68: Scoop up this blood carefully and feed it to the dogs first.

69: Don't go feeding them, don't go feeding them, dogs like this aren't worth a
damn.

70: Because it was no thanks to them\footnote{câʔ ve 'to be connected with; be due to.' cf. the transitive counterpart cÁ ve 'to connect sth, join sth together.'} that we caught anything, was it?

71: Yeah.

72: Well, now let's just go hunt back towards home, shall we? Today--.

73: \{Well-- / Yep\}.

74: We'll just head back. We've had good hunting.\footnote{g̈a lɛ̀ʔ mâ lɛ: "since we have gotten a lot to eat."} But we'll just head back
and see what we can see. Since there's nothing [else] we can do now.

75: I guess that's the end.

