\setcounter{footnote}{0}


1. Let us reflect together on the story of a group of rich men who made offerings
to God and of a certain widow who offered money to God. 2 This story which I would
like to impart to you, my brothers, is to be found in Mark 12:41-44.1 3 A second
passage is to be found in Luke 6:38.2 4 And a third passage is contained in II
Corinthians 9:6-7.3

5. This book of the Bible4 has many \{places / passages\} which are full of meaning.5
6. Therefore, please listen carefully to this story about offering money to God
in the temple.

7. When Jesus was living on the earth, there was a certain very poor widow in the
city of Jerusalem. 8 Her husband had died long before. 9 She had to seek her own
living, and though she did her utmost, [her] money was never enough to gain her
food and drink [or] to clothe and garb herself, so she would go hungry and suffer.

10. This woman, despite the fact that she was so wretched, had not forgotten God.
She would go up to the temple6 as was fitting,7 and not only did she praise Him,
but every day she would pray8 fervently7 to God.

11. "One day this widow went to the great Temple to contribute and donate
an offering of money. 12 For her offering she gave a great deal of money--er, rather,
she \textit{could not} give a great deal of money, but she took all the money she
did have and went to offer it--went to offer it to Him. 13 When she prayed--er,
when she arrived at the Temple, she saw a group of rich men9 putting great sums
of money into the offering-box, and her heart was downcast. 14 She wanted to offer
lots of money just like them, but she didn't have anything to give. 15 She only
had two small coins.10 16 Those two small coins were only worth a farthing a piece.11

17. When the woman put the two small coins in the offering-box--into the offering-box--Jesus
saw it. 18 He also saw the group of rich men offering great sums of money. 19 Then
he said to his disciples: 20 "More than all the people who have put [money]
into the offering-place--er, into the offering-box--this poor and wretched widow
has put the most of all."

