
\textbf{14 Sermon: The Widow's Mites}

21 When Jesus was living upon the earth, in the city of Jerusalem, the great Temple--in
the great Temple in the city of Jerusalem, all the Pharisees who had much earthly
wealth12, and all the High Priest Pharisees, and all the rich people and tycoons13
offered great sums of money to God--were offering [great sums of money to God].
22 At this time there was a very poor and wretched woman. 23 Since her husband
had died, she was obliged to earn her living all by herself: everything, whether
it was her money and effects, or her clothing and raiment, or her food and drink.
24 [That's how she was faring] at that time. 25 Therefore, when the moment for
putting money into the offering-place came, she had nothing to offer. 26 She wanted
to make an offering to God. 27 She had nothing to offer. 28 The other people, all
the tycoons coming up [to the temple] in their horse-drawn wagons pouring money
into the offering-box until it overflowed--this woman saw them. 29 When she saw
this, how could she feel?

30 \texttt{"}These others can offer so very, very much to God. 31 I cannot offer
God anything. 32 I can only give this tiny amount, and my heart is sore indeed.\texttt{"}
33 This is what the woman was thinking. 34 As she thought this she was filled with
embarrassment. 35 She had no idea14 how she could go and put anything into the
offering-place. 36 She had no idea how she could go put anything into the offering-box.
37 There was a great throng [there], thousands and tens of thousands of people.
38 Slowly she carried her money, hiding it [in her hand], and went to put it into
the offering-place. 39 How much was it that she put in? 40 Only two farthings.
41 She only had two farthings. 42 For herself. 43 To put in there.

44 Jesus saw her doing this. 45 As Jesus was sitting there he saw it. 46 What he
said was this: \texttt{"}As far as you can see15, who is the single person who
has contributed the very most to this money that has been offered to God today?
47 Who has put in more than anybody else?\texttt{"}

48 What did the disciples say? 49 This is how the disciples saw things: 50 All
the disciples had their attention fixed on a certain person who had taken an enormous
amount of money and put it into God's offering-place, and they thought \texttt{"}That
is a tremendous amount! 51 [But] what is that woman doing giving a couple of farthings
to God--putting them into His offering-place!\texttt{"} 52 This is what the disciples
were thinking.

53 What did Jesus say? 54 \texttt{"}So16 that's what you think,\texttt{"} he said.
55 \texttt{"}Those people put in a lot of money. 56 That woman put in a little
bit of money. 57 As far as her working for God's sake17 goes, she has offered God
very little: that's what you all think. 58 But this woman does not have a single
farthing's worth of substance18 to keep herself alive. 59 She only had two farthings.
60 She took it all and offered it to God. 61 The one who is the greatest of all
these people, the one who has put in the largest offering, is this woman:\texttt{"}
thus did Jesus speak to his disciples.

62 This means19 that among you Christians, when it is time to make an offering
to God, there should be no embarrassment [such that you say to yourselves]: \texttt{"}I
have so little that I cannot offer God anything.\texttt{"} 63 We ought to make
offerings to God in accordance with our means.19

64 Oh, my brothers! Let everyone give to God to the best of his ability20! 65 Because,
you see, your body, your intelligence, your wealth and property, your pigs and
chickens, all of there things belong to God. 66 \texttt{"}My body does not belong
to me. 67 Everything belongs to God. 68 For this reason we should make offerings
to God. 69 For this reason, when we spend our money, when we give it away on the
outside21 to the people of the world22, to those who would teach us things, we
should make sure23 that this money is doing work for God. 70 Therefore, my Christian
brethren, may the doing of the Lord's work remain always in our hearts, as it should.

71 This is what I have to impart to you.

