\setcounter{footnote}{0}


1. This spirit, you see, when this master-spirit is hungry for meat, he takes a
pitchfork\footnote{\textbf{šo} \textbf{qɔ̀ʔ} \textbf{lɛ̀ʔ} \textbf{ve}: lit. ``a crooked piece of iron.'' \textbf{lɛ̀ʔ} \textasciitilde{} \textbf{ɛ̀ʔ} `adverbializer'. See the \textit{Index Verborum} in GL:647.} and hooks it fast onto a human being, [so that] if he [=the human]
doesn't feed him tasty food, he will fiercely make the human being waste away.\footnote{\textbf{cɔ́} \textbf{cɨ} \textbf{ve} `cause to be thin.'}

2. But if the man goes to consult the spirit-doctor\footnote{\textbf{mɔ́-pa}: a ``ritual specialist'' in animist villages, often an opium addict, who is recognized as an authority in dealing with the spirits by reciting the appropriate incantations. See the many such texts recorded by A.R. Walker (partial bibliography in DL:1421-25).} the doctor [will say]:

3. ``The reason why your body has become so thin is, that the sky-spirit
wanted to eat delicious food through you, so he has treated you this way.

4. ``However, if you cook up some good-tasting food and feed it to him,
you will be cured.

5. ``If you don't feed him, that pitchfork will stay hooked onto you.

6. But\footnote{This speaker uses \textbf{yàʔ-qhâ(-kàʔ}) `but; however' to the exclusion of all other conjunctions. It is left untranslated occasionally, where its iteration would be tedious in English.} as soon as you do feed him, he'll take it all away, and your body will
be healthy [again]."

7. This kind of spirit can also start to cause trouble\footnote{\textbf{chèʔ}: lit. ``bite''.} for the whole household.

8. However, as far as the kind [of spirit] that cannot bite goes, we simply pay
homage to

them, mark you!\footnote{\textbf{nê} \textbf{te} \textbf{ve} `to worship the spirits.' \textbf{qôʔ-pî-ʔ} is a final unrestricted particle meaning literally ``tell it to third persons!'', here translated `mark you'.}

9. We do not worship the kind that can bite.

10. If we were to speak about what forms\footnote{\textbf{ɔ̀-to}: lit. ``bodies'' (< Shan).} the biting kind of spirits have, seeing
as how there are so many of them, since they are everywhere--water-spirits, rock-spirits,
mountain-spirits, valley-spirits, tree-spirits, thunder-and-lightning-spirits--there
are all kinds of spirits.

11. But even though there is such a large number of them, this kind of spirits
that can bite are all inferior to us human beings.

12. This is the way the matter stands.

