
{\textbf{20 How We Came from Burma}}

{\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}
HEAD=======[Tape II, Side 1 \#11]}

{1 [When we were] fleeing here to Thailand [from where]1 we [used to] live
in Burma, when we were fleeing here along the roads and byways,2 we lived off jungle-greens3
and by damming [streams] for fish as we came--that's how we made it here. 2 And
there was a big flock of kids along, too--er, walking along--. 3 And we had to
stay on the road a very long time, too. 4 But4 getting here--uh, when it had gotten
to be over a month, we had come as far as Thayakwei,5 where we made camp.6 5 Looking
for banana-leaves6 to live off and damming forks in streams, that's how we got
here. 6 We made it to Thailand.}

{[7 (whisper) Think carefully before you speak!]7}

{8 Fleeing here from Burma we [crossed] the Nam-Si?8 over to this side.
9 The time we stayed at Thayakwei was about two months--we were there for two months.
10 When we were there--}

{[11 (whisper) Don't laugh!]9}

{12 --we were all suffering from hunger and cold. 13 We would go and eat
the other Lahu's mustard-greens.10 14 We'd go and eat their rice. 15 We were at
Thayakwei for two months. 16 We would chop down Kitty-cane11 to eat. 18 We'd go
eat banana-\{stems / plants\}. 19 We'd dam for fish to eat. 20 We came here begging
and escaping for our food.12}

{21 After staying there for a long time13 we continued on to Shajieh.14
22 When we arrived at Shajieh we and the Chinese--er, all he Chinese welcomed us
gladly and cheerfully, and set down cups of hot water15 for us to drink. 23 And
they fed everybody rice, too. 24 After we ate we took to the road16 again, until
we got to the village of Po-mu17 \{up / over\} there, where we stayed and spent
the night, and we Lahu traded all sorts of things with them for food.18 25 We traded
away our silver buttons,19 and even our bracelets we traded away to fill our bellies.20
26 We had nothing [left]. 27 In poverty and distress we were begging and screaming
for our food.}

{28 Then22 our brethren who were [already] living here in Huey Tad21 came
to lead us here. 29 They came for us. 30 THey came to lead us here and they took
care of us joyfully and gladly, and so we arrived here at Huey Tad. }

{31 Having to flee here was a very distressing time for us.23 32 On the
midst of our trouble we had been separated from one another.24 33 So inasmuch25
as we have [finally], by the grace of God, managed to see each other again, we
now have great and abundant reason to praise God. }

{34 That's all there is [to say].}

{\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}
516355fe221b8becf888516aedf3d8b9e942cb6e}

{1 When we were fleeing here to Thailand from where[1] we used to live in
Burma, when we were fleeing here along the roads and byways, we lived off jungle-greens[2]
and by damming streams for fish as we came -- that's how we made it here. }

{2 And there was a big flock of kids along, too -- er, walking along --.
}

{3 And we had to stay on the road a very long time, too. }

{4 Getting here[3], when it had gotten to be over a month, we had come as
far as Thayakwei,[4] where we made camp.}

{5 Looking for banana tree creepers[5] to live off and damming forks in
streams, that's how we got here. }

{6 We made it to Thailand.}

{7 [whisper] Think carefully before you speak![6]}

{8 Fleeing here from Burma we crossed the Nam-Si river[7] over to this side.
}

{9 The time we stayed at Thayakwei was about two months -- we were there
for two months. }

{10 When we were there --}

{11 [whisper] Don't laugh![8]}

{12 -- we were all suffering from hunger and cold. }

{13 We would go and eat the other Lahus' mustard-greens.[9]}

{14 We'd go and eat their rice. }

{15 We were at Thayakwei for two months. }

{16 We would chop down sticky rattan[10] to eat. }

{18 We'd go eat banana plants. }

{19 We'd dam for fish to eat. }

{20 We came here begging for something to eat.[11]}

{21 After[12] staying there for a long time[13]}{\textsuperscript{
we continued on to Shajieh.[14] }}

{\textsuperscript{22 When we arrived at Shajieh we and the Chinese -- er,
all the Chinese welcomed us gladly and cheerfully, and set down cups of tea[15]
for us to drink. }}

{\textsuperscript{23 And they fed everybody rice, too. }}

{\textsuperscript{24 After we ate we took to the road[16] again, until we
got to the village of Pomu[17] over there, where we stayed and spent the night,
and we Lahu traded all sorts of things with them for food.}}

{\textsuperscript{25 We traded away our silver buttons,[18] and even our
bracelets we traded away to fill our bellies.[19]}}

{\textsuperscript{26 We had nothing left. }}

{\textsuperscript{27 In poverty and distress we were begging for our food.}}

{\textsuperscript{28 Then our brethren who were already living here in Huey
Tat[20] came to lead us here. }}

{\textsuperscript{29 They came for us. }}

{\textsuperscript{30 They came to lead us here and they took care of us
joyfully and gladly, and so we arrived here at Huey Tat. }}

{\textsuperscript{31 Having to flee here was a very distressing time for
us.[21]}}

{\textsuperscript{32 In the midst of our troubles we had been separated
from one another.[22]}}

{\textsuperscript{33 So inasmuch[23] as we have finally, by the grace of
God, managed to see each other again, we now have great and abundant reason to
praise God. }}

{\textsuperscript{34 That's all there is to say.}}

