
\textbf{\#21 Lahu cosmogony}

1 Above the Precious Eternal Lake there was no abundance of trees to be found.

2 There was only the Tree of Eternal Life.

3 This Tree of Eternal Life split up and branched out, splitting up into branches
[which turned into] a crooked river with 99 bends, [but] there was no abundance
of trees to be found.

4 There were a few phɛ̀ʔ-qhwɛ̂ ferns \footnote{\textbf{phɛ̀ʔ-qhwɛ̂}: a kind of soft edible dark green leaf like spinach. The plant grows in high and damp places, and is about two feet tall.  See  DL: 913.}; these phɛ̀ʔ-qhwɛ̂ had pretty
leaves and were spread out on winnowing sieves, but there was no abundance of trees
to be found.

5 There were a few qā-ni trees \footnote{\textbf{qā-ni}: a huge tree indicating good land for all crops, used in shipbuilding [\textit{Duabanga grandiflora}]\textit{. }See DL: 236.} ; these qā-ni came into being, winnowing
sieves and winnowing trays came into being, but there was no abundance of trees
to be found.

6 These were a few aquatic \footnote{\textbf{g̈ɨ̀-yâ: } \textbf{g̈ɨ̀  }means both 'pertaining to God' (< PLB \textbf{*ray}) and 'water' (< PLB \textbf{*rəy}), so this compound means both 'lesser deity created by G'uisha' and 'water creature; aquatic creature'.} phɛ̀ʔ-qhwɛ̂ plants, and these aquatic phɛ̀ʔ-qhwɛ̂
plants had 99 branches.

7 However, one of the branches became a servant. One branch became a servant --
oh, one branch became a servant.

8 Another branch became a race of water creatures.

9 One branch became a servant --  oh, one branch became a servant.

10 Another branch became a race of snakes.

11 But there was no abundance of trees to be found.

12 The divine Tree of Eternal Life alone, this Tree split up nicely.

13 It split up very well, but when one branch fell down sambar deer came out.

14 When another branch fell, wild boars came out.

15 Another branch fell and human beings came out.

16 Thus all kinds came into being and emerged there.

