
{\LARGE{}27 In Unity there is Strength 3/9/65}

{\LARGE{}Sermon by Pastor Càbo}

{\LARGE{}1. Once upon a time there was an old man among us Lahu.}

{\LARGE{}2. His sons were seven in number.}

{\LARGE{}3. As for those seven sons of his, they couldn't get along with each other
at all, and were always just quarreling and bickering among themselves.}

{\LARGE{}4. Even when the father would try to give them advice, they would not
heed their father's words.}

{\LARGE{}5. One day the father said to them, ``Oh, my seven sons! [1] Each one
of you go get a stick and bring it to Father,'' he said.}

{\LARGE{}6. When they got them they gave them to their father.}

{\LARGE{}7. Their father tied up the sticks together, and first had the eldest
son bend them.}

{\LARGE{}8. Then he made the second son bend them.}

{\LARGE{}9. Then he had the third son bend them.}

{\LARGE{}10. Then he had the fourth one bend them.}

{\LARGE{}11. Then he had the fifth one bend them.}

{\LARGE{}12. Until the seventh one, one after the other he made them bend them.}

{\LARGE{}13. [But] since he had tied those sticks together, they could not break
them by bending them. [2]}

{\LARGE{}14. The father said to all his sons, ``Oh, my sons! If people love and
cherish each other it is just like this.}

{\LARGE{}15. ``Whether it be robbers and bandits, whether it's any kind
of thing that might ruin you, they can't destroy a single one of you,'' he said.}

{\LARGE{}16. Then the father untied and separated the sticks, and gave each person
one stick.}

{\LARGE{}17. When he had given each of them a stick, and again had them bend them,
every stick broke.}

{\LARGE{}18. The father instructed them and said, ``Oh, seven sons of mine, if
you separate and disperse, if you don't get along harmoniously, and if you don't
respect one another, if you each go your own way, it will be just like this.}

{\LARGE{}19. If things that might destroy you should befall you, others can always
help you,'' he said, instructing his sons.}

{\LARGE{}20. So if all of us Lahu of today's generation take this story to heart,
it would be good.}

{\LARGE{}[1] Note the use of the bound morpheme }{\LARGE{}\textbf{-hɨ}}{\LARGE{}
`pluralizer' after the quantified noun-phrase }{\LARGE{}\textbf{yâ-pā šɨ̄
g̈â}}{\LARGE{} `seven sons'.}

{\LARGE{}[2] }{\LARGE{}\textbf{thɛ}}{\LARGE{} here functions as a }{\LARGE{}\textit{resultative
complement}}{\LARGE{}. See GL:207-209     }


