\setcounter{footnote}{0}

or: The Advantages of Faith in Fortune-tellers

or: The Power of Positive Thinking

Huey Tat 3/9/65

1. Once upon a time there were two people who were friends.

2. One of them was blind, and the other was a cripple.\footnote{\textbf{khɨ-qɔ̀ʔ-pā}: lit. ``one with a crooked leg''. Later in the text the speaker uses the roughly synonymous word \textbf{khɨ-lù-pā}, lit. ``one with a ruined leg''.}

3. While the two of them were friends, one day they had an idea, and they both
went together to a fortune-teller\footnote{\textbf{mɔ́-pā}: `medicine man; fortune-teller; seer; magician; spirit-doctor; shaman' (DL:1029)} to ask how his eyes could come to see and
how his leg could become straight.

4. When the two of them went to the fortune-teller, the fortune-teller said to
them, ``On your way going back, above you on the road when you see a bird's nest
in the crotch of a tree, make the blind man climb up and take it,'' he said.

5. Just exactly as the fortune-teller had said, as the two of them were going on
their way, when the cripple looked upwards, he saw a bird's nest.

6. Since he caught side of the bird's nest, he told the blind man to climb up.

7. The blind man said, ``Since my eyes don't see, I can't see all these things.

8. ``I can't climb up,'' he said.

9. But the cripple said, ``You \textit{can} climb up.

10. ``I'll show you how,'' he said.

11. As the blind man was climbing, at a certain moment he asked, ``Have I reached
it yet?''

12. But his friend said, ``You're not there yet. Keep on climbing a little.''

13. After he had climbed some more, when he [the blind man] asked again, he [the
cripple] said ``You've gotten there!'' .

14. When he said ``You got there!'' since he [the blind man] really wanted to see,
he opened his eyes wide with all his might, and all of a sudden his eyes could
see again!

15. In just the same [magical] way, when he took the bird's eggs in his hand, these
eggs turned into a snake, so because he was so terrified he dropped it right towards
the place where his friend was sitting, at which point his friend also became terrified,
so he suddenly jumped up and his leg straightened out.

16. So then, we Lahu of this generation also ought to learn from these two people
and take it to heart.\footnote{Lit., ``If we absorb the lesson of these two people it is good.'' \textbf{hên-yù}: `learn from, get something out of'.}

17. Since they heeded the words of the fortune-teller, we've learned that even
the blind man got to see again.

18. We've learned also that the cripple's leg straightened out.

19. Everybody who takes good advice, their descendants and all of their relatives
as well, if they take all this to heart it will be good.

20. Thus our Lahu ancestors told us long ago.

