
{\LARGE{}29 The Potter and the Laundrymen or Greed Leads to Grief}

{\LARGE{}1. Once upon a time there were two people.}

{\LARGE{}2. One of them was an expert laundryman. The other was an expert potter.}

{\LARGE{}3. One day, when they were comparing their wealth, [the potter thought]
he didn't earn as much as the laundryman, money-wise.[2]}

{\LARGE{}4. This potter thought that the laundryman was trying to ruin him.}

{\LARGE{}5. [So] he got an idea about a certain jet-black elephant that the local
prince [3] had.}

{\LARGE{}6. The master potter knew in his heart that if the laundryman were to
wash that elephant [4] and it didn't turn white, he [the laundryman] would lose
his livelihood, so he went [to the prince] and said:}

{\LARGE{}7. ``My lord, O prince, if you would only wash the elephant white, when
people would see it they'd say, ``This is amazing!``}

{\LARGE{}8. So then the Prince said, ``Who is there who can wash this elephant
of mine white?''}

{\LARGE{}9. When he asked this, the potter said, ``A certain expert laundryman
can wash it white.}

{\LARGE{}10. ``In fact he has a lot of money, more than me, more than me the master
potter!}

{\LARGE{}11 ``If you summon that fellow [5] and make him wash it, I believe it
will really turn white,'' he said.}

{\LARGE{}12. When he said this, the lord prince said, ``Mûlò, go and summon here
that master laundryman!''}

{\LARGE{}13. Then when he had been summoned and came, he [the laundryman] said,
``My method of washing is that I must have a great big pot made, a huge basin [100],
to wash in.''}

{\LARGE{}14. When he said this, the prince said to the potter, ``O potter, you
call your assistant(s) and make that pot!''}

{\LARGE{}15. So [they] made the pot and carried it [in], [200] and they took the
elephant's foot and put it in, and it shattered.}

{\LARGE{}16. After [the elephant] broke it once, broke it twice, broke it three
times, the lord prince said, ``Since you first thought you would ruin another person,
now you have been ruined [yourself].}

{\LARGE{}17. ``Do not earn your living as a potter anymore!}

{\LARGE{}18. ``As for the laundryman, earn your living well by washing!'' he said.}

{\LARGE{}19. That's the end.}

{\LARGE{}Notes:}

{\LARGE{}[2] This last word is meant to carry the afterthought-like postposing
of the object }{\LARGE{}\textit{phu}}{\LARGE{} `money, wealth'.}

{\LARGE{}[3] A }{\textit{cāw-fā}}{\LARGE{} ( < Shan caaw phaʔ
[cf. Si. câaw fÁa]) was a Shan ``prince'' or ``sawbwa'', of whom there were many
in the Shan states.}

{\LARGE{}[4] Instead of }{\LARGE{}\textit{h}}{\LARGE{}ɔ}{\LARGE{}\textit{ }}{\LARGE{}`elephant',
the story-teller here says }{\LARGE{}\textit{nô ve}}{\LARGE{} `the one up there',
perhaps out of respect for the }{\LARGE{}\textit{sawbwa}}{\LARGE{} whose possession
it was.  More details on extent of their domains < Wikipedia.}

{\LARGE{}[5] }{\LARGE{}\textit{nô ve phâ }}{\LARGE{}`the fellow up there'.}

{\LARGE{}[100] }{\textit{šÁ-qhâʔ }}{\LARGE{}usually means 'cylindrical
rice-steamer' (see DL:1162), but in this context it seems to mean 'huge basin (large
enough to wash an elephant in)'.}

{\LARGE{}[200] The non-3rd person benefactive particle  }{\textit{lâ}}
{\LARGE{} is used instead of }{\LARGE{}\textit{ }}{\textit{pî}}{\scriptsize{}
}{\LARGE{} , apparently to impart greater vividness.}


