
30 The [Blind] Men and the Elephant

1. Once upon a time there was a great big white elephant.

2. This great elephant was tethered outside [of a house].

3. While it was tethered there, there were twelve [blind] people inside the house.\footnote{The story-teller neglects to mention that all 12 people were blind [substitute mɛ̂ʔ-cú-pā for chɔ], as in all versions of this world-famous fable of Indian origin.}

4. These twelve people, one after the other, went and touched [it].

5. When the first person touched it happened to touch a tusk, he said, ``What we
call an elephant is as sharp as a knife!''

6. He \{went/came\} back in, and the chief clerk\footnote{Where did this scribal element come into the story from?} wrote it down in his book.

7. When they sent the second man [to check out the elephant], since he happened
to touch around its ribs, he said, ``It's like a fenced wall!'' and the scribe
wrote it down.

8. One opinions, two opinions, a third opinion---when they sent out the third person,
he happened to touch a leg, so he said, ``The so-called elephant is like a mortar
because it has a tree-trunk,\footnote{Like a giant pestle?} and a flat foot.''

9. When they sent out the fourth person---like---he said, ``The so-called elephant,
its body is big and round, so we can see it easily.''\footnote{This rather lame comment of person \#4 indicates that the story-teller couldn't come up with another elephantine body part for the moment.}

10. All these different opinions were because each of the twelve men happened to
touch a different place, so twelve different opinions resulted.

