
\textbf{32 How Woman Got her Clitoris} and \textbf{The Lawsuit between the Animals
and the Human Beings}

(Cà-bo)

1. This is a story about how God, once upon a time, after He had created Heaven
and Earth, gave out gifts [to all his creatures].

2. Once upon a time, after God had finished creating Heaven and Earth, \{He/God\}
provided all the things needed by all the people and creatures\footnote{ɔ̀-cə̀ can mean either `thing' or, in this context, `creature'.} \{on earth/in
the world\}, whether it was all the \{human beings/people\} or all the kinds of
animals, God created everything for them.

3. After he finished creating [all this], God spoke thus: ``Come take them, everybody's
things! \footnote{Or ``Come everybody, take your things!''}  Whatever it is [you] need.''

4. When he said this, all kinds of animals came.

5. The creatures who [would] bear horns took away [their] horns, and the creatures
who [would] bear fur took away [their] fur.

6. Having done all this, after each creature \footnote{The story-teller uses g̈â `All for humans', here extended to all God's creatures.} had carried everything off, for
the second time God gave them [things] again.

7. And so, when he said, ``Take it!'' they [all] came.

8. A certain woman came, and this is what she said:

9. ``I've heard that You have made something called `a penis'.''

10. Then, at that point, when [God] said, ``OK, come take a penis that I've made,''
this woman suddenly ran up, and she grabbed a big horse penis that was there.

11. A penis that [God] had made. \footnote{Intending it for a horse.}

12. When [God] said, ``This one is not yours! This is for the horse!'' he picked
it up [hurled/threw it] onto the ground and stamped on it, so that's why [pubic]
hair is all tangled and bushy.

13. That's how the old story goes. \footnote{chɔ-mɔ̂-khɔ̂ `traditional saying/story'}

14. A second thing---``We women are still not quite complete,'' they said.

15. ``Our bodies are still not quite complete.

16. ``Please add something,'' [they] once again went to beg [of God].

17. Well, [the God] cut off the tiny little glans \footnote{nī ó-qō, lit. ``penis-head.''} of a dog there, and threw
it so it stuck onto the woman. \footnote{Not clear whether the whole group of women was so provided. Probably meant to apply to the first woman created.}

18. Because of this, women have \{clitorises/clitorides\}, it is said.

19. An old Lahu story tells it this way.

***

20. Now a second matter, the story about how the animals and the human beings \{went
to court/ had a lawsuit\} with each other.

21. One day it happened this way.

22. The people/humans and the animals took each other to court.

23. These animals went to God and said, ``Oh, those so-called humans of yours,
they're always shooting at us.

24. ``They don't even \{call/ invite/ summon\} us [to make friends].

25. ``They don't treat us [well] in any way at all.

26. ``All they do is shoot at us,'' they said.

27. ``You please teach them properly, those humans.''

28. So God asked the animals, he asked a certain barking-deer, \footnote{\textit{Cervulus muntjac}, known as the best eating in the jungle, now all but wiped out in Thailand.} ``Haven't they
invited you?'' he said.

29. When God asked him this, he said, ``As for inviting, they do invite us.

30. ``[But] when they invite us, they just strike out at us!

31. ``We don't have time to do anything [to escape them].''

