\setcounter{footnote}{0}

1. Well, I'll just tell you a little story now.

2. Once upon a time, a man was going around trading.

3. As he was doing his trading, when he was selling things in a certain village
\footnote{\textbf{šu} \textbf{qhâʔ}, lit. ``others' village''; \textbf{šu} `remote 3p pronoun'.}, he

happened to go into a widow's home.

4. When he happened to enter the widow's house, there was a little girl there.

5. ``Say, daughter, where has your mother gone?'' he asked the girl.

6. ``Oh, my mother has gone to the fields, she's not home,'' she said.

7. The trader said, ``It would have been better if your mother were here.\footnote{The \textbf{contrary-to-fact} sense is conveyed by the combination of the conditional Punf \textbf{qo} and the change-of-state Pv \textbf{ò}.}

8. ``In my hand I have \textit{sparkling red and yellow balls}\footnote{\textbf{à-mī-jɨ̀ʔ} \textbf{nî} \textbf{ší}. A complex \textbf{pun} is involved here: \textbf{à-mī-jɨ̀ʔ} `spark'; \textbf{ni} \textasciitilde{} \textbf{ní} `red'; \textbf{ši} \textasciitilde{} \textbf{ší} `yellow'; \textbf{šī} `round object, ball'; \textbf{nī} `penis'; \textbf{nī-šī} `testicle'; \textbf{nî} \textbf{šī} `two round objects'; \textbf{nī-šī} \textbf{nî} \textbf{šī} `two testicles'.} to launch at
\footnote{\textbf{pə} \textbf{pî} \textbf{ve}: `send at a 3p'. A more vivid verb is used in sentence (14): \textbf{phɛ̂} `release, let fly at; fire off; discharge'.} her!''

9. Having said this, he went off to do his trading, he went on and on, then after
he had sold

all his wares, when he came back, that little girl happened to be in the widow's


house again.

10. When he came back again he asked the girl, ``Daughter, where has your


mother gone?''

11. ``My mother's not home. She went to the fields,'' the girl said.

12. ``Oh no-o-o! If that mother of yours had only been here it would've been mu-u-ch


better! I would have launched sparkling red and yellow balls at her!'' he said,
and went

off again.

13. ``Oh, mother,\footnote{Here the narrator puts the words into the daughter's mouth, quoting what she said to her mother.} a while ago, somebody said that he had `sparkling red and
yellow

balls' to launch at you, if only you had been at home!''

14. After this, when the time came for him to return from his trading, he arrived
[at the

widow's house] once more.

15. Since the widow had happened to return home just as he arrived, there she was


this time.

16. ``You said a while ago, my child told me, `If your mother had been at home,
I'd have

launched \textit{sparkling red and yellow balls} at her!'"

17. ``Yes, if I can't produce those sparkling red and yellow balls, I'll give you
60. rupees,''\footnote{\textbf{hôʔ} \textbf{khá}: Shan `60'. Cf. Si. \textit{hòk }`6', Shan \textit{khan} `group of 10'. With this Shan numeral, Lahu \textbf{phu} `money' is translatable by the Shan currency in use during the British raj in Burma, i.e., \textit{rupees}.}

he said, so she said, ``O.K., do try doing it!''

18. So since she had said this, he got an idea, and he went off looking for fireflies.

19. After he had searched and searched for fireflies, he stuffed his pants full
of them.

20. Then he led her into a dark room, and suddenly he opened up and released them,


all sparkling red and yellow, he let them go at her.

21. But what happened was, this woman had some hot water on, and when she poured
it

down his pants\footnote{Undoubtedly mistaking the flashing fireflies for flames.} she burned the guy's dick.

22. When his dick got burnt, he suddenly ran away, up there in the jungle, and
he set it

down on top of a rock and blew on it.

23. While he was trying to cool it off in a breezy place, the little girl followed
him, and

said, ``Mommy, mommy!''

24. ``Look, look!'' she said. ``When he made the sparkling red and yellow balls
for us just

now, it wasn't sharp enough, so he's sharpening it up there!''

25. ``On a rock!'' she said.

