\setcounter{footnote}{0}

1. Now I'd just like to tell you an old Lahu story. Listen up now.

2. Once upon a time a father-in-law and a son-in-law were farting and playing tag.
\footnote{\textbf{g̈àʔ-yù} \textbf{dàʔ} \textbf{ve}: lit. ``mutually chase to catch''.} I'll tell you about

it now, so listen.

3. The father-in-law said, ``Hey, son-in-law, if you're so smart, try catching
me when I fart.''\footnote{His fart would be the signal to start the game.}

4. ``OK,'' the son-in-law said, ``I'll chase you. I'll wait until you're ready!
\footnote{\textbf{tàʔ-í} \textbf{chɛ̂}: this usually means 'silently; quietly', but in this context it means 'be ready; be settled down'.}''

5. So the father-in-law immediately let a loud fart,\footnote{\textbf{tê} \textbf{qhɛ̂} \textbf{tɛ̀ʔ} \textbf{ve}: here the expression \textbf{qhɛ̂} \textbf{tɛ̀ʔ} \textbf{ve} 'let a fart' is broken up by a numeral-plus-classifier. The noun \textbf{qhɛ̂} takes itself as a classifier for anal events (bowel movements or wind-breaking).} and the son-in-law took
off after him.

6. And after a while the son-in-law caught him.

7. The father-in-law said, ``You're a real son-in-law!\footnote{\textbf{má} \textbf{tɛ̀} \textbf{yò} \textbf{lɔ}: an elliptical way of saying \textbf{má} \textbf{cà} \textbf{jâ} \textbf{tɛ̀} \textbf{yò} \textbf{lɔ}, `Son-in-law is really smart!'} You've chased me and
caught me. You're a

real son-in-law!''

8. That's pretty funny, isn't it! Very amusing, this Lahu anecdote!

9. That's the end.

