\setcounter{footnote}{0}

1. Once upon a time there was a man who had five sons.

2. Every day all they did was quarrel with each other.

3. What one would say the other wouldn't listen to.

4. So having seen this, their father said to them:

5. ``My boys, every day you quarrel with each other.''

6. `` You don't get along with one another.''

7. ``Break off a stick from a tree and bring it to me.''

8. ``Each one of you break one off and bring it to me,'' he said.

9. Then, as their father had told them to, the sons broke them off and carried
them off to where their father was.\footnote{The non-3p benefactive particle \textbf{lâ} is used here, apparently to situate the action from the father's receptive point of view, i.e. 3p > 2p benefaction.}

10. ``Now each of you try to bend and snap the sticks you broke off,'' he said.

11. At this point, when the sticks they had each broken off were bent, they immediately
snapped \textit{crack-crack}.\footnote{\textbf{cɛ̂ʔ-cɛ̂ʔ} (onomatopoeic adverb): sound of dry sticks breaking, mice munching paddy, etc. DL:485.}

12. So then he sent them to break off [sticks] again.

13. Having sent all his sons to go back again to break off [sticks] to carry back,
they set them down before

their father.

14. This time the father tied them nicely together, then he had them bend [the
bundle], one after the other.

15. Even though one [son] after the other bent it, it wouldn't snap.

16. Then the father said, ``My sons, if each of you stays by himself, just the
same as a while ago, with

the breaking of those sticks, you will be destroyed. You will have to fail [in
life],'' he said.

17. ``But just as when you bent them again this time, if you all stick together
and are united in feelings

with each other\footnote{\textbf{ni-ma} \textbf{tê} \textbf{šī} \textbf{tí} \textbf{te} \textbf{ve}: lit. ``making only a single heart''.}, even if others should try to harm you, you will not be destroyed,''
he said.

18. Well, it's just the same with us Christians too.

19. If we put our trust in the Lord Jesus Christ, we will achieve everlasting life.
.\footnote{\textbf{co-tìʔ-ha-tìʔ}: co...ha is a couplet used in elaborate expressions to mean `life force; immortal soul' (see DL:479); tìʔ is a prefixable morpheme (ult. < Skt. \textit{divya}) meaning `eternity; eternal abundance; something supernatural' (see L:608-9).}

20. If we do not place faith in the Lord Jesus Christ, we will fall into temptations
to sin.

21. Therefore I would like to sing a little song about how if we put our trust
in the lord Jesus Christ, the

Precious Son of God,\footnote{\textbf{yâ-phu-tìʔ}: this Christian expression contains the same morpheme tìʔ as in the previous note.} we will achieve everlasting life:

22. [sings] Do welcome gladly God's Precious Son,

And you will win Life Eternal.

