
{\textbf{\#38 The Trickster and the Traders}}

{1. Once upon a time there was a guy called ɛ̀-qhɛ̀ the Trickster, [1]
}

{O Lahu!}

{2. This ɛ̀-qhɛ̀ first fashioned legs and stuck them onto an ashy pumpkin
}

{[2], then he sat down by the side of a road, and stayed sitting there.}

{3. Soon a large group of traders arrived there to do some trading, driving
}

{a hundred mules [loaded with goods].}

{4. So then the Trickster said, ``O traders, let's trade your mules for
}

{mine! [3]}

{5. ``This one of mine brings forth many babies, and it's also very fast!''
}

{he said.}

{6. So then that group of traders thought ``It's probably true,'' and they
}

{made the trade.}

{7. Then, after the Trickster had gotten away with that big herd of
}

{mules, the traders mounted his ashy pumpkin one after the other,     }

{and whipped it with all their might.}

{8. But since they couldn't get it to move, they lifted it up and flung
it into    }

{a mÁ-kwī bush [4], whereupon two rabbits that had been in there    }

{jumped out fast, and when the rabbits jumped out and ran away,    }

{they thought ``It's true! By the Buddha, it's true, it's true!''}

{9. ``Catch `em, catch `em, catch `em, catch `em, go get `em,``
they said.[5]}

{10. So after chasing for three days without catching them, they wept
}

{and went back home.}

{---------------------------------------------------------------------------------------------------------------}

{[1] See DL:129.}

{[2] a-thɛ-phɨ \textasciitilde{} ha-thɛ-phɨ \textasciitilde{} a-thɛ́-phɨ
[}{\textit{Benincasa hispida}}{].}

{[3] I.e., for the ashy pumpkin made to look like a donkey.}

{[4] A kind of bushy gray creeper with sour shoots and edible black fruits
that grow in bunches like grapes.}

{[5] ú-chò: an interjection is used when chasing animals. It was inadvertently
omitted from DL.}


