\setcounter{footnote}{0}

1. Well, then, there was still another Trickster.\footnote{I.e., ``here's another Trickster story.'' The Trickster is here referred to either by his name (\textbf{ɛ̀-qhɛ̀} \textasciitilde{} \textbf{ɛ̀-qhay}) or as ``the orphan'' (sentence \#13, below). The wily orphan is one of Trickster's manifestations. The hapless group of (Chinese) traders is another stock motif of these stories.}

2. He put down a packet of rice on the lower part of a road.

3. On the upper part of the road he put a packet of curry.\footnote{\textbf{ɔ̄-chî} (lit. ``rice-enhancer, rice-elevator''): that which is eaten with rice, conventionally translated `curry'. Cf. Thai \textit{kàpkhâaw}.}

4. When he was ambling along carrying a spear, a band of traders came

along, and [he said to them]:

5. ``Hey you guys, don't do any [other] trading now!\footnote{Instead of the usual negative imperative adverb \textbf{tâ}, Trickster here uses the clause-final verb-particle \textbf{mâ-yɔ}, which conveys a brusque or ironic impression. See DL:974.}

6. ``Just trade me your things!

7. ``Let's trade this spear of mine for your belongings. I'll trade my
spear

with you.

8. ``This spear of mine has magic powers!\footnote{\textbf{câ-tìʔ-dɔ̀-tìʔ}: lit. ``eat-magic-drink-magic'', i.e. `an eternal abundance of food and drink'.}

9. ``If you stick it down into the lower part of the road, you get rice
to eat.

10. ``If you stick it down into the upper part of the road, you get meat
to eat,''

he said.

11. So then those traders said ``Let's check it out!'', and when they did it,


when they stuck it into [the earth on] the lower part of the road, they got


a packet of rice.

12. When they stuck it into the upper part of the road, they got a packet


of chicken.

13. So then they were ecstatic, and they thought, ``It's true!'', and after


the Orphan had obtained their horses and had fled,

when that band of traders got hungry at noontime, they thrust that

spear down into the lower part of the road, but they didn't get anything.

14. When they thrust it into the upper part of the road they got nothing either.

15. So when this happened they wept and went back home.

