
39 The Trickster's Spear and the Merchants

\. Well, then, there was still another Trickster. \footnote{I.e., ``here's another Trickster story.'' The trickster is here referred to}

\. He put down a packet of rice on the lower part of a road.

\. On the upper part of the road he put a packet of curry. \footnote{ɔ̄-chî (lit. ``rice-enhancer, -elevator''): that which is eaten with rice,}

\. When he was ambling along carrying a spear, a band of traders came along, and
[he said to them]:

\. ``Hey you guys, don't do any [other] trading now!\footnote{Instead of the usual negative imperative adverb \textbf{mâ}, Trickster here}

\. Just trade me your things! And I'll give you this spear.

\. I trade this spear of mine for your [stuff].

\. This spear of mine has magic powers!

\. If you stick it down to the lower part of the road, you get rice to eat.

1\. If you stick it down into the upper part of the road, you get meat to eat,''
he said---

1\. So then those traders said \{``Let's check it out!''/``Try doing it!''\}, and
when they did it, when they stuck it into [the earth on] the lower part of the
road, they got a packet of rice.

1\. When they stuck it into the upper part of the road, they got a packet of chicken.

1\. So then they were \{ecstatic/very happy\}, and they thought, ``It's true!''
and after the Orphan had \{obtained/gotten\} their horses [in exchange], when that
band of traders got hungry at noontime, they thrust that spear down into the lower
part of the road, but they didn't get anything.

1\. When they thrust it into the upper part of the road they got nothing.

1\. So when this happened they wept and went back home.

