\setcounter{footnote}{0}

1. Once upon a time the Lahu were fighting a war with the Chinese.\footnote{This story, although recorded in Northern Thailand in 1965, dates from the faraway time when the ancestors of these Thailand Lahu were living in close contact with the Chinese. The Chinese by no means always came out ahead in conflicts with the once fierce Loloish tribes of Yunnan, and it was common for some Lolo groups to keep Chinese war prisoners as slaves. Most of the Lahu people still live in China.}

2. Well, on a certain morning, the morning for going off to war, people were selected\footnote{The Lahu now in Thailand are all quite recent immigrants from Shan State in Burma, where they have been living for centuries as they slowly filtered southward from their ancestral homeland on the Yunnan-Tibet border. It is likely that it was the Shan (a Tai people) who did the conscripting of recruits for military service in skirmishes against the Chinese. There are still many more Lahu living in Shan State than in Thailand.}
and they went off.

3. [But] while others had made rice-packets\footnote{The usual wayfarer's food is a cake of steamed rice wrapped in a banana leaf.} for themselves, this fellow ɛ̀\textit{-qh}ɛ̀\footnote{This is the Trickster's proper name. It is sometimes pronounced more like \textbf{ɛ̀-qhày} or \textbf{ɛ̀-qhàn}.}
had wrapped up a sparse-feathered little chicken\footnote{\textbf{g̈âʔ-qú-ni}: `a kind of tailless chicken without breast feathers', lit. ``naked chicken'' (\textbf{ɔ̀-qú-ni} `something naked').} to take with him, and on the
road, when they ate, when it was time to eat, he unwrapped it and made as if to
eat his meal.

4. As he unwrapped it and the little chick hopped right out, he said, Oh, dear,
I mustn't go [with you]! You all go [on ahead without me]! As for me, since I've
seen this omen,\footnote{According to traditional Lahu belief, anything unusual or scary that happens on a journey is taken as an evil omen, and as a persuasive reason to go back home.} I'm going home. I can't possibly fight a war now. See you guys
later!''\footnote{The expression \textbf{qòʔ} \textbf{e} \textbf{šā} , lit. ``I shall go back'', is a common Lahu formula for leave-takings.}

5. And so the others let him go back.

6. Then, when they had let him go back, he ripped open a bumblebee's nest\footnote{\textbf{pɛ̂-g̈ù-li} `bumblebees'. They live in holes in rotting trees, one insect per hole. The Trickster must have put the bees into some kind of container, though the narrator doesn't bother to say this.},
scooped out some crabs\footnote{From a body of water nearby.}, and tied them so they dangled from a bamboo-sheath
\footnote{\textbf{vâ-lɛ́-qôʔ}: `a crinkly covering adhering to the surface of young bamboo plants.' Trickster selected this material for the noise it would make when scratched.}, and placed this [all] outside the house [of a certain woman].

7. Then when night fell\footnote{\textbf{mû} \textbf{phə̀ʔ} \textbf{ve}: lit., ``the sky is revealed'', is the Lahu idiom for night falling. The idea seems to be that only at night is the sky revealed for what it really is, a black star-studded expanse. \textbf{phə̀ʔ} is the same verb as `unwrap' (sentence 3 above).}, and [the bees went] \textit{bzz-bzz-bzz }and [the
crab] went scratching \textit{teh-teh }against the bamboo-tube―er, bamboo-sheath
\footnote{The narrator momentarily misspoke himself, saying first \textbf{vâ-dɛ́-qō} `tube of bamboo' instead of \textbf{vâ-lɛ́-qôʔ}.}, Trickster said to the woman, ``Aha, the spirits of your husbands\footnote{The Trickster is now talking to all the jittery women of the village, who have evidently all come running out of their houses. He uses the expression \textbf{nɔ̀-hɨ} \textbf{ɔ̀-pā} `your (p.) husbands'. \textbf{ɔ̀-pā}, lit. ``a male'', is a slightly pejorative term for husband, since the ordinary word in Black Lahu is \textbf{ɔ̀-phɔ̂}.} have
come back to you! Don't stay here any more! It's too scary!''

8. So they all went trooping over to the Trickster's house to sleep.

9. Then that Trickster said, ``Oh, come now, don't sleep at my house. Your kids
will shit all over everything.''

10. So then they swore to him, saying, ``Oh, no, we won't let them shit. We guarantee
\footnote{The women use the rather high-flown loanword \textbf{lâʔ-lɔ} `guarantee' (cf. Thai ráprɔɔŋ), which adds to the comic effect.} it!,'' and they were allowed to sleep at Trickster's house.

11. Then, after the bunch of women fell asleep, the Trickster thoroughly mashed
up some shitbeans\footnote{\textbf{nɔ̂ʔ-qhɛ̂}, lit. ``shit-beans'', are fermented soybeans made into paste or cakes and often eaten fried with vegetables and ginger. The name arises from its dark brown color, its consistency, and its pungent odor. This is also called \textbf{nɔ̂ʔ-kɨ̂}, lit. ``rotten beans'', which is also the literal meaning of its Burmese name \textbf{pê-pouʔ} (Written Burmese pâi-pup).}, squeezed [the paste] into balls, and smeared it on the
rear ends of all the children who were there.

12. When dawn broke he said, ``Just look at the way your kids have crapped up my
whole house!\footnote{\textbf{qhɛ̂} \textbf{qò} \textbf{kə}, lit. ``shit into''. The verb \textbf{kə} `insert' is sometimes used as an auxiliary verb to indicate thorough, penetrating action. The phrase `my whole house' is an attempt to render this in English.} What will you give me as compensation?''\footnote{\textbf{šāy} (or \textbf{šɛ̄}) `pay compensation' is another legalistic loanword from Tai (cf. Siamese càaj) `pay'.}, and they said,
``Oh, any compensation at all! Whatever you want to take, take!,'' but when they
offered him money he wouldn't take it.

13. No matter what they offered, he wouldn't take it.

14. Then the Trickster said to the bunch of women, ``Well, go and beat on the Hollow
A-yaw Tree\footnote{\textbf{a-yɔ-qō} `hollow \textbf{a-yɔ} tree'. The \textbf{a-yɔ} (either \textit{Lagerstroemia cylindrica }or \textit{L. macrocarpa}) is one of the four `master trees' (\textbf{šɨ̂ʔ-jɔ̂-mɔ̂}) of Lahu tradition. It is a very tall, straight tree, believed especially apt to be hit by lightning.}, and listen [to what it tells you].''\footnote{The whole complex verbal idea, `beat on and listen to what it says', is conveyed by the concise two-verb concatenation \textbf{dɔ̂ʔ} \textbf{na} `beat' + `listen'. It could equally well be translated `beat on and ask it [a question]', since the verb \textbf{na}, like Japanese \textit{kiku}, can mean either `ask' or `listen to', according to context.}

15. In order to go beat on that Hollow A-yaw Tree-oracle, he made the women take
the long way around.

16. The Trickster himself\footnote{\textbf{ɛ̀-qhɛ̀} \textbf{chi}, lit. ``this Trickster'', here translated `the Trickster himself'. The determiner \textbf{chi} `this' and the demonstrative \textbf{ô-ve} `that' are sometimes interchangeable as anaphoric markers, and are usually both translated simply as `the'. (The narrator tends to alternate them for variety.)} went by the straight road, and got inside of the
Hollow A-yaw, and the women tried beating on the outside, going \textit{thump-thump,
}saying, ``What sort of compensation shall we give to Trickster?''

17. And the Trickster said\footnote{No doubt in a disguised ``oracular'' voice, though the story-teller made no effort to reproduce this in his narration.}, ``Ah, give him compensation, give him compensation!''

18. So they went back home again.

19. The Trickster also ran back after the women as fast as he could, [so he was]
sitting in his house [when they got there], and said, ``Well, what did it say to
you?,'' and they said, ``Well, all it said was, `Give compensation to Trickster!'''

20. ``Well, then,'' he said, ``unless you beat on it once more, you won't know
[what to do],'' so again they [went to] beat on it and ask it [a question]―the
Trickster had quickly gotten inside of it again―and the women in a body beat
on it,\textit{ thump-thump, }and it said, ``Oh, give him cunt, give him cunt!''
\footnote{\textbf{cha} `female pudenda, vagina, cunt'. Lahu has no euphemisms for human sexual organs or excretory functions. \textbf{cha} (or \textbf{cha-pɛ̀ʔ}), like \textbf{nī} or \textbf{nī-qhɛ̀ʔ} `male genitalia, penis, cock', is plain-spoken, neither medical, circumlocutory, nor vulgar.}

21. At this the bunch of women ran back home, and the Trickster also rand back
fast, and sitting in his house working this time with strips of rattan he said,
``Well, what did it say? What did it tell you?'' and they said, ``Well, Trickster―it
just said, `Give him cunt, give him cunt!'''

22. Whereupon Trickster proceeded to fuck\footnote{\textbf{pàʔ} `copulate, have intercourse with, fuck'. See preceding note.} every single woman in the village,
one after the other.\footnote{\textbf{yâ-mî-ma} \textbf{tê} \textbf{qhâʔ-qhâʔ} \textbf{ve}: the meaning `every woman in the village' is conveyed by the reduplication of \textbf{qhâʔ} `village'. The idea `one after the other' is provided by the auxiliary verb \textbf{phôʔ} `perform an action as one of a group; perform several similar actions.' As a main verb \textbf{phôʔ} means `to pile up; make a heap'.}

23. Well, when all their husbands came back [from the war], they said, ``Oh, Trickster
did this and that to us! After you all were gone, Trickster fucked us!,'' so straightaway,
as soon as all the husbands were back, they chased after Trickster trying to kill
him, and he ran away.

24. He ran away and they chased him, but they couldn't catch him.

25. No matter how they chased him they couldn't catch him.

26. Unsuccessful once again, they came back home, while this Trickster, this guy
\footnote{The story-teller uses a succession of two topic noun-phrases in apposition, both to establish sufficient contrast with the ineffectual actions of the husbands, and to give himself a bit of breathing space before launching into the rapid series of clauses which are to follow in the same sentence.}, he scooped out some honey\footnote{Presumably from a nearby tree with a bees' nest. The economy of detail the story-teller permits himself is possible because his whole audience (except for the poor recording linguist) had already heard the story many times before.} and smeared himself with it, and then\footnote{Instead of `and then', the phrase \textbf{qhe} \textbf{te} \textbf{šē} \textbf{lɛ-ɔ̄} could be translated `having done this'. The presence of the verb-particle \textbf{šē} `prerequisite action' makes this expression ``verbier'' than the lexicalized conjunction \textbf{qhe-te-lɛ} which occurs so often in this text.}
he went and lay down inside a storage basket\footnote{\textbf{thɛ-qō} or \textbf{phɛ-qō}: a huge basket, often taller than a man and several feet in diameter, used for storing paddy or other harvested crops.} in a cotton-house\footnote{\textbf{šá-lâ-yɛ̀}: an outbuilding where cotton was stored. The cotton sticking to the honey gave the Trickster a sort of tarred-and-feathered appearance, so that he looked like something ``out of this world''.}, then
when they saw him in the joss-stick temple\footnote{\textbf{ša-tú-kɨ̀-yɛ̀}: lit. ``incense-burn-place-house'', a building where there was an altar for burning incense sticks. Lahu animists burn incense and candles to honor the supreme supernatural \textbf{g̈ɨ̀-ša}, the last syllable of whose name seems to be related to the morpheme \textbf{ša} `joss-stick, incense-stick'. The Pidgin English word \textit{joss}, meaning `idol, cult-image', itself derives ultimately from Portuguese \textit{deos} `god'.} they said, ``Oh, the god has appeared
to us today!''\footnote{\textbf{g̈ɨ̀-ša}, here translated `the god', is the Great Spirit of Lahu animist religion, conceived of as an incorporeal and abstract being far above the nature-spirits, who are rather stupid and often malevolent. \textbf{g̈ɨ̀-ša} has his own priests, who enjoy higher status in the community than the ``spirit-doctors'' who deal with the \textbf{nê} or nature-spirits. It was relatively natural for the missionaries to adapt the notion of \textbf{g̈ɨ̀-ša} to apply to the ``Christian God''.}

27. So they made offerings to him.

28. When they gave him the food he ate it.

29. When he did this\footnote{A real god is supposed to inhale merely the ``breath'' or essence (\textbf{šá}, \textbf{ɔ̀-šá}) of a food-offering, so Trickster is revealed as a non-divine being. This incident is reversed symbolically at the end of the story, where Trickster claims to have made his fortune by ``imbibing the essence'' of the water-- thus indirectly proclaiming himself to be a god.}, they thought ``It's only the Trickster!,'' realizing
it all of a sudden, so they grabbed him and tried to beat him to death, but again
he ran away.

30. Again they chased him, and oh, this time they were chasing and chasing him,
he pulled out his penis and skinned back his foreskin\footnote{\textbf{nī} \textbf{qɛ́} \textbf{ve}: `retract the foreskin'. \textbf{qɛ́} means `to extract through a narrow opening.' It is also used, e.g., for removing mucus from the nose.}, and scratched together
some leaves to cover himself, and made his penis all stiff and straight\footnote{So that only his glans protruded through the covering of leaves. (It is to be noted that English \textit{glans} is itself a metaphorical extension of a plant-name, ``acorn''.)}, and
they said, ``Well, Trickster's trail has petered out now!\footnote{The reader will perhaps forgive this \textbf{pun}. The literal meaning of this sentence is ``There is no long any way to chase after Trickster's trail''.} And here's a red
phalloides mushroom\footnote{\textbf{mù-ní-g̈ɔ} or \textbf{mù-ní-g̈wɛ}: a kind of tasty ground mushroom with a red cap resembling a glans penis. The 2nd syllable \textbf{ní} means `red', but recalls the homophonous (except for tone) \textbf{nī} `penis'. \textbf{mù} means `mushroom' and \textbf{g̈ɔ} means `round'. ``Red phalloides'' is a feeble attempt to render this phytonym in English.} that has sprouted.''

31. After this they went home.

32. After they went home, the Trickster went back again too, and this time when
they were chasing him, this guy, this Trickster, he managed to climb up into a
big clump of trees.\footnote{\textbf{šɨ̂ʔ-cɛ̀} \textbf{ɔ̀-pu-ló} `big clump of trees'. The narrator amends this to \textbf{šɨ̂ʔ-cɛ̀-ló} `a big tree', when this clause is ``chain-repeated'' in the next sentence.}

33. When he had gotten to the top of a big tree, at that point, as they chopped
away at it, trying to get it to fall into the river down below, he said, ``Oh,
in my position\footnote{This renders \textbf{ŋà} \textbf{lɛ̀}, lit. ``as for me'', a strongly topicalized 1st person pronoun.}, if you would only chop so that I fall into the water down
there, I'd be able to swim back out and I wouldn't die.

34. [But] if you chop me\footnote{The verb \textbf{thu} `chop down' is here construed with a personal direct object, although it is much more usual to have `tree' as its object. This is preserved in the translation, and in fact reinforces the symbolic psychosexual meaning of the chopping, i.e. an attempt at castration of the potent by the impotent.} so I land on that rocky cliff up there, I'd surely
be smashed to smithereens!\footnote{\textbf{nɛ̂ʔ-kàʔ} \textbf{qay} \textbf{ve} `go smash; get smashed to smithereens'. \textbf{nɛ̂ʔ-kàʔ} is one of a class or morphemes that forms vivid adverbial expressions with the verb \textbf{qay} `go', much like English `go kaputt', `go blooey', etc.} I would die!''

35. And so they chopped it in such a way that its fall was broken\footnote{\textbf{ŋɛ} \textbf{pî}: \textbf{ŋɛ} `lean on', \textbf{pî} `give; causative auxiliary', i.e. `cause to lean against; cause to be propped up on'.} by the cliff
up there, and he managed to escape into a cave in the rocks.

36. Then, when the Trickster had gone into the cave up there, the others blew smoke
\footnote{The Lahu says `blew fire inside', although \textbf{à-mī} `fire' must be translated as `smoke' for clarity in English. Burning brush at the mouth of a cave is a common method for smoking animals out, or of asphyxiating them so they can be removed after they die.} inside with all their might.

37. But try as they would, they couldn't get it to blow inside.

38. He had taken off his shirt and was fanning it out again―from where he was,
inside there.

39. They poked in sticks―they kept hacking off sticks and poking them inside.

40. So the Trickster took a shit and smeared it on the sticks, and it sta-a-nk
\footnote{The verb \textbf{nù} `stink' is pronounced with exaggerated drawled intonation.} like hell, and they said, ``Aha, the Trickster's shit is coming out of him
now! He's dead!,'' and they set a falling-log trap\footnote{\textbf{va-tɛ̂} `falling-log trap': a delicately contrived trap which the animal springs by stepping on a board which instantaneously activates a series of rods, the last of which releases a heavy log. The husbands, although they believe Trickster to be dead, leave this trap at the mouth of the cave as a precautionary measure. Since the trap was set up outside the cave, \textbf{kə} does not have its literal meaning ``insert'' here, but its more abstract meaning of `in such a way that the verbal action penetrates its goal'.} and went back home.

41. After this, the Trickster took up the jewsharp\footnote{\textbf{á-thâ} `jewsharp': a pair of supple bamboo strips with thin reeds and notes a fourth apart, which the player blows on while flicking the ends up and down. The two notes are alternated an iterated with considerable rhythmic complexity. This instrument is used mostly in courtship. Technically \textbf{á-yɛ̂} refers to a slightly different kind of instrument with its strips tuned an octave apart. \textbf{á-thâ-á-yɛ̂} sometimes means `jewsharps of both kinds', but here it is simply a more sonorous synonym of \textbf{á-thâ}.} he had carried with him,
and blew on it with gusto\footnote{'With gusto' translates the emphatic particle \textbf{qha}, as well as the drawled intonation on the following concatenative particle \textbf{lɛ}.}, and early the next morning two young girls came
to have a look, and when they heard the sound of the Trickster playing the jewsharp,
they said, ``Yoo-hoo\footnote{'Yoo-hoo'' was chosen to render the Lahu vocative particle \textbf{ò} , since its ``Blondie-and-Dagwood'' quality suggests the empty-headedness that the girls display.}, Trickster, please let the two of us come in!''

42. Then he said, ``Well, then, let the one who is truly uglier come in first,''
and when the one who was truly uglier came in first, she was crushed to death.
\footnote{It is hard to understand the girls' docile stupidity here. From our point of view it should have been more effective for Trickster to have said, ``Let the prettier one of you come in first.'' Then presumably they would have vied for the honor of being first. One can only conclude that the girls were so much under Trickster's spell that they obeyed him literally, making an objective assessment of who was uglier.}

43. Thereupon the Trickster managed to escape outside once more.

44. Then, again they chased and chased him, and for once they caught him.

45. When they caught him they tied him to the ass of a water-buffalo, and dragged
them along.\footnote{\textbf{g̈ɔ̀-šɛ}, lit. ``drag-lead''. Here \textbf{g̈ɔ̀} is used in its literal sense of ``drag'', and not merely as an enlivener of the verbal idea. It is immaterial whether we take the unexpressed object to be the buffalo or the Trickster, since they are tied together.}

46. Then they bent his face over downwards and dragged him over a piece of rocky
ground, and he said ``A-lo, a-lo, a-lo!''

47. Then they dragged him over some grassy ground, and he said ``A-lo, a-lo, a-lo!''

48. And when they dragged him [again] over the rocks they said, ``Serves you right,
serves you right, serves you right!''\footnote{The interjection \textbf{qâ-càʔ} is used to gloat at somebody's discomfiture. The story-teller emphasizes Trickster's plight so that the audience can fully enjoy it, but he won't be kept down for long. This episode, among its other functions in the story, serves to put Trickster's ultimate triumph into higher relief.}

49. [But] he didn't die.

50. Again he managed to get out, and once more they chased after him.

51. Once they had seized hold of him, they tied him so he hung underneath a big
bridge over the water, some distance away.

52. After they had left him strung up there, at that point, a large group of traders
\footnote{As we soon find out, these were Chinese traveling traders. This provides an interesting symmetry with the beginning of the story, where the men are going off to fight the Chinese. Evidently the confrontation with the Chinese operated at many levels in ancestral Lahuland.} came along.

53. And they said, ``Hey, Trickster!\footnote{Note that the traders recognize the Trickster without ever having been introduced to him. Trickster is larger than life and known to everybody.} What are you doing there under the bridge?
\footnote{This could equally well be translated `Why are you under the bridge?', since the usual Lahu way of saying `why' (\textbf{à-thòʔ-ma} \textbf{te} \textbf{lɛ}) means literally ``doing what?''},'' and he said to that bunch of Chinese, ``Since my eyes are sore, my eyes
are `imbibing the essence!'''\footnote{\textbf{šá} \textbf{mɨ̀} \textbf{ve}, here translated `imbibing the essence', is an expression of key symbolic importance in the story. It does not occur in ordinary speech, and is clearly meant to be pregnant with meaning. \textbf{šá} means `air, breeze, vapor; breath; vital force, spirit, energy'. The verb \textbf{mɨ̀} means `to pass air noisily through the mouth or nose'', as in \textbf{ə̂ʔ-mɨ̀} \textbf{ve} `hiccup', ə̀ʔ \textbf{mɨ̀} \textbf{ve} `belch', \textbf{qa-mɨ̀} \textbf{ve} `sing', \textbf{tí-šîʔ} \textbf{mɨ̀} \textbf{ve} `whistle', \textbf{há-mɨ̀} \textbf{ve} `yawn', \textbf{há-thî} \textbf{mɨ̀} \textbf{ve} `sneeze'.}

54. ``Well, we would like to imbibe it also, Trickster!'' they said, and he said,
``Let me loose then!''

55. So they let him loose, and he got free.

56. Now that group of traders had been driving their mules, horses, and elephants
around to the villages, having come to trade them.

57. Since they all happened to have sore eyes\footnote{Due to the long dusty roads they had to travel. Trickster had alertly noticed that their eyes were red.}, he tied them up, and as they
were trying to ``imbibe the essence of the water'' under there, under the bridge,
he cut them down so they fell in and died, and Trickster got away with all their
goods, their silver and their gold.

58. After this he went back, and when he had returned to his home again they said,
``Oh, Trickster, how did you get [all this]? How did it happen that you got so
much silver and gold even though we tried to kill you?,'' and he said, ``Well,
I just got it by `imbibing the essence' of the water!\footnote{With this remark the Trickster is indirectly revealing himself to be a god, for only gods can live off the incorporeal essence of things. This is a symbolic inversion of the incident above, where he was found out not to be a god when he ate the substance of the food-offerings in the temple.} Since it was you all
who did this for me.''\footnote{\textbf{nɔ̀-hɨ} \textbf{te} \textbf{á} \textbf{lâ} \textbf{lɛ}: the particle \textbf{lâ} indicates action that benefits a non-3rd person (in this case 2nd person benefitting 1st person). Trickster is pretending to be grateful to the husbands for having strung him up under the bridge.}

59. Then they said, ``Oh, we'd like to get it too!,'' so he led them away, making
each man carry with him to that place a pickling-tub\footnote{\textbf{í-mû-qō}: a large-mouthed vessel made of earth, sand, and cement, used for pickling.} as big as a forehead-basket
\footnote{\textbf{qhá-jū-lu}: a deep loosely woven basket used for transporting water-gourds, carried by a tumpline around the forehead so that the hands are left free.}, and he strung them up there underneath [the bridge], and after the people
had gotten inside [their tubs],he slowly released the ropes and made them fall
into the water.

60. Immediately he made them sink down into the water, and they thrashed around
furiously.

61. Then, since there was one old codger who hadn't gone in, he said to him, ``There,
there, there, look! The others are fighting with each other over the gold!,'' so
he jumped in too, and [Trickster] cut through [the ropes], and after the whole
villageful [of men] had died, he got the women of all the people in the village
for himself.

