
\textbf{\#46 God bless this meeting!}

1 May the blessings of God be upon all of us at our conference!

2 It is a great source of joy for our Christian community to have this privilege.

3 We have much for which to praise God's grace.

4 Our meeting together, while we are making friends with each other, the real purpose
of our meeting is to glorify the kingdom of God.

5 Our convening this conference has made it clear that the work connected with
this great meeting has made a lot of progress.

6 However this work is not yet finished.

7 In the future we must strive to continue doing it.

8 Until what we have hoped for has come to pass, we must strive to continue [this
work].

9 Let all of us who are Christians keep thinking in this way.

10 All of us who are Christians should take good care of our own bodies.

11 How much effort should we exert, how much work ought we to do for God's sake?

12 This work for which we must devote so much time, exert so much of our effort,
pray so earnestly, this is the work of all Christians.

13 So all of us Christians need to take responsibility for this work.

14 If everybody takes responsibility for his own household, if we do this, this
work can be completed speedily.

15 All those people who heed the words of God must strive to do this in the future.

16 This work is not only to be done in our country, but it is the work of Christian
communities everywhere.

17 All of us Christians, not only in our homes, not only in our communities, not
only in our villages, not only in this meeting, need to take responsibility.

18 In the future as we exert even more energy, we have great hopes that the kingdom
of God will come to fruition.

19 May the glory of God's love be upon us all!


