
{\LARGE{}\textbf{\#48 God, Parents, and Village Welfare}}

{\LARGE{}1. Now what do we have to do so that all of our children in the future
can be assured of     }

{\LARGE{}the  stability of our village's site?}

{\LARGE{}2. We have to listen carefully to what the government officials [1] say.}

{\LARGE{}3. More importantly, if you want to achieve a long life, listen carefully
to what your parents     }

{\LARGE{}say. [2]}

{\LARGE{}4. And even above that, if you want your spiritual life to improve, you
must properly follow     }

{\LARGE{}in the footsteps of your God, the Lord Jesus Christ.}

{\LARGE{}5. And also besides that, whether it's your older brother or your older
sister, they are like     }

{\LARGE{}your parents with respect to the material side [3] of your life on earth.}

{\LARGE{}6. One's parents, if we consider the earthly, material point of view,
are like everybody's God.}

{\LARGE{}7. Everything throughout the heavens [4] is truly [the work of] God, the
creator of heaven     }

{\LARGE{}and earth.}

{\LARGE{}8. Therefore all of you boys and girls should take careful heed of the
excellent words of     }

{\LARGE{}Lord Jesus.}

{\LARGE{}9. Finally, may peace be upon all of your households!}

{\LARGE{}---------------------------------------------------------------------------------------------------------------}

{\LARGE{}[1] mì-gɨ̀ jɔ̂-mɔ̂: lit. ``lords of the land''.}

{\LARGE{}[2] See the fifth of the Ten Commandments: ``Honor thy father
and thy mother, that thy days may be long upon the land which the Lord thy God
giveth thee.`` }{\LARGE{}\textit{Exodus}}{\LARGE{} 20:12.}

{\LARGE{}[3] ɔ̀-šɨ̄-ɔ̀-šā-phɔ̂: lit. ``the aspect of one's blood and
flesh''.}

{\LARGE{}[4] mû-nɔ-ma qhôʔ ve qha-pə̀-e: lit. ``everything above
the heavens``.}


