
53 Monkey Hunting:

\. Let me tell you now about how we hill-dwelling Lahu go hunting nowadays.

(Scene 1)

\. A: Hey, buddy!  Since tomorrow is Friday, let's go hunt monkeys and spend the
night [someplace], OK?

\. B: Ah, where shall we go to spend the night?

\. A: We'll just spend the night at Pù-pɔ.

\. B: Are there lots of monkeys [there]?

\. A: They say there are lots of them.

\. B: Is Pù-pɔ in a river valley?

\. A: I hear that it is.

\. B: Let's go, come on!

1\. A: If you're going, I'll go.

(Scene 2)

1\. A: Hey Kwei!  There are monkeys here!

1\. B: (Kwei) Where?

1\. A: There, look up there in front of you!  Come here quick!---Come here quick!
Come and shoot!

1\. B: Where?

1\. A: Look way over there in that branch!  Hurry up and shoot!

1\. B: Look at it quietly!  I'll shoot it down for you.

1\. A: It fell, it fell!

1\. B: I'll cut a tube of bamboo and we'll cook it up.

1\. A: Monkey-meat---shooting monkeys---you go roast\footnote{lÁ: to roast in a bamboo tube ( \texttt{<} Shan \textit{laam})} it in a tube.  Hurry up!
It's getting dark...It's done, come quick, let's eat!

2\. B: Yeah.

2\. A: Kwei, eating monkey meat is really delicious, eh?

2\. B: It's delicious!  The best tasting thing there is!

2\. A: Come on, let's go back.

2\. B: OK, let's go.

2\. A: I'll just sing a little song first.  About eating monkey meat.

2\. Song:

Monkey-meat is great to eat, great to eat!

Delicious to eat, monkey meat!

Monkey meat is scrumptious!

