\setcounter{footnote}{0}


1. Now I shall tell you a story. Listen to it with attention, my Lahu brethren!

2. Once upon a time a tiger and a dhole\footnote{dhole: a wild canine [\textit{Cuon alpinus}] with yellowish fur, a black nose, and a red tail.} were talking together.

3. The tiger said, ``There is no one who is cleverer or stronger than me!''

4. ``Do you really think you're so clever?'' said the dhole. ``You're
not, you know. There is somebody who is even greater than you."

5. ``Who is it?'' said the big tiger. ``Are there still
those things called humans around?"

6. ``There are indeed! There are lots of them. If you would like to test
your cleverness against them, I'll show you where they are."

7. ``Let's go, let's go! Just show me! I'll take a quick bite out of them
and show you how I eat them up,'' he said.

8. One day, when the tiger saw two girls going to the fields, he said to the dhole,
``Are they the ones?'' 9. When he asked this, the dhole, seeing
that they had no gun, said ``No.''

10. Another time, when the tiger spied a man carrying a gun and coming to the fields,
and when he asked the dhole about it, saying, ``Hey, dhole, is this the
one?``, the dhole said ``Yes, yes it is.''

11. When the man noticed that the tiger had seen him and was running at him, he
aimed the double-barrelled gun\footnote{\textbf{nâʔ-tà-cɛ}: \textbf{nâʔ-tà} `length of a rifle', (\textbf{ɔ̀-)cɛ} `pair'.} he was carrying and shot a blast of buckshot
from one barrel.

12. When [the tiger] came at him again, he shot six bullets\footnote{The round-object classifier \textbf{šī} is here used instead of the full noun \textbf{nâʔ-šī} `bullet'.} at him from the
other barrel.

13. The tiger ran away, and the dhole said to him, ``Wasn't I telling you
the truth?"

14. ``Yes! I can't stand up to him. He shot at me--bang, bang!--and I got
bitten and tingly all over, and I just couldn't stand it!'' he told him.

15. ``Just so,'' said the dhole.

16. This story, if each and every one of us thinks it over to himself, has relevance
for us all.

17. We may think, ``How full of cleverness I am!'', but it is not
so.

18. There is always someone greater than me!

19. There are probably some among us Lahu who are still that way\footnote{I.e., taking a conceited attitude.} up to the
present day, aren't there?

20. If there are, O my brethren, make haste to repent, please.

21. Thus am I warning you. Mark my words well. This story is over.

