
60 The Six Stupid Men and the Gophers

1. Once upon a time six stupid men went to dig up gophers .\footnote{fâʔ phî: `gopher; bamboo rat' [\textit{Rhizomys sumatrensis}]}

2. Having dug for gophers, they caught seven of them.

3. So after they caught the seven animals, they [proceeded to] divide them up [among
themselves] .\footnote{`Among themselves' translates dàʔ `reciprocal action'.}

4. As they were dividing them up, one person sat amongst them and said, ``Here,
one for you; here, one for you; here, one for that guy; here, one for that other
guy''---dividing them up like this, the six people got six of the animals, but
one was left over.

5. So then, when they did it again like this, dividing them up again, still just
one animal was left over.

6. When one of the guys \footnote{tê   g̈â   phâ is a slangy way of saying chɔ tê g̈â  one  person (clf)  fellow, guy} took the one that was left over, he had two of them.

7. So they divided it up again.

8. But even after they divided it again, [the last] guy got the two animals that
were left.

9. So \{all the livelong day/ the whole blessed day\} \footnote{tê-ni-tâ-vâ: tê ni `all day' + tâ-vâ `id.' < Tai (Si. thÁŋ wan).  The two equivalent expressions are juxtaposed for emphasis.} they kept on dividing
up the gophers this way, until they were all quarreling with each other.

10. Since they were acting this way, they took a very long time doing it, when
all of a sudden a hunter appeared.

11. When this hunter appeared, he said, ``Hey stupids, what are you all doing?''
When he asked them this, they said, ``We're dividing up these gophers, but we can't
agree about it, so we're quarreling \{this way/ like this\}.''

12. So the hunter said, ``Bring them to me, all the gophers!'' and when he said
this, they brought them \{immediately\} all to one place.

13. The hunter divided them up for them, saying, ``Here, one for you; one for that
guy; one for that other guy,'' and after he had divided up six of the animals among
the six people, the hunter took the remaining one for himself.

14. So then those six fools said, ``Now this is they way that a person who knows
how to divide does it!'' and they were very happy.

15. Ha, ha, ha!

