\setcounter{footnote}{0}

1. Once upon a time there were twelve hunters.

2. Among these twelve hunters there was an orphan, bereft of his mother, of both
parents\footnote{\textbf{yâ-cò-ši} `motherless child', \textbf{fɨ-kôʔ-ni} `orphan bereft of both parents'.}, who worked as their servant.

3. Well, then, [one day] they arrived at a place in the jungle, a hunting ground
in the deep forest, and when they got to their camp\footnote{\textbf{chɛ kɨ̀} lit. ``staying place''}, they found some tiger
footprints, so they couldn't spend the night there.

4. Well, those hunters were all scared stiff, so they made a huge basket.\footnote{\textbf{qhá-jū-lu} \textasciitilde{} \textbf{qhá-pū-lu} `large loosely-woven carrying basket with a headstrap'. See Plate \#11 in DL.}

5. They went inside this basket, and made places for themselves to sleep.

6. When it got dark, they went and hoisted up the basket to a high place\footnote{Probably tied to a high tree-branch by its carrying strap.}, and
climbed up into it to sleep.

7. At this point, the orphan said, ``Master hunters, please lend me a knife\footnote{In order to cut bamboo to make a basket for himself, not to use as a weapon against tigers!}!'',
he said.\footnote{It is common in Lahu narratives to both introduce and follow a quotation by ``he said'', but of course one of these is best left untranslated.}

8. ``I want to be safe just like you guys are safe.''

9. ``I don't want to die either.''

10. ``''Since there are so many places where tigers live around here, since
there are lots of tigers, I want to sleep up high just like you guys!'' he said.

11. But those master hunters didn't lend him a knife.

12. So the orphan picked up some scraps of bamboo that had been whittled off and
discarded\footnote{When the master hunters were making their basket.}, and went inside a great big clump of bamboo.\footnote{\textbf{vâ-pu}: a mass of bamboo with their roots intertwined.}

13. When he went in there and looked around, he saw that it was just big enough
for a person to sleep in.

14. So he picked up the rope he had made out of the bamboo scraps, and having looked
around inside, he went in there to sleep for the night.

15. Then, when it got dark, all those hunters went into the basket to sleep.

16. Well, when the tigers came, they jumped up, bit into them and ate them up,
all of them, the ten of them, they bit into them and ate them all up.

17. At this point, after they had all been bitten and eaten up, those tigers caught
sight of the orphan who had slept in the bamboo clump, so they went around it and
thrust their tails into it, went around and stuck them in.

18. Then, when the orphan [saw] the tigers' tails had been thrust in and stuck
in, he took his rope and tied up their tails, tethering them to the bamboo clump.

19. That's just what he did, to those ten tigers.\footnote{Here we learn for the first time that there were ten of them.}

20. Then, when dawn broke, he came outside [from the bamboo clump], and took all
the knives of the dead hunters, and hacked away at all the tigers.

21. So that whole bunch of tigers all died.

22. That's the end.

---------------------------------------------------------------------------------------------------------------

