
66. Candid Conversation: Working up to being recorded

Cà-lɔ̂: 1. Hurry up and do it!

Headman's wife [HW]: 2. I'll bring my dog and have him \{do it/ record\} for you!

Somebody: 3. Bring your dog and come put him down here [1] for us!

HW: 4. The dogs are right outside, a big pack of them.

***

Headman [H]: 5a. Come on, come on you guys have nothing to be bashful about!

5b. They [2] say they want to get stuff

5c. When it's time for him to go back to America---our Lahu way of life.

5d. Anything and everything, [3] \{this and that/ all kinds of things\}.

5e. Like, we could \{pretend to be/ ``do''\} bachelors [4]: ``I wonder if we should
go down there to pitch a little woo'' [5]---``If we go down there there's not going
to be \{anyplace/ any chance\} to court,'' one guy can say, and the other one can
say, ``Over there you've got to wait! You'd have to stay down there a long time.''

Cà-lɔ̂: 6. Tell a little bit about the Kachin [6] looking for a wife!  Why's
he looking for one, I wonder.

<Woman laughs>

H: 7a. [addressing someone presumably living in the Jingpho's house while he is
on his quest] I hear the Jingpho is coming back from looking for a wife.

7b. You stay at home and pound lots of rice and store it, and when he gets back
you can eat it up---say something like that.

Cà-lɔ̂: 8. Will he get one, or won't he---think about that, and talk nicely
about it!

H: 9a. Hey, hurry up and do it, hurry up, you guys! About everything---

9b. Oh, oh, I'm going to roll me one of those great big Klèt-Thawng [7] cigars
like you've seen the idiots smoke---

9c. In the old days in Shan State we used to smoke great big cigars that would
fill your mouth right up!

Cà-lɔ̂ [henceforth CL]: 10. Were they what was called \textbf{šê-pyîn-leiʔ}
