\setcounter{footnote}{0}

<told by Headman Cà-bí>

1. H: We-e-ll, brethren of mine, I'll just tell you a story about how a rabbit
and a tiger matched wits\footnote{\textbf{cû-yī} \textbf{te} \textbf{dàʔ} \textbf{ve}: lit. ``squabble over intelligence''.} with each other! Do you want to hear it? Who wants to hear it?

2. Several voices: I want to hear! I want to hear it!

3. They say that long, long ago rabbits were very intelligent.

4. And in the old days tigers didn't bite people and eat them.

5. They didn't even eat raw meat.

6. Rabbits lived among humans, and tigers lived among humans.

7. One day a tiger and a rabbit made friends with each other, and went along together
towards a place where humans were living.

8. When they got there, Rabbit\footnote{Rabbit evidently reached the human habitation before Tiger did. From now on we usually omit the article and capitalize the animals' names. They are usually referred to with the determiner \textbf{chi} `this': \textbf{patây} \textbf{chi} `this/the rabbit' \textbf{lâ} \textbf{chi} `this/the tiger'.} said to a child, ``Oh, look there! That tiger
says that when your father and mother come back from finding you food, he'll also be waiting to eat with you all.''

9. When he said this, he was deceiving Tiger, that rabbit was.\footnote{I.e., he was blowing Tiger's cover. The humans probably wouldn't have been willing to share their food with him, if they had been forewarned.}

10. Rabbit was very smart, so he ran the hell away.\footnote{\textbf{gèʔ-e} (< \textbf{gɛ̀} `be fast') is an intensifier added to certain verbs (especially verbs of motion). See DL:408.}

11. As for Tiger, he frightened the children in that [human] house, the little
kids in that house, and they were crying!

12. When the father and mother came home, [the father] took a knife and chased
after Tiger, slashing at him and striking at him.

13. Tiger for his part followed the path over there that Rabbit had taken a while
before, saying [to himself], ``Rabbit betrayed me, so I'll just bite and eat him\footnote{\textbf{chèʔ} \textbf{câ} \textbf{ve}: the usual Lahu expression for being devoured by an animal.},'' and he went chasing after Rabbit.

14. Rabbit went on and on, and in a certain place on the path he
sat down near a great big pile of buffalo-shit and kept watch over it.

15. Tiger said, ``Rabbit, the one who betrayed me just now must have been you.
Today I'm going to eat you up.''

16. R: ``No, no! It sure wasn't me'' he [Rabbit] said.

17. T: ``What are you doing there?'' he [Tiger] said.

18. R: ``Long, long ago, in the age of my ancestors, in the time of my
great forefathers, they gave me this stool\footnote{There is a fortuitous polysemy here between two senses of English \textit{stool: }1. backless chair 2. excrement.} and made me sit near it and watch
over it,'' he said.

19. T: ``Oh, I beg you to let me sit on it for a while!'' said Tiger. (If you say
it in Shan, he said ``\textbf{yɔ̂n naŋ ít nɨŋ}.'')\footnote{Lit. ``beg to sit one time.'' This story is evidently of Shan origin, since this Shan phrase stuck in the narrator's mind. Several other Shan words appear later in the story.} That means ``Let me
sit for a while.''

20. R: ``Oh, that's impossible. In the time of my ancient forefathers, they said
that if I let anybody else sit on it, it would collapse!''

21. ``However, I can't not let you sit on it,'' said Rabbit to Tiger.

22. When Tiger started to sit down, Rabbit jumped up, turned around, and ran the
hell out of there.

23. Tiger sat down heavily,\footnote{\textbf{g̈ɔ̀} \textbf{mɨ}: the pre-head versatile verb \textbf{g̈ɔ̀} (lit. ``drag'') gives a vigorous meaning to the head verb.} and the buffalo-shit stuck to his arse in a big
yellow gloppy mess.

24. When the buffalo-shit stuck onto him, Tiger said, ``That Rabbit deceives people!''
and he suddenly went chasing after him again.

25. When he chased him down again, he said ``The one who lied to me must have been
you.''

26. Rabbit was sitting up on a crooked branch of a tree, near a great big beehive,
with his ears drooping, saying to Tiger that he was watching over a huge gong.\footnote{I.e., Rabbit pretended that the bees' nest was a gong.}

27. Then when [Tiger] said, ``Rabbit, you're tricking me,'' [Rabbit] said ``No-o-o!
Ages ago in my forefathers' time they entrusted me with this great duty [to watch
over the gong].''

28. ``Oh, I beg you to let me try striking it once, please let me strike it just
once!\footnote{Tiger says it first in Shan (\textbf{yɔ̂n mâun}), then in Lahu (\textbf{lɔ̀} \textbf{dɔ̂ʔ})}'' Tiger said.

29. ``Impossible,'' said Rabbit. ``Impossible.''

30. ``Oh, please, I beg you to let me strike it!'' said Tiger.

31. ``Well, if you do strike it, I'll just have to jump up and turn around,'' said
Rabbit, and he jumped up and ran away.

32. Tiger then smashed his fist into the beehive over there.

33. The bees went inside [his body] and bit him, and his nose got all swollen,
his mouth got tiny,\footnote{Because of the stings around his mouth, his oral orifice was almost completely blocked.} his mouth swelled up, his head and his face got huge.

34. So then Tiger said, ``That Rabbit lied to me\footnote{Lit. ``lies to people,'' but the non-3\textsuperscript{rd} person benefactive Pv \textbf{lâ} shows that Tiger means that he is the object of deception.}!'' and went chasing after
him again. He was quite expert at following trails, that Tiger.

35. He was always chasing to attack.

36. Some distance away, by the side of the road in an abandoned field, [Rabbit]
had coiled a green pit viper\footnote{\textbf{vɨ̀-nɔ} \textit{Trimeresurus popeorum}. Only slightly venomous, but with a very painful bite.} around an old dried out tree-stump, a black log, and put it to sleep warming itself in the sun.

37. ``Rabbit, you deceived me,'' he [Tiger] said. ``I'm going to bite into you
and eat you up, this very day.''

38. ``Oh, no!'' he [Rabbit] said. ``Long, long ago, in the time of my forefathers,
in the time of my grandpa and grandma, they put me in charge of guarding this curved tobacco-pipe.''

39. So Tiger said he would like to try smoking it just once. (He was using
the Shan language, [as was normal] long, long ago.) ``I'd like to try smoking just one puff from that big pipe.''

40. ``Impossible, pal of mine!\footnote{He uses a Lahuized Shan loanword \textbf{lùʔ} for `smoke'.} I have been entrusted with it since the time
of my ancient forefathers,'' he said.

41. Well, he finally ended up having to let him have a smoke.\footnote{Rabbit uses the friendly Shan-derived word \textbf{šaháy} `buddy; pal', ult. < Pali \textit{sahāya }`friend, ally'. See DL: 1158.}

42. So tiger eagerly went for that green head of the pit viper, that coiled-up
head, thinking that even one puff would be delicious, and his great big mouth swelled up.

43. So then―and this was not all\footnote{\textbf{tɔ}: causative of \textbf{dɔ̀} `drink; smoke (tobacco)', like Jse. \textit{tabako wo nomaseru }< \textit{nomu} `drink; smoke'.}―Rabbit ran the hell away again and [Tiger]
raced after him, and over there in a place in the mountains, a human being who was selecting some thatch-grass to pick up and carry home caught sight
of the rabbit.\footnote{\textbf{mâ} \textbf{qɔ̀ʔ} \textbf{hêʔ} \textbf{šē}: lit. ``still not the case this time.'' See DL: 972}

44. ``Rabbit, where are you going?'' he said.

45. ``Oh, there's a tiger that's chasing me to eat. The tiger is running after
me and beating me.\footnote{Note the sequence of four verbs in fortuitous concatenation, (\textbf{šɔ̄} \textbf{chi} \textbf{pû} \textbf{qòʔ}), expressing a series of consecutive actions.} I'm nearly dead. Where could I go to escape [him]?''

46. ``Come inside my load of thatch-grass!'' said the human to Rabbit.

47. Whoosh!\footnote{\textbf{g̈àʔ} \textbf{dɔ̂ʔ} \textbf{lâ} \textbf{ve}: ``pursuing me with blows.''} He went into the load of thatch.

48. Well, having followed him again, Tiger got there, all smeared with buffalo-shit,
with swollen eyes where the bees had stung him, with his mouth swelled up where
the viper bit him\footnote{\textbf{vɨ̀-nɔ} \textbf{chèʔ} \textbf{tā} \textbf{ve} \textbf{mɔ̀ʔ-qɔ} bá \textbf{ɛ̀} \textbf{qay} \textbf{lɛ}: Clause added by JAM.}, and said, ``Haven't you seen a rabbit around here?''

49. ``No, I haven't,'' he [the human] said.

50. ``Wow, you're such a tiny little person, but your load is so big! Let me carry
it a little,'' said Tiger.

51. The human said, ``That's impossible.''

52. So he begged him, ``Pu-leeze\footnote{Tiger uses the Shan word \textbf{yɔ̂(n)} `beg' instead of normal Lahu \textbf{lɔ̀}.} just let me carry it a little!

53. Your load is too big for you!''

54. So then [the human] said to Tiger, ``Let's make an end to your damn\footnote{This expletive translates the verb-particle \textbf{qhɛ} `iterative excess'. At this point the human gives Tiger the load to carry.} begging
right now!''

55. From inside that thatch something was going \textbf{khɛ̂ʔ-khɛ̂ʔ}!\footnote{Onomatopoetic adverb: ``with a scratching sound, like a match on sandpaper.'' DL: 379.}

56. ``Hey, I wonder what's inside your thatch? Something is going\textbf{khɛ̂ʔ-khɛ̂ʔ}!''
he [Tiger] said.

57. ``Oh, it's just the thatch rattling around,'' he said.

58. It was Rabbit inside there that was setting the thatch on fire.

59. Tiger was staggering under the weight.

60. It was going \textbf{tɛ̂ʔ-tɛ̂ʔ.}

61. A flint had been struck there.

62. ``Oh, I wonder what in the world is in your thatch! It's going \textbf{tɛ̂ʔ-tɛ̂ʔ}
now.''

63. ``The thatch is just rattling around,'' the human said.

64. All of a sudden the thatch burned brightly and the flames leapt high.

65. [Tiger] didn't have the sense\footnote{For the ramifications of the polysemous morpheme \textbf{khɔ̂}, see DL 382} to throw it away, you see, so he ran away
all on fire, not looking for human beings anymore, and fled into the

most overgrown parts\footnote{\textbf{qhɔ̀} \textbf{ná} \textbf{chò} \textbf{phɔ} \textbf{e}: Correlative conjunction \textbf{qhɔ̀...chò }``where V1, there V2''. DL 307-8.} of some old abandoned fields.

66. Well, there were large herds of horses and cattle grazing there.

67. ``Mr. Ox, Mr. Horse,\footnote{Tiger adds the male-name prefix \textbf{cà}- to the animals (\textbf{cà-nû}, \textbf{cà-mû}), which creates a humorous humanizing effect, since most traditional Lahu men's names were also derived from the day in the 12-day calendrical cycle of animals when the boy was born. See DL 448-451.} where could I run away to today?''

68. ``Run into the most overgrown part of the abandoned field over there,'' they
said.

69. When he tried to flee into the deepest part of the abandoned field he was burning
so badly that his flesh became covered with big stripes,\footnote{As tigers are to the present day!} that

mighty tiger.\footnote{\textbf{lâ-qhâ-pɨ}: a more dignified name for the tiger than simply \textbf{lâ}.}

70. He couldn't flee to a safe place.

71. So then, after running and running he caught sight of a water-buffalo bathing
in a buffalo-wallow.

72. ``Oh, Mr. Buffalo, today I'm burning almost to death. Where can I run to get
away?''

73. ``Come on in here,'' he [Mr. Buffalo] said.

74. When he went into the pool the fire died, the fire just died.

75. The charred parts where the fire burned him, some of those burnt spots were
now black and others were blue, his eyes were puffed out, his mouth had become huge, he had sat on a pile of buffalo-shit, and those black-and-blue
stripes―he was a terrible mess!

76. That hive of bees that he hit mistaking it\footnote{Lit. ``naming it a gong.'' See DL: 1012.} for a gong bit into him so his
eyes swelled up.

77. And the pit-viper making his mouth all bent out of shape was because he [Rabbit]
had told him [Tiger] that it [Mr. Viper] was a pipe, a crooked tobacco-pipe, so he said he would just take a puff, and [the viper] bit into him
and his mouth got all swollen.

78. After that he got burned until he had stripes of bare\footnote{\textbf{ní} \textbf{ɛ̀} (AE ext): `something red; something bare, naked'. DL 765-6.} flesh, all red and
raw\footnote{\textbf{qɛ̀ʔ-qɛ̀ʔ-lè-lè}: DL 249.}, and ran to the cattle, ran to the horses, but wasn't saved.

79. They just showed him the way to an abandoned field.

80. When he reached a pool where a water-buffalo was he went into it, and was saved.

81. So to this day, unless a water-buffalo should do something against God's will,
unless he is guilty of sins and crimes, a tiger will not bite him.

82. But a tiger has no respect for horses or cattle.

83. To this very day tigers eat them.

84. And that's the story.

