
{\textbf{78 Government Offers Us New Lands}}

{\textbf{\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}
HEAD}}{1 H: Well, a message from the government has come to us,}{\textsuperscript{1
to all of us Lahu. We're going to discuss it. Let everybody come here for a while,
please.}}

{\textsuperscript{The matter which we're all going to be discussing this
evening is this. There's talk that they're planning to give us all--all of us [Lahu]
who are living in Thailand--lands, fields, places where we can cultivate the soil,
beginning right now and forever after.2 But how is this going to come about?}}

{\textsuperscript{2 Cà-bo: I'm tired of hearing this all the time. It's
always the same old story. }}

{\textsuperscript{3 H: Well that's where you're wrong! I tell you they're
saying that they're going to give us land that we can live off forever in the future,
for our own adult generation as well as our children's generation! They say they'll
give each person about twenty rai.4 What shall we do [about it]? Shall we go take
a look at the places? }}

{\textsuperscript{4 Cà-bo: Where the hell are these places anyway?}}

{\textsuperscript{5 H: Well, }}

{\textsuperscript{5-A T: It's not just in one place.}}

{\textsuperscript{5-B H: I've heard that it's someplace over at Pha-pu-jawn,5
near where the Hmong are. Go over there and have a look, they said to me!}}

{\textsuperscript{6 Cà-bo: Didn't somebody say that they've parceled that
out to the Hmong already?}}

{\textsuperscript{7 H: The Hmong won't get all of it. They say they'll give
us Lahu our share, too! }}

{\textsuperscript{8 Cà-bo: With the business of giving some to the Hmong
and some to the Lahu there's only going to be quarreling back and forth.}}

{\textsuperscript{9 H: Oh, they're going to separate off each person's land
for him, they say. They'll measure it off for us. So who all is going to go, to
have a look at it?}}

{\textsuperscript{10 Cà-bo: Well, I for one am not going. From as far back
as I can remember until today I have never lived off a wet-rice field.6 I've grown
up working in dry mountain-fields, and that's all. There's malaria [down there
in the plains]! I'm afraid of it.7}}

{\textsuperscript{11 H: Well, since they're giving out the land, let's go
take a look at it. That's the least we can do.8 Even if we don't cultivate it right
away, we still have no way of knowing whether we might not gradually have to use
it in the future. They won't be making this offer forever.9 If there comes a time
when they're not giving it out anymore, even if we want it we won't get it. We
really ought to go out and have a look at it, brothers!}}

{\textsuperscript{12 Cà-bo: You lead them there then. Take Ja-keh and his
group and Ja-pho and his group.}}

{\textsuperscript{13 H: Oh, they'll go, they'll go. One or two of them will
go. You should go early tomorrow morning, after you have something to eat.10 The
official11 said that he'd go with us too tomorrow morning, after we eat. He's going
to wait for us. So after we eat--all of you who want to go, everybody who's going,
have something to eat and go up12 and fetch the leader, then go off together and
have a look. Personally I don't think I'll go along yet. Because I don't have time
to go. Right now I'm just swamped with work,13 you see.}}

{\textsuperscript{14 Thû-yì: For crying out loud, you're the headman.
You have to go! Otherwise how can you look after the villagers' interests?}}

{\textsuperscript{15 H: Well, I'm not one who's getting any [of that land]
yet. It's not there that I'm getting any [land], so I won't go yet. I'm not going
there to look. But all of you who do want to get some, you go and look it over!
}}

{\textsuperscript{16 Cà-bo: Well, when the headman informs the villagers
about it this way but doesn't go himself, I daresay nobody's going to pay any attention
to him, in the future. }}

{\textsuperscript{17 H: For my part, I don't want to get14 any [land] over
there because it's too far away. I'm going to try to get some that's near here.
But all of you who want to get some, go. If you don't want any ---}}

{\textsuperscript{18 Cà-bo: What you're saying is really very selfish.
If it's a place nearby you'd like to get it for yourself. But other people you'd
like to send off to places far away. If you act like that how can this village
prosper?15}}

{\textsuperscript{19 H: No, whether it's far or near, if it suits you they'll
give it to you--no matter where it is and no matter how much you want. I'm not
saying I'll take it all for myself! Whether it's good land or bad land they'll
share it out equally, the same to everybody. That's how it's done, when the authorities
give things out. }}

{\textsuperscript{20 Thû-yì: Well, in that case you certainly ought to
go look at it--tomorrow, after you have something to eat.}}

{\textsuperscript{21 H: But I don't want any of it ever at Pha-pu-jaw--really
I don't!}}

{\textsuperscript{22 Cà-bo: Just look at him, just look at him! With his
own mouth he says he doesn't want any.}}

{\textsuperscript{23 H: Well then, where would }}{\textsuperscript{\textit{you}}}{\textsuperscript{
like to get land, eh? Someplace nice and close, right on the }}{\textsuperscript{\textit{nikhom}}}{\textsuperscript{'s
land, not more than a furlong away from Huey Tat, I suppose, right? Well, I [also]
would like to get [some land] right around here. If it's too far away I just can't
afford to go and work there. There's nobody [to help me].19 I don't have any children,
you know.}}

{\textsuperscript{24 Thû-yì: You won't find any place to make a paddy-field
around these parts! Except for Metalai;20 that's the only place.}}

{\textsuperscript{25 H: No, if you go over around Mì-g̈â-tù, on the
way to the Cattle and Elephant Market,21 there's a great big plain! If we can only
get water it'll be a perfect place. Go look around there. }}

{\textsuperscript{26 Cà-bo: Cultivating paddy fields in a desert where
there's no water and earning a living! Maybe your father could, I suppose!22}}

{\textsuperscript{27 H: Come now! Up above the valley there's a stream called
Stony Brook23 -- now if you could just go and dig a channel [from it to the fields]
there'd be all [the water] that you need for cultivation! I guess you haven't been
over there to take a proper look yet. }}

{\textsuperscript{28 Cà-bo: Without spending about ten thousand baht24
you wouldn't be able to dig the channel. }}

{\textsuperscript{29 H: No, as long as we've found25 a place that we want
to have, whatever has to be spent or whatever has to be done the government will
help us. It's a trifling matter for them.26}}

{\textsuperscript{30 Cà-bo: Yes, a long time ago, the government was already
telling us \texttt{"}We'll help you, we'll help you!'', but up to today I don't
even see a mouthful of tea [that we've gotten out of them].}}

{\textsuperscript{31 H: The government says that they can do whatever they
set their minds to do. }}

{\textsuperscript{32 Cà-bo: Two or three pigs, fifty or sixty big coffee-plants--how
can anybody earn a living from that sort of thing!27}}

{\textsuperscript{33 H: Anyhow, whether we do it or not is up to us ---
since they're offering it. We are backwoods people.28 We're a tiny29 race of people.
We ought to listen carefully to what the government says. Whatever they tell us
we should do. As we go forward in the future, if each of us doesn't have his own
piece of land here in Thailand, our children and our posterity will suffer for
it.30 Those of us who are grown-ups really ought to ponder this very carefully
now. In the past too, in this Lahu nation of ours, for generation after generation
people have been thoughtless,31 so that today whatever we do we are beset with
troubles and difficulties.32 We have no education, none of the various skills that
other people have. It is because we don't listen to the advice of others that we
have come to this pass. So at this point we should all think things over very carefully,
my brothers!Liši, where would you like to get land, if they offer you some?}}

{\textsuperscript{33a Liši: Wherever they give it to me, I will accept
it.}}

{\textsuperscript{33b H: Well, that's it, I feel the same as you do. [But]
there isn't anybody who wants to go to Pha-pu-jaw that we were talking about just
now! And you're telling me, \texttt{"}Go! Since you're the headman!\texttt{"} So
what shall I do? }}

{\textsuperscript{=======[Tape VII, Side 2]}}

{\textsuperscript{1 H: Well, a message from the government has come to us[1],
to all of us Lahu. We're going to discuss it. Let everybody come here for a while,
please.... }}

{\textsuperscript{The matter which we're all going to be discussing this
evening is this. \{They're saying that / There's talk that\} they're planning to
give us all--all of us [Lahu] who are living in Thailand--lands, fields, places
where we can cultivate this soil, beginning right now and forever after.[2] But
how is this going to come about--}}

{\textsuperscript{2 Cà-bo: I'm tired of hearing this all the time. It's
always the same old story. }}

{\textsuperscript{3 H: Well that's where you're wrong![3] I tell you they're
saying that they're going to give us land that we can live off forever \{after
/ in the future\}, for our own adult generation as well as our children's generation!
They say they'll give each person about twenty ray.[4] What shall we do [about
it]? Shall we go take a look at the places? }}

{\textsuperscript{4 Cà-bo: Where the hell are these places anyway?}}

{\textsuperscript{5 H: Well, I've heard that it's [someplace] over at Pha-pu-jaw[5],
near where the Meo are. Go over there and have a look, they said to me!}}

{\textsuperscript{6 Cà-bo: Didn't somebody say that they've parceled out
that [land] to the Meo already?}}

{\textsuperscript{7 H: The Meo won't get all of it. [The government] they
say they'll give us Lahu our share, too! }}

{\textsuperscript{8 Cà-bo: With the business of giving some to the Meo
and some to the Lahu there's only going to be quarreling back and forth.}}

{\textsuperscript{9 H: Oh, they're going to mark off [the boundaries of]
each person's land for him, they say. They'll measure it off for us. So--who all
is going to go? To have a look at it--}}

{\textsuperscript{10 Cà-bo: Well, I for one am not going. From [as] far
back [as I can remember] until today I have never lived off a wet-rice field.[6]
I've grown up working in dry mountain-fields, and that's all. There's malaria [down
there in the plains]! I'm afraid of it.[7]}}

{\textsuperscript{11 H: Well, since they're giving out the land, let's go
take a look at it. That's the least we can do[8]. Even if we don't cultivate it
right away, we still have no way of knowing whether we might not gradually come
to have to use it in the future. They won't be making this offer forever.[9] If
there comes a time when they're not giving it out [anymore], even if we want it
we won't get it. We really ought to go out and have a look at it, brothers!}}

{\textsuperscript{12 Cà-bo: You lead [them there] then. [Take] Ja-keh and
his group and Ja-pho and his group.}}

{\textsuperscript{13 H: Oh they'll go, they'll go. One or two of them will
go. You should go early \{tomorrow / in the\} morning, after you have something
to eat.[10] The official[11] said that he'd go with us too tomorrow morning, after
we eat. He's going to wait for us. So after we eat--all of you who want to go,
everybody who's going, have something to eat and go up[12] and fetch the leader,
then go off together and have a look. Personally I don't think I'll go along yet.
Because I don't have time to go. Right now I'm just swamped with work[13], you
see--}}

{\textsuperscript{14 Thû-yì: For crying out loud, you're the headman.
You have to go! Otherwise how can you \{take care of / look after\} the villagers['s
interests]?}}

{\textsuperscript{15 H: Well, I'm not one who's getting any [of that land]
yet. It's not there that I'm getting any [land], so I wont' go yet. I'm not going
there to look. But all of you who do want to get some, you go and look it over!
}}

{\textsuperscript{16 Cà-bo: Well, when the headman informs the villagers
about it this way but doesn't go himself, I daresay nobody's going to pay any attention
to him--in the future. }}

{\textsuperscript{17 H: For my part, \{I don't want[14] / I'm not getting\}
any [land] over there because it's too far away. I'm going to try to get some that's
near here. But all of you who want to get some, go. If you don't want any--}}

{\textsuperscript{18 Cà-bo: What you're saying is really very selfish.
If it's a place nearby you'd like to get it for yourself. But other people you'd
like to send off to places far away. If you act like that how can this village
prosper[15]?}}

{\textsuperscript{19 H: No, whether it's far or near, if it suits you they'll
give it to you--no matter where it is and no matter how much you want.[16] I'm
not saying I'll take it all for myself! Whether it's good land or bad land they'll
share it out equally, the same to everybody. That's how it's done, when the authorities
give things out. }}

{\textsuperscript{20 Thû-yì: Well, in that case you certainly ought to
go look at it--tomorrow, after you have something to eat.}}

{\textsuperscript{21 H: But I don't want any of it ever at Pha-pu-jaw--really
I don't!}}

{\textsuperscript{22 Cà-bo: Just look at him, just look at him! With his
own mouth he says he doesn't' want any....[17]}}

{\textsuperscript{23 H: Well then, where would }}{\textsuperscript{\textit{you}}}{\textsuperscript{
like to get land, eh? Someplace nice and close[18], right on the }}{\textsuperscript{\textit{nikhom}}}{\textsuperscript{'s[19]
land, not more than a furlong away from Huey Tad, I suppose, right? Well, I [also]
would like to get [some land] right around here. If it's too far away I just can't
afford to go and work there. There's nobody [to help me].[19] I don't have any
children, you know.}}

{\textsuperscript{24 Thû-yì: You won't find any place to make a paddy-field
around these parts! Except for Meh-tha-lai;[20] that's the only place.}}

{\textsuperscript{25 H: No, over around Mi-gha-tu, on the way to the Cattle
and Elephant Market[21], there's a great big plain! If we can only get water it'll
be a perfect place. Go look around there. }}

{\textsuperscript{26 Cà-bo: How can you \{irrigate a field / make a wet-rice
field\} and \{live off it / make a living\} at }}{\textsuperscript{\textbf{Mu-ko-mi-cho}}}{\textsuperscript{[21]
without any water? Maybe your father could, I suppose![22]}}

{\textsuperscript{27 H: Come now! Up above the valley there's a stream called
Stony Brook[23]--now if you could just go and dig a channel [from it to the fields]
there'd be all [the water] that you need for cultivation! I guess you haven't been
over there to take a proper look yet. }}

{\textsuperscript{28 Cà-bo: Without spending about then thousand baht[24]
you wouldn't be able to dig the channel. }}

{\textsuperscript{29 H: No, as long as we've found[25] a place that we want
to have, whatever has to be spent or whatever has to be done the government will
help us. It's a trifling matter for them.[26]}}

{\textsuperscript{30 Cà-bo: Yes, a long time ago, the government was also
telling us \texttt{"}We'll help you, we'll help you!\texttt{"}--but up to today
I don't even see a mouthful of tea [that we've gotten out of them]. }}

{\textsuperscript{31 H: The government says that they can do whatever they
set their minds to do. }}

{\textsuperscript{32 Cà-bo: Two or three pigs, fifty or sixty big coffee-plants--how
can anybody earn a living from that sort of thing![27]}}

{\textsuperscript{33 H: Anyhow, \{whether we do it or not / if we don't
do it\} is up to us--\{but / since\} they're offering it. We are backwoods people.[28]
We're a tiny[29] race of people. We ought to listen carefully to what the government
says. Whatever they tell us we should do. As we go forward in the future, if each
of us doesn't have his own piece of land here in Thailand, our children and our
posterity will suffer for it.[30] Those of us who are \{responsible adults / the
elders\} really ought to ponder this very carefully now. In the past too, in this
Lahu nation of ours, for generation after generation people have been thoughtless,[31]
so that today whatever we do we are beset with troubles and difficulties.[32] We
have no education, none of the various skills that other people have. It is because
we don't listen to the advice of others that we have come to this pass. So at this
point we should all think things over very carefully. Please, brothers!--Liši,
where would you like to get land, when they offer you some?}}

{\textsuperscript{33a Liši: Wherever they give it to me, I will accept
it.[33]}}

{\textsuperscript{33b H: Well, that's it, I feel the same as you do. [But]
there isn't anybody who wants to go to Pha-pu-jaw that we were talking about \{before
/ just now\}! And you're telling me, \texttt{"}Go! You're the headman!\texttt{"}
So what shall I do? }}

{\textsuperscript{\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}
516355fe221b8becf888516aedf3d8b9e942cb6e34 Cà-bo: Since you're the headman, you
really must go.}}

{\textsuperscript{35 H: I don't want to get any [land there], I tell you!
Even if I go there I'm not taking any.}}

{\textsuperscript{36 Cà-bo: But you must go and look for [land] on the
villagers' behalf even though you don't want any for yourself. }}

{\textsuperscript{37 H: If the villagers want it, they'll get it. }}

{\textsuperscript{38 Cà-bo: You've got to go look after things yourself.
If you don't go and watch how they're measuring it off, how will the villagers
know whether it's being done fairly or not? }}

{\textsuperscript{39 H: Well, once they're all through measuring it properly,
I do plan to go, you know.}}

{\textsuperscript{\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}
HEAD40 Cà-bo: When it comes to doing a job, this kind of piecemeal, half-hearted34
way of doing it is really annoying! Year after year [this trait of yours] hasn't
changed,35 right up to today. It's time for you to realize this. }}

{\textsuperscript{41 H: Even if you say you want [a certain piece of land]
it doesn't mean that you have to cultivate it right away. If there isn't anybody
to work it at the moment,36 you can just keep it in reserve37 for the future! We'd
better take advantage of their offer while it still holds.38 So all of you who
have the time ought to go tomorrow! Otherwise you might say someday,}}

{\textsuperscript{``You didn't tell me about this.'' ``Even though the government
said they'd give us land, you didn't tell me.'' ``If you had only told me when
they said they'd give it to us!'' ``If we had only known about it, we would have
gone!'', you might say, so now I'm telling you about it while there's still time,
[I'm telling] you all.39}}

{\textsuperscript{42 Cà-bo: Yes, well, we've known about this for a long
time, this whole business. However, I don't see any advantage or benefit40 for
us if we do work [these new fields]. I can only see difficulties. Rather than listening
to that kind of advice, it wouldn't hurt41 if we paid attention to what the Boss42
says. }}

{\textsuperscript{43 H: In the future, when Thailand, this Thailand of ours
gets to be packed chock-full of people, whoever doesn't have his own land is going
to be in trouble, I bet. Now, while they're telling us this in good time, we all
ought to think it over carefully. They're only going to be offering it to us for
a while. If we don't watch out \texttt{"}By the time our thoughts catch up, the
prey can't be caught!\texttt{"}43}}

{\textsuperscript{44 Cà-bo: Well, as far as I'm concerned, I'm still not
going.44 You all can discuss it and go if you want to. }}

{\textsuperscript{45 H: Well, then, this is where the matter stands. If
there's anybody who wants to go, let him go tomorrow. That's all there is to it.
As for me, I won't be going yet. I don't have the time yet. Because I'm very busy
working, seeing that I don't have anybody [to help me]. Among all of you who want
to go, [why don't] some of you get together and go off hunting or something46 from
the cattle-tending huts to Porcupine47 Mountain, and cook yourselves a good roast
and continue on from there!48}}

{\textsuperscript{46 Thû-yì: Well, but did they say they'd dig out the
paddy-fields for us? What did they say they'd do? }}

{\textsuperscript{47 H: No, they're not going to dig out the field for us.
We've got to dig ours out all by ourselves. We'll have a go at it, then if it turns
out that we can't, in case we're not able to do it,49 we'll just tell the government
officials. The government will take care of us and help us, they said. Also, in
the matter of bringing water [to the fields], even if it's a long way, if the regime--er,
if the government50 helps us, we'll make it all right. Little people that we51
all are, we've been thinking \texttt{"}This is impossible and that is impossible\texttt{"}--that's
how we think. But, look, if we don't have fields, land to till, whatever we might
try to do in the future, it will really be impossible [for us to get anywhere].
Everybody had better think this over mighty fast.... }}

{\textsuperscript{Elephant-and-Cattle-Market, now, [that's a place] I'd
like to get, too.52 They say the tree-stumps [there] are big, but we [could] chop
them out, [to make our] fields. In a year or two we could have them burned out
and be earning a living from our paddy-fields. Oh, I'm sure we could!}}

{\textsuperscript{=======40 Cà-bo: When it comes to doing a job, this kind
of piecemeal, half-hearted[34] way of doing it is really annoying! Year after year
[this trait of yours] hasn't changed,[35] right up to today. It's time for you
to realize this. }}

{\textsuperscript{41 H: Even i you say you want [a certain piece of land]
it doesn't mean that you have to \{work / live off\} it right away. If there isn't
anybody to cultivate it at the moment[36], you can just keep it in reserve[37]
for the future! We'd better take advantage of their offer while it still holds.[38]
So all of you who have the time ought to go tomorrow! Otherwise you might say someday
that I didn't inform you about this: that even though they said they'd give you
land I didn't tell you. I had only told you when they said they'd give it to you!
\texttt{"}If we had only \{known / found out\} about it, we would have gone!\texttt{"}
you might say. So now I'm telling you about it while there's still time, [I'm telling]
you all.[39]}}

{\textsuperscript{42 Cà-bo: Yes, well, we've known about this for a long
time, this whole business. However, I don't see any advantage or benefit[40] for
us if we do work [these new fields]. I can only see difficulties. Rather than listening
to that kind of advice, it wouldn't hurt[41] if we paid attention to what the Boss[42]
says. }}

{\textsuperscript{43 H: In the future, when Thailand, this Thailand of ours
gets to be packed chock-full of people, whoever doesn't have his own land is going
to be in trouble, I bet. Now, while they're telling us this in good time, we all
ought to think it over carefully. They're only going to be offering it to us for
a while. If we don't watch out \texttt{"}By the time our thoughts catch up, the
prey can't be caught!\texttt{"}[43]}}

{\textsuperscript{44 Cà-bo: Well, as far as I'm concerned, I'm still not
going.[44] You all can discuss it and go if you want to. }}

{\textsuperscript{45 H: Well, then, this is where the matter stands. If
there's anybody who wants to go, let him go tomorrow. That's all there is to it.
As for me, I won't be going yet. I don't have the time yet. Because I'm very busy
working, seeing that I don't have anybody [to help me]. Among all of you who want
to go, [why don't] some of you get together[45] and go off hunting our furry friends[46]
from Cattle Market there over to Porcupine[47] Mountain, and make yourselves a
good roast and continue on from there![48]}}

{\textsuperscript{46 Thû-yì: Well, but did they say they'd dig out the
paddy-fields for us? What did they say they'd do? }}

{\textsuperscript{47 H: No, they're not going to dig out the field for us.
We've got to dig ours out all by ourselves. We'll have a go at a it, then if it
turns out that we can't, in case we're not able to do it,[49] we'll just \{tell
/ say no to\} the government, the government will take care of us and help us,
they said. Also, in the matter of bringing water [to the fields], even if it's
a long way, if the regime--er, if the government[50] helps us, we'll make it all
right. Little people that we[51] all are, we've been thinking \texttt{"}This is
impossible and that is impossible\texttt{"}--that's how we think. But, look, if
we don't have fields, land to till, whatever we might try to do in the future,
it will really be impossible [for us to get anywhere]. Everybody had better think
this over mighty fast.... }}

{\textsuperscript{Elephant-and-Cattle-Market, now, [that's a place] I'd
like to get, too.[52] They say the tree-stumps [there] are big, but we [could]
chop them out, [to make our] fields. In a year or two we could have them burned
out and be earning a living from our paddy-fields. Oh, I'm sure we could!}}

{\textsuperscript{\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}
516355fe221b8becf888516aedf3d8b9e942cb6e So, as far as I'm concerned, I want the
big stretch of flat land above Cattle Market--the whole plain, above everybody
else's [fields].}}

{\textsuperscript{48 Cà-bo: You're so greedy! That's just it. That's why
nobody wants to go [with you], why nobody wants to follow your lead! }}

{\textsuperscript{49 H: Ha, ha! That's nonsense, that is!}}

{\textsuperscript{50      : He's right, he's right!}}

{\textsuperscript{51 H: Okay, okay, I'll take [land] at the tail-end of
everybody instead,[53] if that's the way you feel. Way below everybody else, down
at the \{rocky ground / Rock-}}{\textsuperscript{\textbf{thresh\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}
HEAD}}}{\textsuperscript{\}54--}}

\begin{center}
{\textsuperscript{=======\}[54]--}}

{\textsuperscript{\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}
516355fe221b8becf888516aedf3d8b9e942cb6e[Everybody talks at once]}}
\end{center}

\leftskip=0pt
{\textsuperscript{52 Thû-yì: Hold on! The land down there is mine!}}

{\textsuperscript{53 H: The plain ..... [The rest is drowned out]}}

{\textsuperscript{54 Cà-bo: Just look at him, just look at him, now-- [rest
is unintelligible]}}

{\textsuperscript{55 Pà-ɛ́: It's mine, it's mine! The land down there
is mine! }}

{\textsuperscript{56 H: Look, when the authorities give it to us, we'll
carefully measure it off, the good places as well as the bad, and divide it up.
Everybody will get an equal share.[55]}}

{\textsuperscript{57 Thû-yì: Me, if it's bad land I'm not taking it!}}

{\textsuperscript{58 Pà-ɛ́: I won't take it either--I won't!}}

{\textsuperscript{59 H: [That's so unreasonable][56] what you've been saying.
When I said I'd take the upper part, you said you wouldn't give it to me. But according
to what our forefathers said long ago, it's the lower part which is to be taken.
Because when you get land below other people's, you can[57] gradually \{move /
space your way\}[58] a little bit upwards one year, and a little bit further upwards
the second year, until someday the one who got the lower part will get all the
upper fields as well! There's an old saying to that effect. It's a good one. And
[yet] I say I want the upper part, and you won't give it to me!}}

{\textsuperscript{60 Cà-bo: Just look at him, just look at him! Now he's
even teaching you the way to rob other people of their possessions!}}

{\textsuperscript{61 H: I'm not robbing anybody, I'm not robbing anybody!
I [told] you I'd take the upper part even if there wasn't enough water. [But when
I said] I'd take the upper part, what you said was \texttt{"}If it's a good place
you take it for yourself!\texttt{"} So now I'm saying I'll take [land] down there,
way under everybody else in the rocky plains down below! }}

{\textsuperscript{62 Cà-bo: Well, then, what about what you were saying
there about digging your way upwards little by little into other peoples' [land]?}}

{\textsuperscript{63 H: [Laughs] Oh, that! You know there's an old saying
like that, don't you? \texttt{"}When you stake[59] out a paddy-field, stake it
out low! If you get a high field, you may get water, but cold water makes for bad
paddy.[60] If you get land below the others, one day you will get to be greater
than they\texttt{"}--isn't that what they say? As our ancestors have said, \texttt{"}The
first year while you hoe your weeds, undermine[61] the paddy-field boundaries[62]
a little and push up a ways. The next year dig under again and advance upwards
another length. [That is how you become] great.\texttt{"} So I'm just trying to
get other people's [property] for myself! }}

{\textsuperscript{64 Cà-bo: And here's where your greed comes in!}}

{\textsuperscript{65 H: It's not greed at all! You guys don't want [that
land]. It should be given to me!}}

{\textsuperscript{66 Cà-bo: Then let each one of us go down and [work]
only in his own place![63]}}

{\textsuperscript{67 H: Well--}}

{\textsuperscript{68 Cà-bo: This business of knocking over and breaking
up other people's boundaries, you would always have to make new ones and there'd
be no end to it.}}

{\textsuperscript{69 H: [I'm not saying] \texttt{"}knock them over and break
them up\texttt{"}--when the water weeds are thick around there, if you just dig
at them a teeny-weeny eensy-weensy bit, the earth comes tumbling down [into your
field]--if it's soaked with water. }}

{\textsuperscript{70 Cà-bo: Terrific, I'll go scrape yours away too, then!}}

{\textsuperscript{71 H: [Laughing] I'm only kidding![64] You're talking
nonsense![65]}}

{\textsuperscript{72 Pà-ɛ́: Don't cultivate the upper part. Don't cultivate
the [whole] middle part. Dividing up the middle part, each person will get half.
I'll tear them [the boundary markers] down.}}

{\textsuperscript{73 H: Will you do it honestly?[66]--Well, you know what
people say, if they're narrow ones--if all the stripes [of land] are narrow, they'll
say, \texttt{"}It's too long! I don't want it!\texttt{"}[67] So let everybody measure
off his own section squarely. }}

{\textsuperscript{74 Thû-yì: Well, if you keep doing it that way...}}

{\textsuperscript{75 Cà-bo: Don't give any to people who don't want to
accept it.}}

{\textsuperscript{76 Thû-yì: It doesn't look nice, it doesn't look nice.}}

{\textsuperscript{77 H: When I'm making the boundaries[68] between the sections,
I'll certainly try to do it in such a way that each and every person gets enough.}}

{\textsuperscript{78 Thû-yì: Well--}}

{\textsuperscript{79 Pà-ɛ́: [Interrupts] I'll bet they'll tear them down.}}

{\textsuperscript{80 H: Oh, you people are talking at cross-purposes here[69]!
Just talking at cross-purposes. When the government is good enough[70] to give
us something, let everybody at least[71] go and have a proper[72] look at it! No
matter what land falls to any individuals' lot.[73]}}

{\textsuperscript{81 Thû-yì: It depends entirely on the government.}}

{\textsuperscript{\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}\texttt{<}
HEAD82 H: Look, when the government gives us something there's nothing for us to
go complain about, is there? Shaw-hpa has suggested that we draw lots! You'd write
down74 the paddy-field [locations] and put them in [the pool], then you'd draw
the lots, and whichever piece75 you happen to get, it would be that land which
you'd--that area there that you would get. }}

{\textsuperscript{83 Cà-bo: \{It'd be impossible to do it that way / That
would never work\}. When you draw lots and you lose76 it's not a happy business.
}}

{\textsuperscript{=======82 H: Look, when the government gives us something
there's nothing for us to go complain about, is there? Shaw-hpa has suggested that
we draw lots! You'd write down[74] the paddy-field [locations] and put them in
[the pool], then you'd draw the lots, and whichever piece[75] you happen to get,
it would be that land which you'd--that area there that you would get. }}

{\textsuperscript{83 Cà-bo: \{It'd be impossible to do it that way / That
would never work\}. When you draw lots and you lose[76] it's not a happy business.
}}

{\textsuperscript{\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}\texttt{>}
516355fe221b8becf888516aedf3d8b9e942cb6e84 H: It's because nobody agrees on how
to divide it up! So what can we do about it? }}

{\textsuperscript{85 Cà-bo: What'll happen is, when the lots are drawn
a loser[77] will--}}

{\textsuperscript{86 H: The hell with losers! they just won't get it (=what
they want), that's all. }}

{\textsuperscript{87 Thû-yì: Well, if we draw lots and I don't get a good
place, I simply won't take it! Even if it falls to my lot. }}

{\textsuperscript{88 Pà-ɛ́: I'd take it, I'd take it! }}

{\textsuperscript{89 H: Well, if that's the way it is, there's nothing to
be done with you people. }}

{\textsuperscript{... [two inaudible speeches follow]}}

{\textsuperscript{Well, if you won't even do what I advise you to, let somebody
else tell you what to do, let somebody else tell you. We're all just quarreling
with each other and not getting any place. }}

{\textsuperscript{90 Cà-bo: Let me decide, let me decide! I'll solve them
for you, all these problems of yours. If there is anybody now who doesn't want
to get any [land] as the headman said to, let him just give his share to the Lišu,[78]
all of it. }}

{\textsuperscript{91 H: Which place do you mean? }}

{\textsuperscript{92 Cà-bo: Even the [lands] at Pha-pu-ju.[79]}}

{\textsuperscript{93 H: The \{land / ones\} at Pha-pu-ju, now--}}

{\textsuperscript{94 Cà-bo: And let [the lands] at Cattle and Elephant
Market be given away, too! You people are so stubborn[80] right now! }}

{\textsuperscript{95 Lišu: I for one will take them!}}

{\textsuperscript{96 H: So how are you guys going to go and earn a living
in the future?}}

{\textsuperscript{97 Lišu: No matter where we work for own living[81] we'll
just be scratching out a bare existence anyway.}}

{\textsuperscript{98        : Well, if we haven't got places to live off,
someday we'll be in trouble.}}

{\textsuperscript{99 H: So for me, even if I don't get any [new land], I
won't worry about it. If I do get some--}}

{\textsuperscript{100 : That's why you're not going [out to look at the
new lands], eh?}}

{\textsuperscript{101 H: If I don't get any I won't be angry. If I do get
some, I won't be happy.}}

{\textsuperscript{102 Cà-bo: You're just saying that with your mouth![82]
You do that with everything.}}

{\textsuperscript{103 H: See here, now, I've grown old working oldfields[83]
these many years[84]! \{None of you people earn your living [that way] / Among
you there're nobody who earns his living [that way]\} Me, each year I work somebody
else's oldfield, a different one each time. I go roaming around and earn my living.[85]}}

{\textsuperscript{104 Lišu: Okay, keep scrabbling for a living then! All
by yourself.}}

{\textsuperscript{105 H: Scrabble I will. I'm not going.}}

{\textsuperscript{106 Lišu: Stay sitting all forlorn[86] and keep scratching
for a living!}}

{\textsuperscript{[A few inaudible words follow]}}

{\textsuperscript{107 H: When people give you fellows advice, you don't
pay any attention.[87] Even when those important people advise you and tell you
what to do, you still don't want to listen.}}

{\textsuperscript{108 Lišu: The one who doesn't want to listen is you!}}

{\textsuperscript{109 Cà-bo: You just sit under an eggplant-berry bush[88]
and scratch for a living!}}

{\textsuperscript{110 H: There's a saying about the Lahu people[89], that
\texttt{"}when one of us is working, the other turns away. He is obstructed. When
one of us is moving, the other \{reins him in / pulls him back\}.\texttt{"} We
can't go on like this. We Lahu, no matter what we do--}}

{\textsuperscript{111    : \texttt{"}While one man is climbing a tree, the
other is pulling at his heels\texttt{"}--that saying is to the point here.}}

{\textsuperscript{112 H: Because we don't want to obey the laws and customs
of our superiors or even listen to what they're saying--if there's anything unpleasant
to hear.[90]}}

{\textsuperscript{113 Cà-bo: It's all because you didn't listen from the
very beginning when people were telling you \texttt{"}Don't do this, don't do that,
it's impossible!\texttt{"} So today, you are, it's nothing but troubles, whatever
we do. No matter who teaches you, you won't listen!}}

{\textsuperscript{114 H: Well, unless we let those government officials
handle it for us I don't see what we can do. }}

{\textsuperscript{115    : Then let's leave it to the government. You people
\{stop quarreling about it / don't quarrel about it anymore\}!}}

