
\textbf{\#84 A. The Buffalo-Bugger and the Rabbit Hunters}

[Tape VIII, Side 2]

1 Once upon a time there was \{this / an\} old fellow. 2 He would go off to his
wet-rice field to plow, spending the night in the paddy-field, [but] his wife would't
go [along].[1] 3 One day a son of his, who had gone [home] to fetch some polished
rice [for his father's meal], came back [to the field], and as he reached the edge
of the mountain-field--I mean, the edge of the paddy-field,[2] he looked at his
father--and there was his old man, screwing a female water-buffalo! 4 So the kid
ran back home right away, and told his mother. 5 Then when the mother learned about
this, she packed up some more rice to carry out and went off to her husband,[3]
and there they were sitting and eating [together] \{outdoors / in the fields).

6 Now that day some rabbit-hunters came to the village! 7 Well, as they came chasing
the rabbit, making a great racket, the old fellow asks his wife \{\texttt{"}What's
going on there?\texttt{"} / \texttt{"}What are they doing?\texttt{"}\} 8 So the
wife says, \texttt{"}they've come to beat that buffalo-bugger to death.\texttt{"}
9 So all of a sudden, without even finishing his rice, he runs off fleeing for
his life. 10 Well, as he was running away, the rabbit happened to flee after him
in the same direction, and when they [=the hunters] saw this they \{shouted / said\},
\texttt{"}Come on, let's chase him hard!\texttt{"}--and he ran away with all his
might.

11 Finally when he had run [all he could] and he was \{at bay / at the end of his
rope\}, he said, \texttt{"}But the gray buffalo wasn't yours! And the black one
wasn't yours either!\texttt{"}[4]

