
12 H: Hm, this old fellow your [just] told about--did he live a long time ago?\footnote{The Headman is just trying to get the conversation started again.}

13 Khù-mɔ̂: Yes.

14 Thû-yì: Well, Khumaw, where would you like to go anyway?

15 K-m: I've come to visit.

16 T-y: Aren't you planning to go visit over \textit{there} now?\footnote{I.e., 'wouldn't you rather go find some girls?'}

17 K-m: No, I came to visit here.

18 T-y: Pa-eh, what about you--

19 H: He's probably going to go around visiting on a day like today, Sunday. There's
no sense in just hanging around\footnote{gɨ̂ chɛ̂ "be doing for pleasure, be fooling around" vs. ca gɨ̂ tô "go around for pleasure, go on a round of visits."}--you shouldn't\footnote{Clause + (m)â hêʔ o =  "one shouldn't + clause." Lit, "it is not (a case of) clause." See also below, speeches 46, 76, etc.} just lie around \{blinking
your eyes sleepily / lackadaisically\}! Get moving!

20 K-m: We'll go count the girls over there.

21 H: So you're going courting, huh?

22 K-m: Don't' you guys want to go?

23 T-y: Where to?

24 K-m: Over there--you know.

25 T-y: Over there, at Meh-Jweh--\footnote{The rest of the sentence is indecipherable, as is most of the Headman's following speech.}

26 H: Up there .... don't' you know the way [to get there]?

27a  : See there! She's wearing black!\footnote{This sentence was said at the same time as Thû-yì's speech 27b.}

27b T-y: ... I'm going to go.

28 Pà-ɛ́: We have three of four people [who want to go]!

29 H: Get together and go, get together and go, you young fellows. When I was your
age\footnote{Lit, "when I was like you."} I used to go, all right!

30 Kachin: You should please cut me in on the deal!\footnote{The Kachin is an older man, something of a clown. He was still single, since Lahu girls hesitate to marry members of different tribes. Lit: "Please share together with me."}

\begin{center}
[Laughter. Everybody talks at once.]
\end{center}

\leftskip=0pt
31 H: How many are there, did you say? How many girls?\footnote{Less colloquially: yâmîhÁ qhà-nî g̈â cɔ̀ le qôʔ ve.}

32 Pà-ɛ́: There are four or five of them.

33 H: Four or five?

34 Pà-ɛ́: Yes.

35 T-y: How many good-looking ones are there?

36 Pà-ɛ́: They're all good-looking.

37 T-y: There's one who isn't so bad as the others anyway!\footnote{khɛ̂ ve < Shan khɛn 'superior': "to be the best of a poor lot, to be comparatively good."}

38 Pà-ɛ́: There is one--she's a little better than the others.

39 T-y: That's the one! The one for me!

40       : No, no, no! ... She's mine!

41       : Like hell she's yours\footnote{the khɔ̂ + X: "what nonsense to say X!" thôo ꞊ pòthôo (Interj.), khɔ̂ ꞊ 'words.'}!

42 Pà-ɛ́: She's mine. I'll give you guys the inferior\footnote{ɔ̀-hɔ́ "that which is below, something inferior."} ones.

43 T-y: Well, in that case I'm not going--I won't go.

44 H: I'll divide them up for you, I'll divide them up for you. As for Numero Uno,\footnote{The headman uses Thai numbers for the rankings, for comic effect. This is rendered by the Spanish translation.}
she'll be for this gentleman. Numero Dos\footnote{The headman uses Thai numbers for the rankings, for comic effect. This is rendered by the Spanish translation.} is for this little mother\footnote{The word for 'mother' seems here to be used ribaldry for 'guy' (CHECK)--a striking cross-cultural coincidence.}, and
this guy gets Numero Tres.\footnote{The headman uses Thai numbers for the rankings, for comic effect. This is rendered by the Spanish translation.}

45 T-y: Hey, that's no good, Numero Uno is mine!

46 H: Oh, you can't act that way!

47 Càbo: Good grief, you have no sense of shame\footnote{Lit., "you don't even know the way of shame."}!

48             Numero Uno is mine!

49 T-y: Well, don't fight about it like that!

50 Pà-ɛ́: The one who is Numero Uno is mine, I tell you! She's not yours.

51 T-y: You guys are friends, and there you are fighting!

52 Pà-ɛ́: We're laying conflicting claims to a girl, see?\footnote{vɛ dàʔ ve: 'to vie; dispute a claim with one another.'}

53 Càbo: He says they're fighting over a girl.

54 Kachin: Well, if you people don't want her, why don't you parcel her out to
me!

55 T-y: She's tiny, why don't you tell him! That "Numero Uno"
of his.

56 Pà-ɛ́: Yes, Numero Uno certainly\textit{ is} mine.

57 T-y: Let's take her--if only we succeed in counting her.\footnote{mâʔ g̈a ve 'to have one's courting crowned with success.'}

58 H: Well, so this Numero Uno is a tiny one, is she!

59 Pà-ɛ́: You guys are impossible, just impossible! We can't go on arguing this
way.\footnote{"As for speaking words, doing this way is not mutually possible anymore.""}

60 Càbo: Are you going to have a fight?

61 Pà-ɛ́: We'll have a fight.

62 Càbo: Who was the first one to see that girl?

63 Pà-ɛ́: I saw her first.

64 T-y: After he saw her he said he'd give her to me if I wanted her! Right there
on the road [was where he said that]!

65 Càbo: Then whoever saw her first, that fellow gets the good-looking one. As
for the inferior ones, the ugly ones, we'll give them all to Jakeh and the Little
Red Savage\footnote{Cà-qu-ní "Mr. Red + Naked." This is Thû-yì's nickname. [CHECK - who is Cà-qɛ́? Is that somebody's nickname, too? Maybe they both refer to T-y]} here!

\begin{center}
[Laughter]
\end{center}

\leftskip=0pt
66 Pà-ɛ́: The Kachin said he'd take one like that.

67 T-y: Well, whoever wants that kind can have them.{\small{}24} Pà-ɛ́, let's
you and me go visiting someplace else! It's not as if there were no other fish
in the sea,{\small{}25} you know. Otherwise,{\small{}26} no matter how nicely we
discussed it, we'd have to come to blows, friends that we all are.

68 Pà-ɛ́: There are some over around PÁ-ló and PÁ-màʔ.

69 H: There are millions of them, I bet, when it comes to that.

70 Pà-ɛ́: Plenty of 'em, plenty of 'em.

71 H: Oh, so you're very well informed, are you?\footnote{"Do you know very well?"}

72 Pà-ɛ́: I know whereof I speak!\footnote{"I really know!"}

72b       : I'll let\footnote{All that is decipherable of this (probably obscene) remark is the Pvf expressing probability, nɛ̀-\emph{ɔ}.} ...

73 Kachin: I know a thing about it too.

74 Pà-ɛ́: Since there are pretty ones \textit{and} ugly ones--

75 Càbo: Well, if you observe how a female\footnote{ɔ̀-s\emph{ɛ}-ma, lit. 'a female body.'} does her work, and her behavior\footnote{<tâ-hêʔ-tâ-càʔ> < Shan; càʔ ult < Bs.},
then you'll know!--whether she has a kood\footnote{Càbo says tàʔ instead of dàʔ, imitating the Kachin's accent.} character or not.

76 H: Look carefully [for a wife], my boys! It's not as if [you were buying] a
basket\footnote{<KhÁ-pù khÁ-lu> "baskets" CHECK.} or something, you know. You don't get one today and throw it away tomorrow!\footnote{hə̂ʔ means both 'get' and 'marry.' bà means both 'throw away' and 'divorce.'}

77 Pà-ɛ́: They \textit{are} like baskets!

78 H: According to the wise old saying, [a wife] is the "flower of the
hearth and home."\footnote{<yɛ̀-vêʔ-qa-vêʔ>.}

78a Càbo: [inaudible]

79 H: [If you marry (inaudible) someone unsuitable] you're liable to get tired
of being married to her. So watch your step!

80 Pà-ɛ́: If she doesn't have a good character then she's an old basket.\footnote{<KhÁ-pù khÁ-lu> "baskets" CHECK.}
If she does have a good character, she's not a basket.

81 H: Hm, well, then, why don't you fellows stop this business now. Let me tell
you a story. Just listen. How about it?

82    : O.K.

