
[Sermon: Pastor Cà-bo]

1. What we \{must have/need\} is people who help others, and people who are humble
in spirit.

2. We must be able to have patience.

3. We must be tolerant and broadminded.

4. We must not say unrighteous or unjust things.

5. Thus we have been taught.

6. However, let us carefully consider what it says in the Bible about this:

7. Even though all of us humans say this---although God has been teaching us his
word, and is continuing to \{guide/lead\} us, why is it that we humans can say
such ugly things?

8. Why are we capable of uttering such things?

9. Let's think about this now.

10. When I think about it, I am---that's the way it is, right?---\{being/ acting
as\} the leader who is guiding your whole \{group/community\}.

11. But your friends, the brethren on the committee \footnote{kɔ̄mītī: a group of respected villagers consulted by the headman and pastor on matters of importance.} \{are also like that/play
a similar role\}.

12. \{They/We\} are the ones who help the community, who guide it, who help people.

13. However, do not [go so far as to] say that we in leadership positions are guiding
and helping others.

14. Even we for our part have in full measure done unrighteous things, been proud
and haughty, gone against God's will, and have lacked \{patience/humility\}.

15. Therefore, since we \{behave/are\} this way, all those who follow us also have
now <seen> their own strength \{grow less/diminish\}.

16. For this reason all of us Christians must think things over.

17. The disciple Paul says, ``We \{must have/need\} in \{the/our\} community all
the people who look after, care for, and help [others], those who are tolerant
and broad-minded.''

18. Anything which would break up and divide our community should not be considered
by us. \footnote{Evidently serious disputes within \{Huey Tat/the village\} had led to a threat that some households would split off and form a village of their own (or join a different pre-existing one).}

19. We should not consider it.

20. We must not consider things which would bring us to ruin and destruction.

21. No matter how others may treat us, no matter how they may scold us, whatever
they may say to us, we must remain humble and patient.

22. We haven't managed to accomplish [this] yet.

22a. Let us consider why we brethren haven't yet managed to do that.

23. For what reason can't we do it?

24. In a certain place in the Bible it says ``your heart, your soul, your strength
and \{spirit/ health\} \{everything belongs to/all belong\} to God.''

25. Give it all up to God!

26. Yet for us human beings it's \textit{my} heart, soul, strength and \{spirit/health\}
\textit{my} wealth and property, \textit{my }household---it's all mine and I myself
control it.

27. It's because of this [attitude] that we have not succeeded.

28. What does it say in another part of the Bible about this ``I myself control
it''?

29. ``In your heart there is a sacred dwelling place \footnote{chɛ̂-kɨ̀ bo-yɛ̀: `dwelling' + `holy house, temple'} for the holy spirit.''
[Corinthians 3:16, 6:19 -- note by DB]

30. Do not let it be a dwelling place for \{Satan/the evil spirit\}

31. In that second part of the Bible it says that.

32. But is our dwelling [i.e., body] place the temple of God's Holy Spirit?

33. This body of ours is Satan's temple, Satan \{rules/controls\} it, we have allowed
Satan to rule it.

34. Since I all by myself cannot take care [of things], I let Satan rule me.

35. And when he let him rule, and when Satan takes care of us, when he ``improves''
us in his image\footnote{This translation is meant to convey the ironic tone of qɔ̀ʔ gu la ve tê yân thâ `when he then \{improves/repairs\} [us]': \textbf{gu} (V) `prepare, practice; mend, fix, repair; be cured'.}, what happens?

36. We quarrel and bicker with each other, we don't get along harmoniously, we
act contrary to God's will, we cut ourselves off from what God has ordained.

37. That's how it has come to be within the hearts of us human beings.

38. Besides this happening, now it's become nothing but quarreling and bickering,
fighting and squabbling.

39. Today we'd like to consider this disharmony, people not listening to what anybody
else is saying, all these things.

40. Let each and every one of us consider this.

41. If all this is according to God's will, then \{fine/ all well and good\}.

42. If it's according to God's will then we can continue this way in the future.

43. But if it's not according to God's will, once we have walked on the \{path/road\}
that is contrary to God's will, we will \{be destroyed/come to grief\}.

44. We will be ruined and destroyed.

45. Besides being ruined and destroyed, we will not be cared for by God, and you
can't put your trust in \{Satan/the Evil Spirit\} either.

46. The Evil Spirit will not help you, and God won't \{care for/take care of\}
you either.

47. Our community will not make progress, will not prosper in the future.

48. These are the things we'll [have to] face.

49. We who are Christians are all \{capable of sin\}.

50. There is no one who is not a sinner.

51. However, among God's words it is said that one ought to respect during the
time that has been granted for sinners to repent.

52. We should repent \{properly/in the right way\}.

53. My brethren, let us think over everything that has happened recently.

53a. Let us consider this with extra special attention.

54. What was it that we put our trust in that has made us become like this---was
it because we did put our trust in God's word that we became like this, or was
it because we did \textit{not} trust in God's words that we became like this?

55. If we all had truly put our trust only in God's words, we could not have become
like this.

56. We could not have become like this.

57. During the past two years, because we loved and esteemed each other, \{happily/cheerfully\}
helped one another, and listened to what each other was saying, we were able to
do God's work, that's how it was.

58. The reason is, during those two years, even though we didn't give everything
[we owed] to God, we gave him our hearts, so \{it could be that way/we got along
fine\}.

59. But now we can't accomplish anything anymore.

60. What it says in another passage, in Philippians, what does it say?

61. We who have already become Christians should all have our thoughts, beliefs,
actions, and hearts \footnote{dɔ̂-qhâ, tā-qhâ, te-qhâ, ni-ma:  think-path nom     act        heart  \textbf{ta} is glossed in field-notes as `thinking, believing', and said to be < Shan thān. But this morpheme does not appear in DL.} on a single path. (Philippians 2:1-2 -- DB)

62. And our hearts \footnote{tê caɨ tí: `one heart only'. Cà-bo here uses the Shan word caɨ `heart' instead of native Lahu ni-ma, interestingly preserving the Shan centralizing diphthong -aɨ (which has merged in Siamese with -ai, but which is still retained in Thai orthography).} should be united in love.

63. What do you think, brethren, are there several different ways of loving?

64. There are not.

65. There is only one.

66. As for beliefs, are there several different ways to believe?

67. There are not.

68. There is only one.

69. We received but a single one.

70. But when we \{think about/consider\} this matter, there are many problems.

71. We Christians do not speak harmoniously one with the other.

72. It's gone so far that even if one person simply insults somebody else that
\{sin/offense\} is never forgiven.

73. Think about it.

74. Jesus, the son of God, was totally free from sin.

75. He sacrificed himself upon the cross for your sins.

76. People spat upon him

77. They beat him, they reviled him, they cursed him, until they killed him on
the cross.

78. What did Jesus say?

79. ``Father, forgive them, for they know not what they do,'' \footnote{More literally, ``Father, they, not knowing, do this to me; forgive their sin.''} Jesus said. (Luke
23:34 -- DB)

80. And yet we human beings, even if we have only been insulted a little bit, we
cannot forgive that \{sin/offense\}.

81. How much this must sadden the heart of God!

82. Everybody read [this] in the Bible and think it over!

83. Our community going and splitting up is not according to God's will. We cannot
be content this way.

84. The thing is, we must love one another.

85. In another passage in John I \footnote{Couldn't find it either in the Gospel of St. John, or in chapter 5 of The First Epistle general of John (Note: 1 John 4:11 -- DB)}, what does it say?

86. In Chapter V, ``Dearly beloved, the Lord Jesus loves us so very deeply that
we ought to love one another.''

87. Jesus would say that.

88. The way Jesus loved, that's how we ought to love one another.

89. If we do not love each other, in the future how will we be able to do the work
that will \{bring about God's kingdom on earth/``make God's country progress''\}?

90. Brethren, there is nobody living on earth who does not make mistakes.

91. [We] make mistakes.

92. But even though we are prone to err, while God has given us the time to repent,
we should all repeat and \{get on with things/continue as before\}!

93. Just as I have communicated the words of God to you, my brethren, everyone
ought to think them over and over and act on in the future.\footnote{te qay cɔ̂ : `ought to keep doing'  do go correct  Vn Vv Vv}

94. My brethren, do not just think of ideas that will ruin us and split us apart!

95. The thing is, if we entertain such ideas, besides its being extremely depressing,
there's another thing---it would be a hame for us in the eyes of others.

96. The other day we were happily listening to God's words, and together we resolved
to unite our hearts, through thick and thin\footnote{hā-thɔ̂-ša-thɔ̂: lit. ``difficult-even-easy-even.''  Vadj Punf Vadj Punf}, in spite of how angry we might
have been with each other, and we had made progress up to today.

97. \{But/So\} today it's all of a sudden like this, and our community has the
idea of ``let's split up,'' which is inconceivable! \footnote{mâ dɔ̂ tɔ̀ʔ: ``not think emerge''  NEG V Vv}

98. Therefore, in the future everybody should think this over carefully.

99. When you ``intend to act,'' think carefully.

100. When you ``intend to eat,'' sniff first.

101. There is still time to somehow get better, to somehow have an easier time,
to somehow come up in the world in the future.

102. We're not dead yet.

103. God has not yet called us back to him.

104. Within the time that has been allotted to us, let all of us think things over
and continue as before.

105. Today this is all that I have prepared about the words of God.

***

106. If there is anything that the headman or members of the committee want to
let \{everybody/the whole congregation\} hear about, there is time.

