
86b [continued from \#85]

Sermon: the recent disharmony in our village

\. As for Hell \footnote{\textbf{nâʔ-hòʔ}: \texttt{<} N. Thai (cf. Si. \textbf{narók}) \texttt{<}}, how can it possibly be mixed with the Light?

\. In the very same way, we who are Christians are the People of Light.

\. You are all people of the light.

\. You are the salt of the earth.

\. As it says in Matthew Chapter VI, \footnote{Actually, this is Matthew 5:13. ``Ye are the salt of the earth: but if the} you are the salt of the earth.

\. And in another place it says that you are also the light of the earth.

\. Just think about it, shall we.

\. Is it good that we Christians split up and go our separate ways?

\. Is it good that we live in disharmony with each other?

1\. Everyone without exception, all of us, must think this over.

1\. \{Listen properly to how we ought to act./ How should we act? Listen carefully.\}

1\. You are the Temple \footnote{\textbf{bo-yɛ̀-ló}: lit. ``great house of grace''. \textbf{bo-yɛ̀} `church'} of the Living God.

1\. We are all temples of the living Holy Spirit.

1\. What does it say about our being temples in the Holy Bible?

1\. You are the temples of God's pure Holy Spirit.

1\. Do not allow anything evil or impure to enter your hearts.

1\. Thus it is \{written/ said\}.

1\. Therefore every one of us must \{take to heart/ think over\} \{the word of
God/ God's words\} and while there is still time we must repent our sins.

1\. God has granted us the time to do His work.

2\. Therefore everyone, while reading the Bible, remember God's words and return
to the right path! \footnote{\textbf{qɔ   phûʔ   qhɔ̀ʔ  la  lò}: lit. ``come turning back around}

2\. If we act from now on according to God's will, loving and cherishing one another,
getting along with and understanding each other, it will be good.

2\. If we now split up into little groups here and there what will become of us?

2\. \{As an analogy/for example\}, if we burn together four of five logs of wood,
when they burn there will be abundant flames.

2\. However, if we only burn two together, the flames will not be very bright,
and there will be very few of them, won't there!

2\. But if people take even these two logs and split them into separate places,
the flames will disappear altogether.

2\. In the very same way, if the community of Christians should split up and go
their separate ways, quarreling and bickering with each other, we are also splitting
ourselves apart from God.

2\. We would be violating and offending against all the words of God.

2\. Among us would arise all sorts of things that are contrary to God's will.

2\. And even though we are said to be the light of the earth, that light would
not \{emerge/ come forth\}.

3\. All the light would disappear from among us, and we would cause pain to the
heart of Jesus.

3\. Thus today am I teaching you, my brethren, about the words of God.

3\. At this point if our headman, or the committee of elders, has anything to \{discuss/
say\}, there is time for it.

3\. Isn't there anything?

