
\textbf{87 Candid conversation before Building an Airstrip}

1a: I think this is funny.

1b: Don't laugh, mother!

2: I'll be the one who asks. You people will answer.

3: [About] how the airstrip is coming along--how building the airstrip is coming
along.

4: You people are now making a landing place for airplanes in Shatodo village.

5: Don't do it yet.\footnote{I.e. Don't turn the tape recorder on yet.} We'll talk it over among ourselves first.

6 Paul: O.K.

7: We'll talk it over first. Don't do it yet. '[We are] building an airstrip in
Shatodo village.' '[Planes with] how many engines\footnote{Lit. ``heads''.} do they say will be able to
land?' I'll ask you, all right? Then you guys will answer me, right? 'Three-engined
ones can land.'

8: Say that we're making a landing place for ones with three engines. Think carefully
[before you speak]. Remember [what to say] and do it just exactly right! Otherwise
it could turn out the way it did before with Jamu's father\footnote{The Lahu are teknonymous.} --- he got stuck\footnote{Lit. ``his words were blocked''.}
and didn't know what to say.

9: One person can say what sort of thing it is that's called an 'airplane.'\footnote{Lit. ``how it goes with an airplane''}

Dav 10: Yes, we'll say that over here, then you both should continue talking on
that subject. Now this time keep everything in mind, please! [Laughs] In a minute
we'll do it.

11: It's going in already! Listen! Into the hearing-machine.

Dav 12: I don't think it's going in yet. It's not going in yet. They're not doing
it yet.

13: Turn it on! We'll speak now.

Paul 14: Now you can do it.

15: Can we do it already?

Dav 16: Just speak the way you planned it out a while ago, O.K.?

Paul 17: Talk as loud as you can, please.

