\setcounter{footnote}{0}

[Sermon by Pastor David in Šátodū vilage, 1965]

1. Woman: Don't laugh! Be quiet!

2. Another person: Don't laugh! If you talk now it'll be recorded!\footnote{\textbf{chi} \textbf{bə̀ʔ} \textbf{te} \textbf{a} \textbf{qo} \textbf{lòʔ} \textbf{e} \textbf{ve} \textbf{yâ}: lit. ``if you do it now it will go in!''}

---------------------------------------------------------------------------------

3. Pastor David: Once upon a time there was a certain wealthy man.\footnote{\textbf{šathê} < Burmese < Pali < Skt. śre(stha) 'most splendid; preeminent'. <subscript dots in Skt. form>}

4. And he had one pig.

5. This sow was called Miss Blackie.\footnote{\textbf{na-nâʔ}: \textbf{na}- (Bn) `prefix to female names', \textbf{nâʔ} (N; Mpfx) `black'}

6. So one day when she came back after bathing in a mud-hole, when her owner saw
her, when he looked at all that mud, it wasn't good, so he took her and washed
away all that mud from her\footnote{\textbf{í-kâʔ} \textbf{chɨ̂} \textbf{bà} \textbf{pî}: ``washed away with water for her benefit.''}, and made her sparkling clean.

7. Then when he let her go again he said to her, ``From now on, don't you ever
again bathe in mud!'', the owner said.

8. But then after he let her go, she saw a mud-hole again and took a bath again
in the water---er, I mean the \textit{mud }.\footnote{The verb \textbf{hɛ́} `bathe' almost always occurs with the specifying noun \textbf{í-kâʔ} `water', so the speaker lets the word `water' slip out before correcting himself.}

9. But the owner's concern\footnote{\textbf{hàʔ-qá}: 1st syll. < Shan hak `love' (cf. Si. rák)} for her was not exhausted yet.

10. On this matter there is something in the Bible.

11. While all human beings are wallowing in sin\footnote{\textbf{vên-bá} \textbf{ɔ̀-qhɔ} \textbf{lo} \textbf{šɨ} \textbf{chɛ̂} \textbf{ve}: \textbf{šɨ} `die' can be used metaphorically to mean `persist in a wretched state'. See DL: 1232.}, the Lord Jesus Christ has
great compassion for all sinners, and therefore with His blood he washes away all
sins.

12. For He says, ``From now on let no one sin again!" Thus He says.

13. However, all of us human beings still do not heed the words of Jesus, so we'll
continue to sin in the future.

14. However, Jesus' love for all of humanity is boundless.

15. He is always waiting for them to repent their bad actions and return to Him.

16. Therefore that mud-hole means\footnote{\textbf{qôʔ} \textbf{gâ} \textbf{ve}: ``want to say,'' i.e., `mean'; cf. Fr. \textit{vouloir dire}, Bse. pyô tšin \textbf{te}, etc.} the sins of all of you!

17. As for the water, that means the blood of Jesus.

18. May all Christians who have been washed in the blood of Jesus remember what
Jesus said: ``From today onward, do not sin again!''

