
90 Laying up Riches in Heaven

Pastor Cà-bo

1. Once upon a time there was a certain woman.

2. This person was a great Tutoress \footnote{šālā-     ma-   ló. ``Tutoress'' is in the OED.  teacher-FEM-great} to a royal family .\footnote{Probably a Shan princeling or a \textit{sawbwa. }cə̂ʔ-càw < Shan.}

3. She was the one who took care of the nobles and who led them [in the path of
proper behavior].

4. Well one day (=night) she had a dream.

5. What she dreamt was that she went up to heaven!

6. When she arrived there, in this [heavenly] century she saw a very beautiful
big house.

7. So she went on, and when she got near to it she saw a person.

8. She asked this person, ``Whose house is this?''

9. The person said, ``This is the house of your servant.''

10. When she went to another place nearby, she saw a house that was even more beautiful.

11. When she asked about it, she was told ``This is the house of your servant who
takes care of the pottery.''

12. ``Well, if the houses of all my servants are as beautiful as this, mine, since
I am a great Tutoress, must be even more beautiful,'' she though.

13. Then she went on a little further.

14. When she reached a certain point, she saw in a little dry riverbed a little
tiny house thatched with ragweed .\footnote{A ragweed hut is taken as the archetype of a miserable place to live.}

15. That town/city \footnote{That surrounded the ragweed hut.} was a beautiful one.

16a. So she asked this fellow, ``Since this is such a fine town, why would [anybody]
go and build such a tiny house?''

16b. When she asked this, he said to her, ``This one is your house.''

17. She immediately got very angry.

18. ``Why is it that all my servants have such beautiful houses, while the house
built for me is only like this?'' she asked.

19. ``We built it with what you sent to us.

20. ``As for this [house], since you've only sent us so much, that's all you'll
get,'' he said.

***

21. Today, all we Christians while we are living on earth should not accumulate
wealth and property. \footnote{There was actually little danger of this in impoverished Lahu villages.}

22. To the best of our ability, in order to attain beautiful houses in heaven,
we should accumulate good deeds here on earth.

