
{\LARGE{}92 The Trickster and the Old Bonze}

{\LARGE{}1. Once upon a time, a certain old bonze [1] took the Trickster [ɛ̀-qhɛ̀]
along with him to go carry salt.}

{\LARGE{}2. The old bonze was riding a horse.}

{\LARGE{}3. As for the Trickster, he made him go on foot.}

{\LARGE{}4. Then after a while he made the Trickster carry the salt.}

{\LARGE{}5. He [the bonze] arrived at a certain place on horseback, and stopped
to rest.}

{\LARGE{}6. At this point Trickster knocked loose [2] some cow-sage beans [3] onto
the horse's saddle.}

{\LARGE{}7. Then when the two of them continued on their way and reached a certain
point, [the bonze] said, ``Hey, Trickster, are you finding it easy to carry the
salt?''}

{\LARGE{}8. The Trickster said, ``Oh, it's very easy! It's not at all tiring. And
how is it going with you?'', he asked.}

{\LARGE{}9. ``Oh, I can't ride this horse. My ass is itching like crazy. It's most
unfitting!'' he said.}

{\LARGE{}10. So then he said, ``Say, Trickster, let me carry your salt. You come
ride the horse.''}

{\LARGE{}11. Then [the bonze] went further ahead.}

{\LARGE{}12. Once he [the bonze] got further ahead, the Trickster knocked away
what he had sprinkled onto the horse, and went riding along.}

{\LARGE{}13. Riding along he reached the outskirts of a strange village [4], and
stayed there [waiting for the bonze to catch up].}

{\LARGE{}14. By the time the old bonze carrying the salt also reached the village,
night had fallen.}

{\LARGE{}15. When it got dark he said to Trickster, ``How shall the two of us hide
away this salt?'' [5]}

{\LARGE{}16. Trickster said, ``Let's hide it down there inside that pond!''}

{\LARGE{}17. Early the next morning, when the two of them went to look for the
salt they had hidden, they couldn't find it.}

{\LARGE{}18. It had all become [dissolved in the] water.}

{\LARGE{}19. He said to the Trickster, ``ɛ̀-qhɛ̀, what could have happened!
I can't find the salt!''}

{\LARGE{}20. Trickster said, ``The fish have eaten it, the fish have eaten it!
Let's hurry and drain this pond.''}

{\LARGE{}21. As they were draining it, some pa-nÁ-tɔ̄ [6] came out, and Trickster
slapped them to death.}

{\LARGE{}22. When some catfish [7] emerged, [Trickster] made [the bonze] slap them
[too]. [8]}

{\LARGE{}23. So this old bonze suffered a world of pain.}

{\LARGE{}24. Since he hurt so much, he said to Trickster, ``You're going to have
to carry me.  If you don't carry me I can't make it.''}

{\LARGE{}25. Thereupon Trickster said, ``Take it easy, take it easy! I'll go look
for somebody to carry you.''}

{\LARGE{}26. So he wound his way upwards and dislodged a rock so it would roll
downhill.}

{\LARGE{}27. When [Trickster] said, ``Run! If it's not a tiger, it's a bear! Run!''
the old bonze ran all the way back home.}

{\LARGE{}[1] I.e., a Mahayana Buddhist monk. The word comes from French }{\LARGE{}\textit{bonze}}{\LARGE{}
< Portuguese }{\LARGE{}\textit{bonzo}}{\LARGE{} < Jse. }{\LARGE{}\textit{bonsō}}{\LARGE{}
< Chinese, ult. < Sanskrit.}

{\LARGE{}[2] dɔ̂ʔ-pū: `sprinkle by striking' (``strike-pour'').}

{\LARGE{}[3] phɨ̂-(h)Á-le `cowsage; horse-eye bean' [}{\LARGE{}\textit{Mucuna
pruriens}}{\LARGE{}] a creeper with fuzzy brown beans containing an itchy substance.
See DL:921.}

{\LARGE{}[4] šu qhâʔ: lit. ``other peoples' village''.}

{\LARGE{}[5] So it wouldn't be stolen during the night.}

{\LARGE{}[6] An edible black fish with a white belly, about a foot long. Also called
pa-nÁ-qāy. See DL:802.}

{\LARGE{}[7] Spiny bottom feeders [}{\LARGE{}\textit{Clarias batrachus}}{\LARGE{}].
See DL:803.}

{\LARGE{}[8] Thus lacerating his hand on the spines.}


