\setcounter{footnote}{0}

1. Once there was a certain lord.

2. Under this lord there was a high minister.\footnote{\textbf{a-màʔ}: < Burmese (cf. WB \textbf{ʔamat}) `minister; vizier; high official'.}

3. One day during the time this minister was in the lord's service, while he was
staying close\footnote{`While staying close' translates the reduplication of \textbf{chɛ̂} (V) `stay; be in a place; dwell'.} by him, one of

his [the lord's] middle fingers rotted and broke off.\footnote{The Lahu verb-particle \textbf{šē} (called šēr or the ``\textbf{šē} of regret'' in GL) here signifies that the speaker regards the verbal event as unfortunate. It is normally left untranslated.}

4. When this middle finger rotted off he showed it [his hand] to the minister.

5. ``Minister, one of my fingers, one of my middle fingers has rotted off.

6. What shall I do? Come here, and look!''

7. When he said this, and when the minister came and was shown it, [he] said, ``My
lord, everything that happens

is for the best.''\footnote{\textbf{ɔ̀-bo} \textbf{ɨ̄} \textbf{ve}: lit. ``the advantage/grace'' is great. For a discussion of the different nuances of \textbf{ɔ̀-bo(n}) in the animist, Christian, and Buddhist contexts, see DL 186-7.}

5. So the lord immediately became angry, and since he was so furious, while he
was thinking [about what to do],

he took [the Minister] along far into a forest, where he [the lord] saw a place
where there was a deep pit in

the ground.

6. When he saw this place, he said to the minister, ``Come here quickly, come here
quickly! I wonder what that is over there!''

7. And the minister rushed to come, and while he was straining to see, [the lord]
slowly pushed him so that he fell into [the pit].

8. After the minister had entered the pit, the lord continued on to another place,
where he found himself face to face with a large band of naked savages who ate
human flesh.

8. When they saw him they wanted to eat his flesh, but as they were about to kill
him, they said ``Take a good

look at everything about this person that is different from other people\footnote{\textbf{šu} \textbf{lɔ} \textbf{mâ} \textbf{šū}: ``not be the same as others' situation''. This expression recurs in the common phrase \textbf{šu} \textbf{lɔ} \textbf{šū} \textbf{la} `come up in the world; be as respectable as others'.} --
one of his middle fingers is missing.

9. So we can't eat him!'', and when this happened, all of a sudden [the lord] realized
something -- he recognized

his good fortune.

10. So since he came to realize [the truth of] the minister's [words] ``Everything
that happens is for the best,''

he straightaway turned back for home and had the minister pulled out [of the pit].
\footnote{Four-verb concatenation: \textbf{qɔ̀ʔ} (vV) 'back' + \textbf{g̈ɔ̀} (vV) 'vigorous action' + \textbf{tɔ̂ʔ} (V\textsubscript{h}) 'emerge' + \textbf{pî} (Vv) 'causative'.}

11. If we say ``What meaning can we extract from this?'', everything that has ever
happened, all the things

we complained and bickered about, are all for the best!\footnote{The high minister has the same optimistic view of the world as Pangloss in Voltaire's \textit{Candide}!}

12. \direct{Somebody says skeptically} : I suppose so.\footnote{The narrator ends with the emphatic final unrestricted particle \textbf{mɛ̄} after the declarative Puf \textbf{yò}. The skeptical heckler substitutes the suppositional Puf \textbf{nɛ̀-ɔ̄} for \textbf{mɛ̄}.}

