\setcounter{footnote}{0}

1. Once upon a time there was a certain person called Eh-Hk'eh (ɛ̀-qhɛ̀).

2. Day after day he would fish out his pecker\footnote{Since Lahu has only a single word for the sexual organs (\textbf{nī} `penis', \textbf{cha} `vagina') it is a problem choosing among the myriad English synonyms to find the ones that works best in a given context. I here choose to translate \textbf{nī} as 'pecker', both because it fits the jocular mood of the narrative, and because it suggests a connection with the episode of the \textit{bird} later in the story.} and trap insects\footnote{It is not clear what aspect of the Trickster's organ was so dangerously attractive to insects.} with it.

3. When a fly would come within reach he would cover it up under his clothes [to
trap it].\footnote{\textbf{šôʔ-bɛ} (V) lit. ``press-cover''; `cover an exposed part with one's clothing'.}

4. Once it happened to be a honeybee.

5. When the honeybee happened along, he tried to cover it up under his clothes,
and it stung him.

6. When he got stung, because where he tried to cover it up and where it stung
him was the same place, he cursed it, saying, ``This bee is a wretched foul bee!''
\footnote{lù...kɨ̂ is an elaborate couplet (\textbf{lù} 'be ruined', \textbf{kɨ̂} 'be rotten', occurring in a large number of similar expressions, e.g. \textbf{yâ-lù-yâ-kɨ̂} 'little brat' (\textbf{yâ} 'child'); \textbf{chɔ-lù-chɔ-kɨ̂} 'miserable person'.}

7. So after that he whipped out his pecker again and managed to catch some insects
[with it], then he went and set them free in a place where chickens were being
fed.

8. Right away the chickens pecked at [the trapped bugs] and nibbled at them.\footnote{The painfulness of this experience for the Trickster is emphasized by \textbf{šē} (Pv) `regretful state of affairs'.}

9. When the chickens pecked and nibbled at him he sobbed bitterly, that guy.

10. So they gave him one chicken [as compensation].

11. After he got the chicken he then went to a place where pigs were fed, and proceeded
\footnote{`Proceeded' is an attempt to translate \textbf{qɔ̀ʔ} (vV) `go on and do something else'. See sense \#5 of \textbf{qɔ̀ʔ} in DL:259.} to let it [the chicken] go free.

12. So then the pigs bit it to death.

13. When they bit it to death, he sobbed bitterly.

14. Since he was sobbing so much, they proceeded to give him a pig---they [just
gave it] to him.

15. So leading the pig, he went and set it free in somebody's cattle-pen.

16. When he set it free in there, the cattle kicked it the hell\footnote{`The hell' translates \textbf{g̈ɔ̀} (vV) `vivid/violent action'.} to death.

17. So then he sat down there and sobbed with all his might.

18. Since he was crying [so much] they gave him a cow.

19. Leading this cow, this time he went and set it free in an elephant corral.

20. So then the elephant(s) killed that cow of his.

21. Whereupon he sat down there and cried his eyes out.

22. Since he was crying [so much] they finally gave him an elephant.

23. Then, leading this elephant he went and let it go free in a potter's workshop.

24. His elephant stepped on the pots and shattered them.\footnote{\textbf{nâʔ-khô} `shatter by stepping on'}

25. Since it had trampled his pots to pieces, the other guy [the potter] \{took/got\}
his elephant.

26. So then he sat down there and wept bitterly.

27. While he was crying, a bird came and called out to him.

28. What that bird was calling was \textbf{kru-g̈âʔ kru-g̈âʔ}.\footnote{This imitative sequence kr- does not occur in any other Lahu word. This binome is also used as a noun, to signify a species of dove-like bird. See DL:362.}

29. So then he let fly with a potsherd, and he got it.

30. He caught it, plucked it and roasted it, and it was very fat.

31. Since it was so fat, he smeared the fat onto his pecker.

32. As soon as he had smeared it on, his pecker began to sing.

33. It sang \textbf{kru-g̈âʔ kru-g̈âʔ}, and no matter what he did it didn't
get better.\footnote{I.e., he couldn't stop his penis from chirping.}

34. But then, when he scorched [his penis] with the [hot] potsherd, he was cured.

35. So then he took along some of that bird-fat and went to a place where a princess
was doing some weaving.

36. Then after he had smeared some of that bird-fat onto her stool\footnote{The raconteur leaves out any mention of the princess sitting down on her weaving-stool, but that is taken for granted.}, her pussy
began to sing.

37. When her pussy sang, it just sang \textbf{kru-g̈âʔ kru-g̈âʔ}.

38. So then they sought all the physicians in the city to have them cure her.

39. [But] no matter who treated her, she did not get better.

40. Whoever it was who treated her, she was not cured.

41. Finally they said to ɛ̀-qhɛ̀, ``Oh, if you can cure this woman, take her
for your wife!

42. ``And furthermore, if you can cure her, you will be given half of the kingdom,''
they said to him.

43. So then ɛ̀-qhɛ̀ went and scorched [her] with that hot potsherd, and she
was cured.

44. She was cured, so he got the princess, and also got control of half the country.

