
{\Large{}\textbf{97 The Lion and the Frog}}

{\Large{}1. Once upon a time there was a certain country. }

{\Large{}2. In this country there was a great forest. }

{\Large{}3. In this great forest there was an abundance of all beasts and animals.
}

{\Large{}4. Above all these there was one animal called the lion who was the greatest
by far. }

{\Large{}5. Well, one day the lion was roaming around in that country. }

{\Large{}6. Wandering around and around he said, ``In this country there
is no one who is greater or higher than I am! There is no one who can overcome
me--not a single person!``, so he said. }

{\Large{}7. When a frog heard him saying this, the frog said, ``What are
you wandering around for?``}

{\Large{}8. When the lion heard this he thought, ``Well, here I am so great
and I'm talked to like that!`` }

{\Large{}9. And he got very angry. }

{\Large{}10. Since he was so angry he went to see who had addressed him thus. }

{\Large{}11. When he went to look, there was a big lake. }

{\Large{}12. In it there was growing a water lily. }

{\Large{}13. Although somebody was talking to him, the lion could not find him.
}

{\Large{}14. When he had gotten near, the frog said again, ``What are you
walking around like that for?``}

{\Large{}15. Then the lion became truly furious. }

{\Large{}16. He rushed violently forward. }

{\Large{}17. When he got to the place, there was a frog sitting on top of a lotus
leaf and talking to him. }

{\Large{}18. When the lion tried to bite him, all of a sudden he stumbled into
the water. }

{\Large{}19. There he roared with all his might, with all his might he bellowed,
so his heart burst and he died. }


