\setcounter{footnote}{0}

1. Once there was a certain man.

2. He was a tobacco-smoker.

3. So he had all kinds of tobacco seedlings.\footnote{\textbf{šú-yè}: `tobacco seedling'. This word was inaccurately glossed as `dried and cut tobacco ready for smoking' in JAM 1969:175.}

4. Well, he went looking for [a certain kind of] tobacco at a Shan's.

5. Then he said to the Shan, ``Mister, don't you have any šú ɔ̀-ɛ́?''\footnote{The Lahu does not know the Shan word for `cheroot, cigarillo', so he uses the Lahu periphrasis \textbf{šú} \textbf{ɔ̀-ɛ́} `little tobacco'. He has some of them wrapped up (with other belongings) in a large package, in case it should prove necessary to show the Shan just what kind he wants, but he has not unwrapped it yet .}

6. At this the Shan said, ``What is this `šú ɔ̀-ɛ́'?''\footnote{In this sentence, and in the rest of the text, the Shan speaks perfectly good Lahu. This is just a narrative convention, and the hearer is supposed to assume that he is really speaking Shan throughout. Cf. the convention in American movies where foreigners who are presumably talking together in their own language are made to speak in English so the audience will understand, but with a foreign accent, so that an exotic distance will be maintained.}, so he repeated
``šú ɔ̀-ɛ́''.

7. Then the Shan said to him,\footnote{The words \textbf{Pî-chɔ̂} \textbf{yɔ̂} \textbf{àʔ} `the Shan (said) to him' are tacked onto the end of the sentence as an afterthought. Such inversions are common in colloquial Lahu.} ``Oh!\footnote{The Shan now thinks he understands, and recoils in horror. The Lahu word \textbf{šú} `tobacco' sounds something like the Shan [shə̌] `tiger'. Perhaps the Shan understands the word \textbf{ɔ̀-ɛ́} `little thing', which would lend plausibility to his hypothesis that there was a `little tiger' or tiger-cub in the Lahu's package, since a big one would not fit. This is not essential to the story, however, since the sequence of vowels \textbf{u-ɔ-ɛ} in \textbf{šú} \textbf{ɔ̀-ɛ́}, when rapidly pronounced, sounds very much like the schwa vocalism of the Shan form.} In that case don't up wrap it yet, don't
unwrap it yet!

8. I'll go look for a policeman first!''

9. Then, since the Shan hadn't understood, he unwrapped it and put it down [for
him to see] --- he'd been talking about tobacco seedlings!

