\setcounter{footnote}{0}

ɔ̀-pū ve ɔ̄-kâʔ

1. qhe-qo tê pɔ̂ʔ thâ chɔ tê g̈â cɔ̀.

2. šā-mɔ́ ló yò.

3. yɔ̂ ɔ̄-qhe šā tí bɔ̂ʔ chɛ̂ ve yò.

4. šā ɔ̀-phû tí câ chɛ̂ ve yò.

5. tê ni pə̀ tê ni yɔ̂ šā tí bɔ̂ʔ chɛ̂ ve yò.

6. tê ni thâ gà e lɛ yɔ̂ qay-qay lɛ ô vɨ̂-kɨ̀ ɔ̄ yɛ̀-mí-twɛ̄
tê khɛ g̈a ve yò.

7. yɛ̀-mí-twɛ̄ tê khɛ g̈a lɛ, yɔ̂ pû mâ g̈â ve pa-tɔ, ɔ̀-gɨ̀-qú
qhɛ̀ʔ lɛ ô á-chû-pu

mú-dɔ àʔ phɛ-hú á ve cê.

8. qhe-te-lɛ yɔ̂ qhɛ̂ qò á lɛ qòʔ e ve yò.

9. qòʔ e lɛ yɔ̂ ve ɔ̀-má-pā àʔ qôʔ pi ve, "má ò, ô ɔ̄
yɛ̀-mí-twɛ̄ ɔ̀-gɨ̀-qú ca pû e," tɛ̀ qôʔ pî ve yò.

10. qhe-te-lɛ ɔ̀-má-pā qay ve yò.

11. yàʔ-qhâ qhà-thâʔ-kàʔ mâ qɔ̀ʔ la mɔ̀.

12. ɔ̀-má-pā chi ca ni lɛ á-chû-pu mú-dɔ àʔ cɨ́ tā ve yɛ̀-gɨ̀-qú
àʔ g̈â-thèʔ pû chɛ̂ ve lɔ.

13. qhe-qo á-chû cɨ̀ a qo, ``nɔ̀ chèʔ qo ŋà kàʔ chèʔ!''
qôʔ lɛ, yɔ̂ ô kàʔ tí qɔ̀ʔ chèʔ phɛ̂ chɛ̂ ve yò.

14. qhe-te-lɛ ɔ̀-pū à-šɛ-thâ qhɛ̂ qò á kɨ̀ ɔ̄ gà lɛ, ɔ̀-má-pā
ô ve tí g̈ɔ̂ʔ câ chɛ̂ ve thàʔ g̈a mɔ̀ ve lɔ qôʔ-ma.

15. ``má ò, nɔ̀ à-thòʔ-ma te chɛ̂ ve le?'' qôʔ a lɛ,
"alôo, à-pū ò, yàʔ-ni lɛ̀ nɔ̀

ɔ̄-kâʔ â tɛ á lâ qo ŋà šɨ e yò," qhe qôʔ pî ve yò.

1. Well, once there was a certain person.

2. He was an expert hunter.

3. All he ever did was hunt animals.

4. He earned his living only by the money he got from hunting.

5. Day after day all he did was hunt.

6. One day he walked and walked and caught a bear in a distant place.

7. Having caught this bear, since he couldn't carry it, he stripped off its skin
and tied it up to dry on the stump of a thornbush.

8. Then he took a crap and went home.

9. After he got home he said to his son-in-law, "Son-in-law, go and carry
back the bearskin over there!"

10. So the son-in-law went off.

11. But nobody saw him come back.

12. The son-in-law had gone to look, and was struggling to carry the bearskin that
had thorns stuck to it.

13. So when the thorns pricked him he said, "If you bite me, I'll bite
you too!",

and he kept on biting them back violently.

14. Then when the father-in-law got to the place where he had taken a crap, he saw
his son-in-law eagerly eating it!

15. ``Son-in-law, what are you doing?'' he said, so the son-in-law
said, "My goodness, father-in-law, if you hadn't left this cold rice for
me I would have died of hunger!"

