\setcounter{footnote}{0}

1. Well, now, once there were two men.

2. These two men both went into the jungle, and caught a monkey--they had gone monkey-hunting.

3. After they caught the monkey they went to sell the monkey-meat to a Shan.

4. The Shan said: ``šaháy, nə̀ i-šá ɛ̀?''

5. That means ``Friend, what kind of meat is it?''\footnote{Presumably the Lahu either didn't have the meat with them, or else had butchered it beyond recognition.}

6. Now the two of them didn't know the Shan language at all.

7. So when they were spoken to, they kept quiet.\footnote{\textbf{tàʔ-í} \textbf{chɛ̂} 'to be silent.' \textbf{tàʔ-í} 'silently' occurs only with the verb \textbf{chɛ̂} 'remain, keep doing something, stay.'}

8. They didn't know what to say.

9. ``You, you tell him!'' they said (to each other), and they remained
silent.

10. Well the Shan kept asking them again and again, ``nə̀ i-šá ɛ̀?''

11. Finally,\footnote{\textbf{tê} \textbf{khɨ} \textbf{gà} \textbf{e} \textbf{lɛ}: lit: ``when a time had arrived.''} they said ``Meat live in tree, meat live in tree!''
\footnote{\textbf{nə̀} \textbf{yú} mài: i.e., 'the meat of a kind of animal that lives in a tree.' The Lahu word \textbf{šā} means both 'animal' and 'meat', but the Shan words are different (\textbf{nə̀} 'meat', sat 'animal'). The Lahu only knew the word for 'meat.'}

12. But the Shan didn't understand.

13. He just kept saying ``nə̀ i-šá ɛ̀?''

13. ``Meat live in tree'' they said.

14. Well, he couldn't understand, he just couldn't, so finally one of the men [jumped
up and] swung from

a beam in the Shan's house while the other one fixed an arrow to his bow\footnote{\textbf{khâʔ} \textbf{thɔ} \textbf{ve} 'put an arrow in a bow, cock a bow.'} and
aimed it at him.

15. Whereupon the Shan had to beg them [for mercy].

16. ``Now I understand, now I understand!'' he said.\footnote{The Lahu were trying to make him understand by their playlet. The Shan misinterpreted their actions as showing hostility to him, and hastened to assure them that now he understood, although obviously he still did not. The Shan's last words are in Lahu, but this is a narrative convention, and is not meant to imply that he could actually speak Lahu .}

