\setcounter{footnote}{0}

1. Then there were two [other] people, once upon a time, named Qɔ̀-nɛ́ and Qɔ̀-ŋɔ́.
\footnote{\textbf{Qɔ̀-nɛ́} is the Black Lahu, and \textbf{Qò-ŋɔ́} is the Yellow Lahu. The names are humorous, since \textbf{nɛ́} and \textbf{ŋɔ́} are both verbs meaning `to get stuck' in BL. In a previous translation I called them ``Stickney'' and ``Stuckney'' respectively.}

2. One morning they were going along together, when [one of them] said:

3. ``Let's go parakeet-trapping early tomorrow morning, okay?''

4. So the next morning they went off together.

5. When they had finished laying the parakeet-traps, a parakeet came and got stuck
\footnote{The traps were some sort of contrivance on which a sticky substance (bird-lime) was smeared to capture the birds. The traps set by the two men were apparently some distance apart.} on it--on [the trap] of Qɔ̀-nɛ́.

6. Thereupon Qɔ̀-ŋɔ́ [the YL] said, ``Friend, it's stuck\footnote{\textbf{nɛ́}.} on yours!''

7. Then, another one got stuck over on [the trap] of Qɔ̀-ŋɔ́ [the BL].

8. At this, he\footnote{\textbf{Qɔ̀-nɛ́}.} said, ``Friend, there's one stuck\footnote{\textbf{ŋɔ́}.} on yours too!''

9. Then the two of them got furious at each other and began fighting.\footnote{Actually \textbf{Qɔ̀-ŋɔ́} must have started the fight, since it was he who took exception to the word \textbf{ŋɔ́}. See note to the next sentence.}

10. As they fought, they ended up down in the river--and [all because] he [the YL]
had said ``The parakeets aren't paying any attention to yours!''
\footnote{The misunderstanding arose primarily through the word \textbf{ŋɔ́}, which means `to stick' (=\textbf{nɛ́}) in BL. but `look at, pay attention to' in YL.}

