\setcounter{footnote}{0}

1. Once there were two people.

2. They went hunting.

3. These two people, father-in-law\footnote{\textbf{ɔ̀-pū}: etymologically `grandfather' (< PTB *bəw), used also to mean `father-in-law'.} and son-in-law, were hunting porcupines.

4. They arrived at an old-field\footnote{\textbf{hɛ-šā}: a field previously cleared for a swidden, now reverting to jungle as it lies fallow. These old-fields were good hunting-grounds, since the regrowing plants provided good cover for animals.} in a certain place, and the old man made his
son-in-law follow a trail in the upper part of the field, while he went into the
middle of the field.

5. Then, when the son-in-law saw some red ants going up and down on a tree stump,
he said, ``Father-in-law, they've gone up already!''\footnote{The humor here resides in the father-in-law's misunderstanding of what was ``going up'': the son-in-law meant the ants, while the father-in-law thought he meant a porcupine. Yet another demonstration of possible ambiguity in a ``pro-drop'' language like Lahu!}

6. When these red ants went up, he said, ``Father-in-law, they've gone up!''

7. When he said this, the father-in-law climbed up, exerting himself he ran up
that hill.

8. When those red ants came back down again, he shouted loudly, ``Father-in-law,
they've gone down already!,'' so the father-in-law kept on running up and running
down all the time, so he [finally] said, ``Aw, this son-in-law is no damn good!''

9. Having run up and down like that he said ``Son-in-law, lick this!''\footnote{\textbf{chi} \textbf{ve} \textbf{lɛ̀ʔ} \textbf{ôʔ}, lit. ``lick this,'' a shorter version of \textbf{ŋà} \textbf{ve} \textbf{mɔ̂} \textbf{lɛ̀ʔ} \textbf{ôʔ}, lit. ``lick my thing'', is a frequent Lahu put-down more or less equivalent to English ``kiss my ass.'' The second misunderstanding here resides in the polysemy of \textbf{lɛ̀ʔ}, etymologically `lick' ( < PTB *\textbf{m-lyak}), but often used to mean `eat', especially with respect to snack food. See DL: 1393.}, and
he went back [to his son-in-law's house], and when he got there he met up with
his son-in-law again.

10. So he said, ``Son-in-law, open up the rice that I wrapped up before.''\footnote{Father-in-law had wrapped some rice up in a banana-leaf for them to snack on during their hunt.}

11. ``But you said `Eat it!', so I ate it up,'' he said.

12. When he said this, the Father-in-law said, ``Aw, this son-in-law is no damn
good!'', and he went home [to his own house].

