\setcounter{footnote}{0}

1. Once there were two guys, one was named Crazy Man\footnote{\textbf{Cà-g̈û}: \textbf{cà}- `prefix to male names', \textbf{g̈û} `be crazy'.} and the other was named
Weasel.\footnote{\textbf{fɨ́-lâ-qō}: glossed variously as `golden marten; yellow-throated marten' or `yellow weasel.' See DL:1314.}

2. The two of them\footnote{\textbf{yɔ̂} \textbf{nî} \textbf{g̈â}: this is a colloquial variant of \textbf{yɔ̂-hɨ} \textbf{nî} \textbf{g̈â}, which is possible to use when the plurality is clear from the context, in this case the numeral + classifier \textbf{nî} \textbf{g̈â} `two people'.} were clearing a swidden when one [Crazy Man] said to the
other, ``There are hornbills, lots of them, munching on hog-plums.''\footnote{\textbf{màʔ-chu-šī}: \textit{Spondias pinnata}, a tree with edible bark and fruits. See DL:980.}

3. Weasel: ``Whereabouts?''

4. Crazy Man: ``Down there, in the plains, on the banks of the Kok River.''

5. Weasel: ``You crazy piece of trash, nobody has each heard of such a thing, hornbills
like that!''

6. Crazy Man: ``It's true! It's true that there are hornbills munching hog-plums.''

7. Weasel: ``Nobody has ever said such a thing!''

8. Crazy Man: ``I'm going to take my musket and go shoot them and come back. There
are lots and lots of them!''

9. Weasel: ``Nobody will go with you. You're lying. You crazy fool!''

10. Crazy Man [talking to somebody else]: ``Grandpa, grandpa\footnote{\textbf{lɔ̂-pū}: (1) old man (2) jocular form of address to any man. See DL:1400.}, there are hornbills
munching on the hog-plums. Come and shoot them, come shoot, come shoot, come shoot!
The hornbills are munching the hog-plums.''

11. Old man: ``Where, where? I don't see them!''

12. Crazy Man: ``Over there, in the jungle.''

***

13. Crazy Man: ``Let's go! I'm going to catch fish with a bell-net.\footnote{\textbf{ŋâ} \textbf{phɛ̂} \textbf{ve} (OV)} Down there
in the Kok River.''

14. ``Let's go, let's go, let's go, we're heading out!''

15. ``Wow, these fish are really making noise splashing around tonight!''

16. ``You're lying! I don't hear anything.''

17. ``They're making noise like a helicopter!''

18. ``Yes. As they're flying, they're landing over there where the net bird-trap\footnote{\textbf{pɛ̄-pâ}: A net set up in a narrow mountain pass, so that when a bird hits it, it rolls up and entraps the bird.}
is.''

19. ``I still can't hear!''

20. ``Those fish in the water.''

21. `` I can't see them.''

22. ``It's true! I'm going to scoop them out. I'm going to release my net right
away. Like I did once before, I'm going to get a whole shoulder bag full, my pants
\footnote{If a fisherman runs out of receptacles he might take off his trousers and stuff them full of fish, tying the ends to make a bundle.} full, my rucksack full!''

23. ``Hmm. I didn't see that you shared them with me that time!''

24. ``I was out doing it so late at night down there in the plains before I came
back home here again.''

