
\textbf{110 The 'Empty Coconuts': the Lahu and the Shan Traders}

[Tape X, Side 1]

1 Once upon a time there was a certain man. 2 One day, leading a group of children,
he went along a road. 4 As they were going along the road, they arrived at a certain
place, where they saw a group of important-looking1 Shan \{traders / merchants\}.
4 Well, there were very many of these important traders. 5 When he saw them, he
said--he asked the Shans [a question]. 6 What he said was, \texttt{"}\textasciitilde{}\textasciitilde{}\textasciitilde{}\textasciitilde{}\textasciitilde{}\texttt{"}
7 Which means \texttt{"}Sirs, friends, what are you carrying [with you]?\texttt{"}

8 At this a Shan said, \texttt{"}\textasciitilde{}\textasciitilde{}\textasciitilde{}\textasciitilde{}\textasciitilde{}.\texttt{"}

9 So then he asked again: \texttt{"}\textasciitilde{}\textasciitilde{}\textasciitilde{}\textasciitilde{}?\texttt{"}
10 That is, \texttt{"}Sir, what are you carrying? 11 Then the Shan answered him
again, \texttt{"}\textasciitilde{}\textasciitilde{}\textasciitilde{}.\texttt{"}
12 That is, the Shan was saying \texttt{"}We're not carrying anything. 13 Our saddlebags
are all empty.\texttt{"}

14 Then \{the man / he\} said to the boys: \texttt{"}Hey, boys, boys! They say
they're carrying coconuts2! 15 Wouldn't you like to eat some? 16 If you'd like
to eat some coconuts, why don't you all chip in some money3!\texttt{"} 17 So the've
got some money together. 18 When he got the money they had collected, they said
to the Shans, \texttt{"}Please sell us some coconuts.\texttt{"}

19 At this the Shan said, \texttt{"}We have nothing to sell. 20 pao.....\texttt{"}
21 That is, \texttt{"}Our saddlebags are empty4!\texttt{"}

