\setcounter{footnote}{0}

1. Once upon a time there was a certain man.

2. One day, leading a group of young people, he went along a road.

3. As they were going along the road, they arrived at a certain place, where they saw a group of important-looking\footnote{\textbf{ló} < Shan lon `great, big, important.'} Shan traders.

4. Well, there were very many of these important traders.

5. When he saw them, he said--he asked the Shans [a question].

6. What he said was, ``\textbf{pɔ-lè, šaháy, šú hàʔ šá ɛ̀}?"

7. Which means ``Sirs, friends, what are you carrying [with you]?''

8. At this a Shan said, ``\textbf{ma hàʔ šán. hàʔ pāu láj-láj ə̀.}''

9. So then he asked again: ``\textbf{pɔ-lè, hàʔ šá ɛ̀}?''

10. That is, ``Sir, what are you carrying?''

11. Then the Shan answered him again, ``\textbf{ma hàʔ šá. hàʔ pāu láj-láj ə̀}.''

12. That is, the Shan was saying ``We're not carrying anything.

13. Our saddlebags are all empty.''

14. Then he said to the boys: ``Hey, boys, boys! They say they're carrying coconuts\footnote{'The word \textbf{pāu} (also < Shan - cf. Thai \textbf{məphráaw}) means `coconut' as well as `empty.'}!

15. Wouldn't you like to eat some?

16. If you'd like to eat some coconuts, why don't you all chip in some money\footnote{\textbf{phi šɔ kə}: `collect (\textbf{š}ɔ) and put in (\textbf{kə}) money (\textbf{phi})'.}!''

17. So he collected some money.

18. When they got the money together, they said to the Shans, ``Please sell us some coconuts!''

19. At this the Shan said, ``We have nothing to sell.''

20. ``\textbf{pāu láj-láj ə̀}!''

21. That is, ``Our saddlebags are empty\footnote{The Lahu gloss he gives is slightly inaccurate. The translation of 20 is simply "They're empty."}!''

