\setcounter{footnote}{0}

\textit{<Pastor Cà-bo>} 

1. At this point I would like us to sing\footnote{Lit. ``do a hymn''. \textbf{chî-mu-khɔ̂} `hymn', lit. ``words of praise'' (\textbf{chî} `lift up', \textbf{mu} `high', \textbf{khɔ̂} `words').} a hymn.

2. This hymn is from The Acts of the Apostles, Chapter V, verses 1 to 5.\footnote{\textbf{ganân ɨ} `chapter' (``big number''); \textbf{ganân i} `verse' (``small number'')}

3. So now this hymn will be sung by two women, two mature women.\footnote{\textbf{yâ-mî-ma} `woman'; \textbf{chɔ-mɔ̂-ma} `mature woman'.}

\textit{\direct{Singers}}

4. It says in the Bible,

5. that having lied, they died, Ananias and Sapphira.\footnote{The relevant passages in the King James version are as follows: \textit{Acts of the Apostles, Chapter 5 1. But a certain man named Ananias, with Sapphira his wife, sold a possession, 2. And kept back a part of the price, his wife also being privy to it, and brought a certain part, and laid it at the apostle's feet. 3. But Peter said, ``Ananias, why hath Satan filled thine heart to lie to the Holy Ghost, and to keep back part of the price of the land?...'' 5. And Ananias hearing these words fell down and gave up the ghost...}}

6. When they were making offerings to God, the apostle asked him, ``For your land


you only got this much?

7. Is that the truth, Ananias?''

8. ``Yes, I only got this much. Truly I only got this much.''

9. The apostle said, ``Do not lie to God, Ananias!''

10. Peter said, ``Do not lie to God, O Ananias!''

11. Suddenly Ananias keeled over and died.

12. Sapphira also had talked it over with him in just this way.\footnote{I.e., she had colluded with her husband in preparing the lie.}

13. For this reason the couple\footnote{\textbf{yɔ̂} \textbf{tê} \textbf{ca}: the singular 3rd person pronoun \textbf{yɔ̂} is used (instead of the plural \textbf{yɔ̂-hɨ}) since the plurality is already indicated by the numeral-plus-classifier \textbf{tê} \textbf{ca} `a couple'.} both had to die.

14. For lying, one must truly die!

15. My brethren, for this reason let us cast aside all lies!

16. Lest we end up like Ananias!

17. The end.

