
{\LARGE{}\textbf{\#114 The Merchant and the Widow's Daughter}}

{\LARGE{}1 Once upon a time there lived an older brother and a younger brother.}

{\LARGE{}2 First the younger brother, carrying some perfume, went off to sell it.}

{\LARGE{}3 He went off to sell it, he kept on going to sell it, until he came to
a certain place, until he reached a strange village, when he happened to climb
up to a house, a widow's house. [1]}

{\LARGE{}4 A little girl was there.}

{\LARGE{}5 ``Little girl, where has your mother gone?'', he said, and she answered
``My mother isn't home. She's gone to the fields.''}

{\LARGE{}6 So then he said, ``If your mother had been at home, I'd have given her
some sparkling red and yellow balls!'' [2]}

{\LARGE{}7 Then he went away, and after he had gone to some other place he came
back, and when he went back up to that house, when he went up to it again, when
he got there, he said, ``Little girl, where has your mother gone?'', but this time
also the mother was not there.}

{\LARGE{}8 ``If your mother were at home, I'd give her some sparkling red and yellow
balls! Some sparkling red balls! If she hadn't gone away, I'd also have given her
ten pieces of silver,'' he said. [3]}

{\LARGE{}9 After that he went back home, not having sold all his perfume he returned
home, and his older brother said, ``Hand it over, hand it over! Your big brother
will have a turn going there!''}

{\LARGE{}10 So then the elder brother carried it in his turn, and set off again
to sell the perfume.}

{\LARGE{}11 He went there and said, ``Little girl!''}

{\LARGE{}12 He happened to come up to that same house, and said ``Where has your
mother gone?'', and she said ``My mother isn't home'' --- this time also when it
was the older brother's turn.}

{\LARGE{}13 So when she said ``My mother isn't home'', the older brother also said
``If your mother were only at home I'd give her some sparkling red and yellow balls!''}

{\LARGE{}14 So having said this again, after he had gone away and come back, this
time the mother was at home for him.}

{\LARGE{}15 When he had come back she said, ``A while ago you said that you'd give
me some sparkling red and yellow balls. So just try and do it!''}

{\LARGE{}16 Now he had taken some bright red fireflies at the outskirts of the
village and stuffed them into his shirt.}

{\LARGE{}17 Having stuffed them in there, he said ``Run, run, into that other room!''}

{\LARGE{}18 In the other room he unwrapped and released them and the whole house
glowed bright red, so [the mother] said, ``Oh, daughter, daughter, go get some
water to pour on it!'', [4] and the daughter got some water and when she poured
it, it happened to be hot water, and it burned that fellow's prick.}

{\LARGE{}19 So then in a flash he suddenly jumped up and went out, he ran away
and sat down on a rock, and then while he was rubbing [his sore penis] the widow
and her daughter had also left, happening to run off in the same direction where
he had fled, and the daughter said, ``Mother, mother, he's sharpening that thing
that wasn't sharp enough to stick into us just now!''}

{\LARGE{}20 That's what she said. That's what they say that she said.}

{\LARGE{}21 That's the end.}

