\setcounter{footnote}{0}

Šá-to-dō Village

1. Now I'd just like to tell you all about how the folks in Šá-to-dō used to
slaughter a pig to eat.

2. Tell us until we're satisfied!\footnote{Emendation of text from \textbf{qôʔ} \textbf{mā} \textbf{qha} \textbf{ni} \textbf{ɛ̀ʔ} to \textbf{qôʔ} \textbf{mā} \textbf{ni-qhâ} \textbf{qha} \textbf{šɨ} \textbf{ɛ̀}.}

3. Well, then, as for slaughtering pigs, we also just recently killed one. Today
when we killed another one there were a few parts that we didn't divide up among
us, so some people are probably angry.

4. That's right. A while ago a person took some pork and went home with it, but
then I saw him bring it back again.

5. That guy was not happy about it.

6. He wasn't satisfied.

7. Hey, I'm not satisfied either!

8. We shouldn't do it that way. We ought to divide it up fairly to eat.

9. Even with the one that we slaughtered the other day, the butcher took the intestines
all for himself. The head too. Even the skin was gone. Just the bones. I saw him
take half of the head for us.\footnote{'For us' translates the verb-particle \textbf{lâ} 'action benefitting a non-3rd person', here used ironically, since the butcher is benefitting only himself.}

10. How many people are there who don't agree?

11. Three people.

12. That's a lot, that's a lot!

13. That is a lot---those who don't agree.

14. Yeah, so if you're not happy about it you should have said something to us
before now. Since you didn't say anything we can't do anything about it now. Now
there aren't even any intestines left. They've all been eaten up.

15. But they---the police---told us to divide it up fairly!\footnote{The pig in question was apparently donated to the village by a local squad of the Border Police.}

16. Yes, but the insides and the head are for whoever does the slaughtering to
eat!

17. From today on we'll definitely divide up the intestines, the innards, and the
head also!

18. Never mind what we do from today onwards---right now you didn't tell us that,
so it's all your fault.

19. That's right! About their saying ``Divide it up fairly!'' when I heard what
they said to us, I disagree with you a little bit.\footnote{I.e., about the interpretation of the policeman's admonition.}

20. If you guys aren't satisfied, if things aren't according to your wishes, why
didn't you say anything when that animal was slaughtered without dividing up the
intestines?

21. When you say that now it's too late.

22. So in the future, if we're able to get another one, just divide it up equally
to eat, every single part!

23. If you just keep quarreling about it, in the future you won't get anything
to eat, you guys.

24. If we quarrel like this, when it reaches the ears of the police we won't get
anything to eat anymore, in the future!

25. Oh, let's just stop worrying about this. Even if one or two people aren't satisfied,
most people are still perfectly happy with it.

26. I wonder how much there is to eat in the innards and the head?

27. Well, it doesn't amount to very much.

28. As for the head, there's not even any meat on it.

29. As far as the feet go, we should chop them into small pieces and mix them up
thoroughly with the rest of the meat.\footnote{'Thoroughly' translates the post-head versatile verb \textbf{pə̀} 'finish V'ing; V to a conclusion'.}

30. If you guys aren't satisfied, go ask anybody you want!\footnote{Lit. ``Where you want to ask, go ask!'' I.e., other people will vouch for the fairness of our behavior.} I don't care.

31. On the head there's nothing to eat. Look over there at the one on the storage-rack.
There isn't even any skin on it.

32. Wherever you make me ask, I'll ask!

33. From now on, you guys can be the ones to do the butchering! We're not going
to kill them anymore.

34. We won't do it anymore. Starting from today. Even if we do get another one
to eat.

35. Now we \textit{won't} get any more to eat.

36. Let those who aren't satisfied do it next time, fairly. This time we managed
to do a job that we didn't want to do. We made a mistake. Since we didn't do it
right, you guys aren't satisfied.

\direct{A pause}

37. I bet you can't even divide up the intestines!

38. Each person would get such a short piece, if we divide it up.

39. We can't do it with so many people!

40. So we wouldn't even get enough to make \textit{náam phrík}.\footnote{A spicy condiment made of salty fish sauce, hot peppers, and bits of vegetables or offal.}

41. I suppose we could use it to wind around a knife-sheath\footnote{\textbf{á-thɔ꞊gú-pi} `coils (of string, bamboo-fiber, or metal) holding together the two halves of a knife sheath.}, eh?

42. I guess that's what we could do with it!

43. Come on, tell them not to have a tug-of-war\footnote{\textbf{g̈ɔ̀} \textbf{dàʔ} \textbf{ve} (lit. ``pull mutually'').} with the pig guts! They're
not long enough.

44. Pig-guts would break if you pulled them like that---they're not bear-guts,
you know!

45. Bear-guts, even more than bamboo, no matter who pulls it, it won't break.

46. Oh, we shouldn't quarrel over this so much.\footnote{\textbf{qôʔ} \textbf{dàʔ} \textbf{ve} can mean either 'talk to each other' or 'quarrel' (\textbf{dàʔ} (Pv) 'mutual action'.}

47. We'd be embarrassed before other people.

48. Since it's such a tiny amount of food, for us to talk about it at such length---if
you look at those pig-guts, you practically can't divide them up at all, just enough
for those who did the slaughtering, maybe four people.

49. Because there are lots of people.

50. With 58 people\footnote{\textbf{ŋâ} \textbf{chi} \textbf{hí} \textbf{g̈â} (\textbf{ve}): This phrase taken out of context has two totally different meanings. The sense here is ``58 people'' (\textbf{ŋâ} \textbf{chi} \textbf{hí} ``58'', \textbf{g̈â} ``classifier for people''), or ``catch this fish by competing'' (\textbf{ŋâ} ``fish'', \textbf{chi} ``this'', \textbf{hí} ``compete, vie for'', \textbf{g̈â} ``catch'').} how can you go divide it up!

51. This time it's a done deal, it's all finished, there's nothing to be done about
it. They're all eaten up.

52. Yep.

53. From today on let's divide up the intestines and the head [along with everything
else].\footnote{The phrase in square brackets translates \textbf{kàʔ} (Punf) `also'.}

54. That's what we ought to do. From today onward. There have been lots of people
who are discontented.

55. Otherwise, if we have a chance to do it again in the future, we should set
up a big \textit{wok} down there in the headman's house, and all in the same place
we should chop up all the intestines, and the meat and the feet and the head and
put it all in and then give it out for everybody to take\footnote{This is a four-verb concatenation: \textbf{qɔ̀ʔ} `perform a new action', \textbf{pî} `give', \textbf{phɛ̂} `send forth', \textbf{yù} `take'.}, and---

56. If we do it that way people wouldn't come to get their food all at the same
time---they'd be embarrassed.

57. If there are people who won't eat, the hell with them!

58. We probably wouldn't manage to divide it up.

59. Well, for my part, even if they don't give me stuff like the guts, it's all
the same to me.

60. Me too, even if I don't get to eat the guts---

61. ---I'd be embarrassed about that!

62. They stink of shit.\footnote{\textbf{nù-qhâ}: lit. ``stink-bitter'', i.e. have a bitter stench.}

63. I don't want any either. The headman says not to come divide it up in his house!
From now on he says we should divide it down there at the police station!

64. They eat people's shit\footnote{\textbf{qhɛ̂} \textbf{ò} \textbf{ve} \textasciitilde{} \textbf{qhɛ̂} \textbf{qò} \textbf{ve}: 'take a crap' (plainspoken but not vulgar).} \direct{giggles}! The inside of their
guts stinks like hell.

65. So from now on if we're lucky enough to get a chance to eat a pig again, who
will slaughter it?

66. Let's all do it together!

67. If that's what you guys do you're just bearing a grudge against us.\footnote{\textbf{châʔ} `bear a grudge'; the verb particle \textbf{lâ} `non-3rd p. beneficiary' does not specify number, but is here interpreted as plural since more than one person did the slaughtering.}

68. You should do it after talking it over with the whole community---in the future
it shouldn't be done like it was this time, not like this time.

69. If you don't agree with us, after you finish working over there\footnote{The work in question is ``Building an airstrip''. \textbf{See Text \# XYZ}}, when
it gets dark you can go down and light pine-torches and divide it up there.

70. If we can't see what we're doing, the dogs might bite into the meat and scatter
it around\footnote{This idea is concisely expressed by the two-verb sequence:\textit{ chè}ʔ `bite' + \textbf{šêʔ} `scatter'.}, a real pain in the ass. You could cut your hands with the knives.
If it's dark and you can't see.

71. If we do it that way [i.e., at night] that's what will happen to us who do
the killing.

72. With us watching, they probably wouldn't grab at the meat, if we've got torches
lit.

73. Well, a while ago when we gave out a portion of meat to this guy, even when
we sent a kid to take it to him\footnote{Evidently sending a child to deliver the meat was an attempt to mollify the hurt feelings of some villagers.}, he said, ``Don't come bringing me that, don't
bring me that! Better to give it to the dogs!'' No matter who offered it to him
he wouldn't take it. He sent it back to me. It's hanging up there, see it?! What
are we going to do with it?

74. Who is it that you're talking about?

75. Cà-gɛ̂---.

76. After dividing it up for him\footnote{\textbf{pɛ̀} \textbf{cā} \textbf{lɛ}: \textbf{pɛ̀} 'divide' +\textbf{cā} 'feed'\textit{ + l}ɛ `suspensive non-final unrestricted particle'.} if he won't eat it, I'll gobble\footnote{\textbf{vâ} `go at it vigorously'. This is a vivid pro-verb, here used instead of \textbf{câ} `eat'.} it up
myself, if that's the way it is!

77. Well, I didn't know about this! If I had known, we wouldn't have gone back
and forth aimlessly like this. Let's take it and just feed it to the dogs. Even
that would be better [than giving it to that guy].

78. Don't feed it to them. I'll tuck into it!

79. Yes, well, there's still another thing. Brothers\footnote{Lit. ``older and younger siblings.''}, are you feeling good?
You're feeling good. Is the pork tasty?

80. (Two people): It's good!

81. Well then, if we talk about what we should do in order to clean up this village
of ours, we must first sweep up all around the houses.\footnote{\textbf{á-qàʔ-á-ji}: ``outside between [the houses].''}\footnote{This sentence, as well as the four following ones, all end with the quotative Punf \textbf{cê}, the function of which here is to stress that others (the Thai authorities) have told the headman what the villagers must do to improve hygiene.}

82. We should keep sweeping the insides of our houses clean.

83. Another thing is, we should dig latrines.

84. And we should build sties for the pigs and coops for the chickens.

85. And if everybody washes all his clothes properly, we'll all be healthy and
fit.

86. I'm not going to dig a latrine yet. Not long ago somebody dug one down there,
but sand flowed into it and filled it up.

87. Yeah, then some people have their bedroom and kitchen in the same place. Sometimes
ladles and blankets are all wrapped up together.

88. That's normal for a Lahu! \direct{laughter}

89. When some people eat their meals, the dogs sit together with them and watch,
and they're eating too. I've actually seen the dogs and the people sitting back
to back and eating away!

90. That's what happens. Since those people raise dogs. I guess I've had that experience
too. You can't teach them anything, people like that.

91. Right you are! You could say it's insulting to the dog's owner. Absolutely!

92. As for a latrine, I won't dig one. I have three dogs.\footnote{Dogs are known for their coprophagic propensities.}

\direct{laughter}

93. Those who have studied [in a Thai town] ought to do it, I guess. We need latrines.

94. We ought to do it.

95. We re-a-ally\footnote{Spoken with exaggerated intonation on the final particle.} ought to do it, I suppose!

96. Even though we've done what those returning students have told us to, look
down there!\footnote{I.e., at the Thai authorities down in the plains.}

97. They're instructing us to build pigsties and chicken-coops, but once we've
built them the government doesn't bring us any [animals to put in them]. They've
been abandoned, all empty. And they're actually rotting away.

98. Maybe the time hasn't come yet, for those down there. They'll bring them to
us.

99. The government won't give them to us! I doubt they'll ever give them to us.

100. They'll probably never give anything to us.

101. It's their fault for not giving us anything.

102. If we hadn't built them [the pigsties and chicken coops] it would have been
\textit{our} fault.

103. You're right!

104. Yeah, since that's the way it is, our food and our health will probably now
be just like the way it's always been.

105. We're done.

------------------------------------------------------------------------------------------------------

