\setcounter{footnote}{0}

1. Once upon a time there was a husband and wife.

2. One day they went to wander around for fun in the forest.

3. Having roamed around a good while they got to the forest, where they met up
with a bear.

4. This bear was very ferocious.

5. When it saw the two of them, it suddenly came running.

6. At this point the husband was scared stiff and went running into a cave.

7. Meanwhile the wife stayed at the mouth of the cave.

8. Now the wife was holding onto a rattan stick.

9. When the bear went at her, the wife took the rattan stick and thrust it into
its mouth, piercing it to death.

10. Then, after it was dead, she called out to her husband, ``Come on out---it's
done\footnote{The Lahu here is \textbf{šɨ} \textbf{ò} `has already died', with no subject expressed. The husband mistakenly supplies a first person subject, thinking his wife is screaming ``I'm dying! I've been done to death!''}!'' she said.

11. The husband said, ``Oh, it has bitten the woman, it's bitten the woman, it's
bitten the woman!'' he kept saying.

12. ``Come on out---it's dead,'' she said, while he just kept saying\footnote{The pre-head versatile verb \textbf{qɔ̀ʔ} 'V again' is modifying the verb \textbf{qôʔ} 'say', which follows the quoted material.} ``It's
bitten the woman!''

13. Well, after a certain time he came out, and said to his wife, ``Now then, woman,
tell people that I was the one who killed it!'',\footnote{Lit. ``do [as if] I got him with a shot.''} he said.

14. Then the wife said, ``Cock your crossbow and shoot the mountain goat!'', she
said.

15. So he cocked his crossbow and got off a shot at the mountain goat.

16. When he fired the shot, it went too low, and didn't strike the body.

17. But this mountain goat was already very old, and when it heard the sound of
the shot it was terrified, and suddenly died.\footnote{This sentence was added by JAM (4/29/12) for clarity: \textbf{yàʔ-qhâ} \textbf{hâ} \textbf{chi} \textbf{tê} \textbf{khɛ} \textbf{lɛ̀} \textbf{mɔ̂} \textbf{jâ} \textbf{ò} \textbf{lɛ}, \textbf{khâʔ} \textbf{tû} \textbf{ve} \textbf{ɔ̀-khɔ̂} \textbf{nà} \textbf{lɛ}, \textbf{tê} \textbf{pɔ̂ʔ} \textbf{tí} \textbf{šɨ} \textbf{e} \textbf{šē} \textbf{ve} \textbf{yò}.}

18. When he had fired off his crossbow, the husband said, ``Aha, it's dead!''

19. But when they took a look, [they saw that the bolt] did not strike its body,
but was stuck someplace below it.

20. ``Look, woman!\footnote{The rest of this text has also been added for clarity.}

21. You see what a great marksman I am!\footnote{Lit. ``how skillful I am at shooting animals!''}

22. Even though I didn't hit him with my shot, the animal died anyway, didn't you
see!'', he said.

