
120 The Lazy Men and the Princess

(Told by Headman Cà-bí)

\. Well, my brethren, once upon a time, as our ancestors---er, let us discuss and
recount to each otehr the story of the four lazy men, shall we?

\. Listen up well!

\. Once upon a time, there was a lazy fellow who went around \{trading/buying and
selling things\}.

\. Since he was very lazy, he went along until he came to a place where four roads
met, and sat himself down there.

\. He sat down there by the roadside waiting for people to come, and after a long
time another lazy guy appeared, right there, at the crossroads.

\. When the two of them met, they asked questions of each other, and kept on waiting
over there.

\. While they were still waiting, another person appeared!

\. Now \{there were/it reached\} three people there waiting.

\. They still kept on waiting, all of them there.

1\. Since they were so lazy they just waited by the minute.

1\. Then---another guy appeared. There were now four of them.

1\. So they all talked to each other, and discussed things with each other.

1\. ``So, what can you do, sir?\footnote{\textit{mɔ́ ò }[vocative]: ``O teacher.''}'' [somebody] said.

1\. ``Well, I can \{tell fortunes/predict the future\},'' he said.

1\. ``And you, what can \textit{you} do?''

1\. ``I know how to make crossbows,'' he said.

1\. ``I can make crossbows.''

1\. ``What can \textit{you} do?''

1\. ``I know how to swim,'' he said.

2\. Well, while they were all \{discussing these things/talking this way\}, one
of them asked [the fortune-teller], ``What is going to happen?''

2\. ``\{Pretty soon/In a short while\}, around nightfall, the daughter of the ruler
of that city there, the \{daughter of the king/princess\}, will pass above us having
been \{bitten and lifted up by/lifted up in the beak of\} \{a hawk/an eagle\}.

2\. ``You make a crossbow carefully!'' he said.

2\. They were staying near a lake, these guys.

2\. The crossroads they were at was near a big lake.

2\. Then, after he had finished making the crossbow, that princess who had been
\{bitten by/carried in the beak of\} the great eagle passed above them.

2\. He cocked his crossbow and shot it off, and [the eagle] fell down with a big
splash, into the lake.

2\. When it had fallen in, [the \{swimmer/guy who could swim\}] swam out and got
her, that child---she was a beautiful maiden, they say.

2\. So these three guys argued with each other: ``I got her!'' ``No, I got her!''
they said.

2\. After they had \{gotten/retrieved\} the girl, they \{debated/quarreled/argued\}
with each other as if in a court of law, those three guys.

3\. What should be done?---when they went to the great place, the great and mighty
lords' place, they were all arguing and quarreling with each other.

3\. Which person would win?

3\. ``Me, me, I win because I predicted the future!'' he said.

3\. Thus they fought with \footnote{\textit{g̈ɔ̀ dàʔ ve}: lit. ``pulled against each other.''} each other.

3\. ``I made the crossbow, and got him [the eagle] with a shot,'' he said.

34-A. ``Yeah, I made the crossbow and shot it.''

3\. ``I got her by swimming,'' he said.

3\. The three of them argued with each other.

3\. Arguing and arguing, they arrived at the palace and said, ``Which person won?''

3\. ``She was recovered by swimming, so the swimmer, the last person, is the winner,''
so they said.

