\setcounter{footnote}{0}

1. Well, today our friends, our Caucasian\footnote{\textbf{Kâlâ-phu}: the original meaning of Kâlâ was ``India'', but this was later extended to any foreign country inhabited by non-Mongoloid people. Adding the morpheme \textbf{phu} 'white' specifically refers to Caucasians. See DL:325-6.} brethren, have visited us, us hillfolk,


us Lahu, and while we are celebrating the New Rice Festival\footnote{\textbf{Cà-šɨ́-ɔ̄} \textbf{câ} \textbf{ve}: 'celebrate the New Rice Festival' (lit. "eat cooked rice from the new paddy"). A harvest festival celebrated in October, both in animist and Christian villages. See DL:446.} they have come

to participate with us in joy and gladness.

2. And we, all of us, must welcome them happily.

3. We will all have fun.

4. It's a good thing, to do it this way.

5. The merit is very great.

6. And the Westerner has graciously provided food and drink for us, having bought


a pig for all of us, to feed us all, feeding us and giving us to drink.

7. The merit is very great, is it not!

8. Since the grace of God has reached all of us, we must praise Him, for God's


grace is great and never-ending.

9. And today we are celebrating together, three Westerners, and three Karens,


together happily with us Lahu, preparing food and drink together, and since

we are praising the grace of God for helping us, all of us can celebrate, and


the merit is very great.

10. And besides this, since great and important people have become friends of all


of us, if it were not for God's helping us, we would never have been the sort


of people who would have been able to meet them.

11. The fact that people living in different countries, in different lands,


because of the grace of God, are celebrating together today, is indeed


because God's grace is very great, is it not!

12. In the future we look forward to meeting each other again, and we will


always be properly grateful for God's grace.

