\setcounter{footnote}{0}

1. Once upon a time there was a widow.

2. One day when that widow had gone to cut thatch-grass, she got terribly thirsty
and looked around

for water, but couldn't find any.

3. Then she scooped up and drunk the water that had been left in an elephant's
footprint in an old

abandoned field, and returned home.

4. Then not long afterwards she became pregnant, and gave birth to a child.

5. That child had no body, but only a head---just a head.\footnote{\textbf{ó-qō} is the ``short form'' (\textbf{khɔ̂-ŋɛ}) of `head', while \textbf{ó-qō-šī} (with the addition of the morpheme for round objects) is the ``long form'' (\textbf{khɔ̂-yɨ̀}). See GL 74-76.}

6. But the mother took care of it properly, and after a fair amount of time had
passed, one day since it

was time to cultivate her field, she was going off to clear the undergrowth.\footnote{\textbf{hɛ} \textbf{phɔ̂} \textbf{ve} (OV) `clear away undergrowth from a swidden with a heavy buush-knife'.}

7a. The head-child said, ``Mother, don't you do the clearing. Let me do it,'' he
said to his mother, and

7b. The mother burst out,\footnote{``Burst out'' translates the emphatic particle \textbf{qha} after qɔ̂ʔ \textbf{pî} `say to'.} ``I doubt that you could do it!'' she said, and

7c. The head-child replied, ``I can do it. Just leave me here and go back.''

7d. ``Pretty soon when it's getting dark come and carry me back,'' he said, and

7e. The mother for her part did as he said, and left him there.

8. When it got dark and the mother came back to look around, she saw that all the
clearing of

the undergrowth had already been done for her.

9a. In the same way,\footnote{``In the same way'' translates \textbf{thɔ̂} (Punf) `also'.} when the time came for felling the large trees\footnote{\textbf{hɛ} \textbf{thu} \textbf{ve} (OV): ``slash a swidden'', i.e. clear a field by felling the big trees in an area before burning it off. Here translated ``clear-cut''.} in
the swidden, he said to

his mother ``I know how to clear-cut the field,'' so when he said that, his mother
went and carried him

to the field and left him there, and she [later] found that the field had been
completely clear-cut for her.

10. And then one day that country was about to be invaded\footnote{\textbf{bɔ̂ʔ} \textbf{yù} \textbf{ve}: lit. ``shoot-take'', i.e. `be taken over by fighting'.} by others, and everybody
predicted\footnote{\textbf{šɔ̄} \textbf{dàʔ} \textbf{ve}: ``calculated mutually''}

that those others would win.

11a. Then he [the Head] said, ``Even if you just let me fight all by myself, I'll
win.''

11b. And when he said that, the King said, ``If you fight and win, I have six daughters.''

11c. ``Whichever one you want to marry, you may marry, and also I can give you
about one half of

my kingdom,'' said the King.

12. Thereupon the Head-child went to fight, and conquered everything, and he [the
King] gave him

his daughter and the land.

13. Then one day, after the Head-child had gone\footnote{His means of locomotion is not specified.} to market, the old lady\footnote{\textbf{ɔ̀-pi-ma}: `grandma, mother-in-law, old lady'. This refers to the Head-child's mother, who was evidently living with him and his princess.}
was tidying up the house,

and accidentally set fire to the place where the Head used to be put, and when
the Head-child got

back from the market he no longer had anywhere to hide, so he turned into a very
handsome man.

