\setcounter{footnote}{0}

1. Once upon a time there was a widowed mother and her child.

2. One day they went to pick fruit to eat --- pine-nuts.

3. Well, when she said, ``These fruits now, if there were someone to pick them for
me, whichever of my seven daughters he wanted I would give her to him,'' a great
snake came and shook those fruits down for her.

4. Then when the widow went back home, that big snake came following after her.

5. ``If you are indeed and in truth my son-in-law come right into the house,'' she
said, and the snake came and she had him choose [a wife] from among some of the
seven daughters she had --- but they wouldn't have him.

6. The last-born one accepted him, saying ``mother, however you have me marry, that
is how I shall marry!'', and when she had wed him, as dawn broke, there was a shining
lantern and a shining screen! [Ed: behind the shining screen was a handsome young
man --- the snake had been metamorphosed]

7. Then one day, the elder sister was jealous, and again went off to pick pine-nuts
to eat.

8. Taking [her younger sister] along, they went off to pick them, and having cut
the tree part-way through [on a previous occasion], she sent her younger sister
climbing up it, then took up her younger sister's son and said:

9. ``Sister, your turban --- er, your son wants to have your turban, I think'';
so she took it off and dropped it down to them.

10. Then, when she had dropped it down to them, she [the older sister] pushed the
tree over.

11. Then, after her younger sister had died, the elder sister picked up all her
clothes and put them on [herself] and lived thereafter with her younger sister's
husband.

12. The younger sister's breast remained hanging from a \{rattan bush/clump of rattan\}.

13. Then one day three doves came and called out, saying,

14. ``Sheh-khao!\footnote{This was the name of the younger sister's son.} O Sheh-khao! Don't you remember the place where you would eat?''
[the husband is impressed by the dove's accusation of the elder sister]

15. ``Sheh-khao's father\footnote{The Lahu are teknonymous. It is the eldest sister who is talking to her husband, the former snake.}, you must be crazy! I'm the one who's your who is your
wife!

16. Just look at my tunic and my skirt!'', she said...

