
{\LARGE{}\textbf{133 The Snake and the Widow's Daughters}}

{\LARGE{}\textbf{<Retelling by Paul Tcalo>}}

{\LARGE{}1. I am retelling in November, 1965, this story which was told by a girl
who lives in the Lahu village of Pa-yeh-shu in Thailand.}

{\LARGE{}2. The story is about a widow and her eldest and youngest daughters.}

{\LARGE{}3. Once upon a time there was a certain widow who had seven very beautiful
daughters. }

{\LARGE{}4. One day this widow got a great craving for pine-nuts, so she went into
the forest and found a pine-tree. }

{\LARGE{}5. Well, she couldn't manage to climb up the pine-tree. }

{\LARGE{}6. So she couldn't pick them to eat. }

{\LARGE{}7. After a while she said, ``Ah, if there were only someone to
pick those pine-nuts for me, I would give him one of these seven daughters of mine
to marry!`` }

{\LARGE{}8. Just as she said this, a big snake came out and picked them for her.}

{\LARGE{}9. Then, when she turned back for home, the big snake also came following
after.}

{\LARGE{}10. They got near the widow's house, but the big snake would not enter.
}

{\LARGE{}11. So she said, ``If indeed and in truth you are my son-in-law,
enter my home!`` }

{\LARGE{}12. And the big snake came into the house. }

{\LARGE{}13. When the snake had come into the house, she had him choose from among
her seven daughters. }

{\LARGE{}14.The girl that the big snake wanted to have was the first-born, the
eldest sister. }

{\LARGE{}15. The eldest sister, however, did not want the snake. }

{\LARGE{}16. But finally the very youngest of all the seven daughters did accept
the snake, in obedience to her mother. }

{\LARGE{}17. Well, after they were married, towards the morning of the next day,
the big snake had a lantern brightly lit in the room where his wife, the youngest
daughter slept. }

{\LARGE{}18. Then, when the other six girls and their mother came sneaking up to
have a look, they found there behind the partition a very handsome man. }

{\LARGE{}19. The big snake was no more to be seen. }

{\LARGE{}20. After this had happened, that eldest daughter who hadn't accepted
him the night before --who had said she wouldn't have him--she was most distressed
at heart. }

{\LARGE{}21. For she was very jealous that her youngest sister had fared thus,
and had won such a fine young man. }

{\LARGE{}22. So she decided she would try to cause her younger sister's death and
then marry her husband. }

{\LARGE{}23. Well, after the younger sister had married the snake, there came a
day when she gave birth to a child. }

{\LARGE{}24. So one day the eldest sister got an idea and put it into action like
this: }

{\LARGE{}25. She chopped through a certain pine-tree in the forest just far enough
so that it wouldn't fall over by itself but would remain attached by a thread--and
left it there. }

{\LARGE{}26. Then one day she took her youngest sister along and went off to pick
some pine-nuts to eat. }

{\LARGE{}27. When they reached that pine-tree, when they got to the pine-tree she
had cut half-way through, she had her younger sister climb up to pick the nuts.
}

{\LARGE{}28. And then, when her sister had climbed up and reached the top of the
tree, she pushed the tree over. }

{\LARGE{}29. The younger sister died when the tree was pushed over. }

{\LARGE{}30. After her death the elder sister removed the tunic, the headdress,
and all the other clothes from her younger sister and put them on herself. }

{\LARGE{}31. Then she took her sister's child and led him back home. }

{\LARGE{}32. When they had returned home, she went in to live with her younger
sister's husband.}

{\LARGE{}33. When the younger sister died, she turned into a dove. }

{\LARGE{}34. Then, on a certain day, this dove came to sit on a tree-top near their
house--near her husband's house--and called out to her son. }

{\LARGE{}35. Her little son's name was Sheh-khao.}

{\LARGE{}36. The words she called were these: ``Sheh-khao! O Sheh-khao!
}

{\LARGE{}37. ``Don't you remember your mother anymore? }

{\LARGE{}38. ``Look, I am your mother! }

{\LARGE{}39. ``Don't you remember the place where you would eat?``[1]
}

{\LARGE{}40. Then she called out to her husband as well: ``Serpo, Serpo![2]
}

{\LARGE{}41. ``It is I who am your wife, it is I!``}

{\LARGE{}42. At this Serpo came out and said, ``Hey, what is going on here
with this dove! }

{\LARGE{}43. ``It can talk like a human being! }

{\LARGE{}44. ``Wife, oh Wife, hurry and come look! }

{\LARGE{}45. ``This dove is calling out and saying 'I am your wife!' }

{\LARGE{}46. ``It's amazing!`` }

{\LARGE{}47. Thus he called to his ``wife`` (it wasn't his real
wife, you know, it was the elder sister), and she came to look on. }

{\LARGE{}48. Then the dove conversed with her husband, saying, ``I am your
wife. }

{\LARGE{}49. ``I am your true wife.`` }

{\LARGE{}50. She told him the whole story of how she had died on that day. }

{\LARGE{}51. She told him many things: }

{\LARGE{}52. ``This person who is now called your wife is not your true
wife. }

{\LARGE{}53. ``She is my eldest sister.``}

{\LARGE{}54. If we are to explain how this could have happened to Serpo, it is
because all of the widow's seven daughters had the same face. }

{\LARGE{}55. Their bodies and their figures were identical. }

{\LARGE{}56. So you couldn't tell one from the other.}

{\LARGE{}57. Then Serpo asked the dove, ``How did she lie to you that day
to make you go into the woods with her?``}

{\LARGE{}58.  ``Well, my sister said to me, 'Let's go pay a visit to our
relatives in yonder village for a few days!}

{\LARGE{}59.  ``But that's not what happened at all. }

{\LARGE{}60. ``On the way, at the side of the road we saw a pine-tree and
she made me climb it to pick the nut-bearing cones, and she pushed me over. }

{\LARGE{}61. ``The tree fell. She had chopped into it. So I died. }

{\LARGE{}62. ``Now, having been killed that way, I have become a dove,``
she said.}

{\LARGE{}63.  ``I am your true wife,}

{\LARGE{}64. ``But I am no longer a human being. I have become a dove.
I have become an animal.}

{\LARGE{}65. `` I have become a bird. I am no more a human being,``
she said. }

{\LARGE{}66. ``So I am very sad, you see, because of you,`` she
said. }

{\LARGE{}67. ``Today I wanted to come and visit you. }

{\LARGE{}68. ``And I wanted to see my son Sheh-khao too, so I came to visit,``
she added. }

{\LARGE{}69.  When she had finished saying this, the elder sister said, ``Look
you, this is impossible--that filthy, miserable bird! }

{\LARGE{}70.  ``Don't listen to what it says! }

{\LARGE{}71. ``Anything a bird like this comes and says is just stuff and
nonsense! [3]}

{\LARGE{}72. ``It's a great bird of lies! }

{\LARGE{}73. ``}{\LARGE{}\textit{I}}{\LARGE{} am your real wife.}

{\LARGE{}74. ``Just look at my face! }

{\LARGE{}75.  ``And look at my tunic and skirt! }

{\LARGE{}76.  ``Am I not your wife? }

{\LARGE{}77.  ``This is my child. }

{\LARGE{}78.  ``Don't let your thoughts be led astray by a wretched bird
talking to you like this!`` she said.}

{\LARGE{}79.  But in Serpo's heart something had changed. }

{\LARGE{}80.  In his heart he did not really want to believe the elder sister who
called herself his wife. }

{\LARGE{}81. Day after day he thought about it. He pondered all sorts of things.
}

{\LARGE{}82. Then at last he got an idea. }

{\LARGE{}83. He carefully brought seven very sharp knives, and placed them on top
of a board, arranging them so that they were spaced out each at a little distance
from one another. }

{\LARGE{}84. When he had finished arranging them this way, he summoned his wife.
}

{\LARGE{}85. ``Wife, come here a moment,`` he said. }

{\LARGE{}86. ``If you are indeed my wife, you just climb up onto these
seven knives and try to walk. }

{\LARGE{}87. ``Try walking up there over them all, one after the other.
}

{\LARGE{}88. ``When you're walking there, if the knives don't cut into
your feet, I shall believe that it is you who are really my wife,`` he
said. }

{\LARGE{}89.  ``The reason for this is that I have sworn an oath before
God.}

{\LARGE{}90.  ``'If she is really my wife, do not let these seven knives
cut her!' }

{\LARGE{}91.  ``But if you are not my wife, they will cut you. This is
the oath I have sworn.`` }

{\LARGE{}92.  Thereupon his wife--that is, the elder sister who had been playing
the part of his wife--carefully lifted up [her feet in a pair of] excellent shoes,
great shoes that nothing could pierce, and climbed up [onto the board]. }

{\LARGE{}93.  But, when she stepped up onto the first of the knives, it cut into
her foot, and she suddenly fell over forward. }

{\LARGE{}94.  Falling forward, she sprawled flat onto the seven knives, and the
seven of them, one after the other, cut her body up into seven pieces. }

{\LARGE{}95.  And there she died. }

{\LARGE{}96.  Then Serpo said, ``Aha! Now I know that she was not really
my wife. }

{\LARGE{}97. ``That dove was really my wife. }

{\LARGE{}98.  ``The widow's eldest daughter was no good. }

{\LARGE{}99.  ``She was not a good daughter.}

{\LARGE{}100. `` She acted out of envy. Now I know all.`` }

{\LARGE{}101.  That's the end of the story.}

{\LARGE{}102. If we ask what we ought to learn from this story, it is this: }

{\LARGE{}103. If we are envious of others, if we try to bring about someone's ruin
and downfall, it does not profit us, and it does not profit him.}

{\LARGE{}104. Thus, thinking it will profit us, if we try to ruin someone else,
that person may indeed be brought to destruction. }

{\LARGE{}105. But we also must thereby be destroyed, you see.}

{\LARGE{}106. It does us no good, and it does him no good. That's the way it is.}

{\LARGE{}107. So, let no man envy anyone else, and let no one bring his fellow-man
to ruin. }

{\LARGE{}108. This story is a bit like the story of the potter and the laundryman.[4]
}

{\LARGE{}109. Do not be envious of others, it means.}

{\LARGE{}110. If we envy other people there is no advantage for us. This is what
we have said. }

{\LARGE{}111. The story is finished. }

