\setcounter{footnote}{0}

1. What we shall read in the Bible, is in John 16:27-28:\footnote{King James translation: The division between the sentences in Lahu is not the same as the verses in the King James Bible, naturally enough.}

\textit{2.} \textit{``For the Father himself loveth you, because ye have loved
me, and have believed that I came out from God.}

\textit{3. I came forth from the Father, and am come into the world:} 

\textit{4. again, I leave the world, and go to the Father,'' he said.}

\textit{5.} Elsewhere in John 14:6: 

6. \textit{``Jesus saith unto him, I am the way, the truth, and the life:} 

\textit{7. no man cometh unto the Father, but by me.''}

8. Let us praise the Lord.

9. Oh, our lord Jesus, so full of love, day after day in thy great love thou givest
us all life.

10. While we are alive, since thou givest all of us the strength and force to earn
the money that we need---we seek and we find.

11. Today, now at this very hour when we praise Thee, out of all the wealth Thou
hast given us, each person will take a little and we'll give it to Thee.

12. The money that we give is not much, it is very little.

13. But since we have heard the Gospel\footnote{\textbf{ha-lɛ̀} \textbf{ve} \textbf{bo-khɔ̂}: lit. ``the happy blessed words.''} [like] so many people in the past, and
since [we] have also come to believe in you, putting our trust\footnote{\textbf{cû}: `put one's trust in'. See JAM 1989.} in Jesus' name,
[we] offer it up to you the Father.

***

14. We all, everybody in the world, hope that God will love us.

15. If indeed we are people who have those hopes\footnote{Lit. ``who hope this way.''}, in whom should we put our
trust?

16. Jesus said, ``I have come from the Father.

17. Since you believe in me, that Father loves you.''

18. Oh my brethren, since God wants to love all of us peoples of the world, we
all of us ought to put our trust in this path that God has prepared for us.

19. And what is that path that He has prepared for us?

20. God's precious Son\footnote{\textbf{yâ-phu}: this poetic word, perhaps meaning literally ``white child/pure child", has a wide range of meanings, including `human being; person; man; son.' It is also used as a humble first person pronoun for a man in love poetry.} called Jesus, that person arrived to us in the world
among us human beings.

21. He definitely arrived here.\footnote{\textbf{gà} \textbf{la} \textbf{pə̀}, lit. ``arrive'' (\textbf{gà}), ``hither'' (\textbf{la}), ``do completely/finish up'' (\textbf{pə̀}).}

22. For this reason, if we humans are all people who hope for heaven\footnote{\textbf{g̈ɨ̀-ša} \textbf{mû-mì}: `heaven' (lit. ``God's country'').}, we must
truly believe in him.

23. How are we able to know what is the road that goes to Heaven? If we humans
here on earth ask where the King lives, now in Thailand the place where the King
lives is Bangkok.

24. If we ask where is the road that we must travel to go to Bangkok, the road
that has been prepared for us by the King is from the railroad station in Chiang
Mai, from Nawpati\footnote{\textbf{nɔ̀-pa-tîʔ}: an area of Chiang Mai near the train station.}, through Lampang\footnote{A city about 30 km south of Chiang Mai.}, and going on to Bangkok.

25. When we go along that road, we \{get to see/ find\} the city where the King
of us all lives, we get to see his house, we come to know him himself.

26. In the same way we humans on earth can all be people who seriously ponder God's
heavenly home, and who wait expectantly for it.

27. For that reason, since we can all be people who want to enter the land that
He has prepared for us, we must go along the road he has given us.

28. As for that road, what kind of road is it?

29. Jesus has said, ``I am the Way,

30. the truth, and the life.

31. No man cometh unto the Father, but by me.''

32. Therefore for all of us human beings, trusting in Jesus, believing in Jesus,
is very good indeed.

33. O my brethren, today if we all go along to the very end of these paths which
God has prepared for us, we will get to see our King.

34. Thus God sent Jesus to this world for us, in order to atone for all of our
sins, and if we say how Jesus taught us this: ``I have come into the world in order
to save you people of the earth from your sins.''

35. So then, in order for God to truly forgive all of your sins, so that you may
follow on his path, he said, ``I have died upon the Cross for the sake of you all.''

36. Therefore the Father will forgive all the punishable sins of everyone who trusts
and believes in me, and when you are in His presence God will judge you to be virtuous
and righteous.

37. So then, [among] all of us who are now happily going along this path, not even
one of us must think of hastily turning around and going on any other paths.

38. I have now preached God's holy words to all of you, my brethren.

39. Among all of us the living Holy Spirit\footnote{\textbf{tèʔ} \textbf{ve} \textbf{co-ha} (Left Relative Clause) or \textbf{co-ha} \textbf{tèʔ} \textbf{ve} (Right Relative Clause).} did not dwell.

40. Therefore, in order to have the Holy Spirit appear among us human beings, God
poured\footnote{\textbf{qha-bî-bî} \textbf{pî} \textbf{tā} `gave in plenitude'.} [its essence] into his Jesus.

41. Therefore He said in another place that by giving [eternal] life to all the
people of the world, Jesus was \textit{manna} \footnote{\textbf{khɔ̀-mô} \textasciitilde{ \textbf{khɔ̀-mû}} `cake, bread' < Shan \textit{khaw-mun}. The Christian sense of `manna' was inadvertently omitted from DL 384.} for human beings.

42. So whoever partakes of this manna will never be hungry again.

43. So all of us human beings ought to partake of this manna until we are utterly replete.

