\setcounter{footnote}{0}

1. Once upon a time, when God created the [first] human being, He created him to
stay inside a gourd\footnote{The Freudian analysis of this creation myth is obvious: gourd = womb. The gourd has become the logo of the Lahu people in China and Burma, just as buffalo horns have become the logo of the Wa.}, it is said.

2. One day a bat lit off a \textit{tawkaw}-fruit\footnote{The \textbf{tɔ́-kɔ} iš a kind of palm with very large leaves, bearing plentiful fruit.} so that it fell right down,
and since there was a barking-deer dozing under the \textit{tawkaw}-tree, it fell
crashing down on him, and because he was startled he suddenly jumped up and ran
away into an old field.\footnote{\textbf{hɛ-šā}: a swidden allowed to lie fallow and revert to jungle over several years}

3. A jungle-chicken also flew over, and alighted\} on a cassia shrub\footnote{\textbf{mì-cɛ̀}: cassia sp. (a shrub with red flowers) [\textit{Cinnamomum cassia}].}

4. At that time a horse was also grazing under the cassia.

5. That horse got very scared, so he ran away, and this reached the ears of an
elephant.

6. Since the elephant was also very frightened, when he was running away full tilt
he happened to step on the gourd and made it roll\footnote{\textbf{g̈a} \textbf{nâʔ-pu}: happen to (\textbf{g̈a}) + step (\textbf{nâʔ}) + roll (\textbf{pu})} [away down a slope].

7. One day when God came to visit the human, he couldn't find him, so he summoned
all the animals and all the trees, and asked them [where the human was].

8. Then when God asked the pine tree, the pine tree said ``I didn't see anything.''

9. So God said, ``When the human gets out [of the gourd], may he chop you down!

10. And do not ever again put out sprouts\footnote{\textbf{ɔ̀-cā} \textbf{cā} \textbf{ve}: lit. ``sprout sprouts,'' a cognate object construction that does not translate well as such.}!'' Because God said that, to this
very day the pine tree puts out no sprouts.

11. Then when he went on to question the butterfly, but since the butterfly also
said ``I didn't see anything,'' God said, ``And as for you, every day, rain or
shine, you will have to go look for food.'' And since He said that, every day they
have to look for their food.

12. Then he went on to ask the \textit{peh-meu}\footnote{\textbf{pɛ̂-mə̂(n)-cèʔ}: a tiny bee-like insect, the size of a fly, that secretes a very sticky liquid that makes beeswax. Its bite does not hurt, and its larvae are edible. See DL:848.} bug, and the \textit{peh-meu}
bug said, ``I saw it go down into the water down there." So since he
told Him this God said, ``May your kind always have liquid!'' And because God
blessed him thus, to this very day his species produces a liquid.

13. Then God called all the animals together again, and made them [try to] take
the gourd back out of the water.\footnote{This sentence contains a four-verb concatenation: \textbf{qɔ̀ʔ} \textbf{yù} \textbf{tɔ̂ʔ} \textbf{cɨ}.}

14. All the animals, all the different kinds of animals then tried to take it out,
but they could not.

15. A kind of crab then grabbed it, and placed it on the bank for Him.\footnote{This very sentence, used as an example in DL 854, there receives a different translation, also possible in a different context: ``He took a kind of crab and put it on the bank for him.''}

16. Then God went on to say: ``Whoever\footnote{\textbf{a-šu} \textbf{tê} \textbf{g̈â}: The classifier for humans (\textbf{g̈â}) is used, since all the animals are personified in this tale.} manages to take this human out of the
gourd will get to eat the first fruits of their successful labor in the fields!''
When he said this, the starling said, ``I'll get him out, if that's the way it
is.'' And when God said, ``If you can do it, do it!,'' the starling said, ``My
beak is very long!'', and all day and all night he pecked away at it.

17. But even though that beak of his got all worn down, it couldn't pierce through.

18. Then the rat said in his turn, ``\textit{I'll} now get him out!'', and he gnawed
away and pierced it through.

19. When he pierced through, unfortunately\footnote{\textbf{šē}: one use of this particle is to express regret at the verbal event. See DL 1200-1201.} he slightly bit the human's nose
as he gnawed.\footnote{\textbf{gə̀ʔ-hə̂ʔ}: `strike by gnawing'.}

20. Thus to this day, we have an indentation\footnote{\textbf{chèʔ-kú-lɛ̀ʔ}: 'a dent; an indentation.'} between our eyes.\footnote{The before-and-after appearance of human beings was illustrated for me by a sketch!}

21. And to this very day, rats live secretly with humans, and have permission to
keep sneakily eating people's food and drink, people's rice.

