
{\LARGE{}\textbf{\#156  Simple Conversation}}

{\LARGE{}1 Cà-vàʔ: Have you gone swimming in the river yet?}

{\LARGE{}2 Cà-g̈âʔ: I haven't swum yet.}

{\LARGE{}3 Paul: Speak a little louder!}

{\LARGE{}4 Cà-vàʔ: Where are you going today?}

{\LARGE{}5 Cà-g̈âʔ: I don't have to go anyplace.}

{\LARGE{}6 Paul: I'm certainly not going anywhere.}

{\LARGE{}7 Cà-vàʔ: For my part, I want to go to the opium field! [1]}

{\LARGE{}8 Paul: Who are you going with? Hurry up and say it, Cà-vàʔ.}

{\LARGE{}9 Cà-vàʔ: I'm going with you!}

{\LARGE{}10 Paul: Before this I never even got to go with you.}

{\LARGE{}11 Cà-vàʔ: We're off to cultivate the opium field.}

{\LARGE{}12 Cà-g̈âʔ: Where is it anyway?}

{\LARGE{}13 Cà-vàʔ: It's at Na-li.}

{\LARGE{}14 Cà-g̈âʔ: Haven't you finished cultivating it yet?}

{\LARGE{}15 Cà-vàʔ: There's a little we still haven't finished.}

{\LARGE{}16 Cà-g̈âʔ: How much still is left?}

{\LARGE{}17 Cà-vàʔ: There's just a little left still! By tomorrow we'll have
cultivated enough.}

{\LARGE{}18 Cà-g̈âʔ: So tomorrow you'll have it all done!}

{\LARGE{}19 Cà-vàʔ: It'll be done. Tomorrow.}

{\LARGE{}20 Cà-g̈âʔ: I'll be able to help you tomorrow too. Together with some
of you.}

{\LARGE{}21 Cà-vàʔ: Do help!}

{\LARGE{}22 Cà-g̈âʔ: All this staying at home for one or two months, I got
tired of working in the swidden.}

{\LARGE{}23 Cà-vàʔ: So come and help! Come help me cultivate tomorrow, just
with me.}

{\LARGE{}24 Paul: Hey Crocodile, [2] tomorrow are you going to help along with
me then?}

{\LARGE{}25 Cà-g̈âʔ: I will help you a little bit. I've been such a lazybones
this year.}

{\LARGE{}26 Paul: So if you all are going to help I'll help, me too.}

