
1 Paul: Now what I'd like to know is, how do we Lahu people cultivate our swiddens,
from the day that we start to cultivate until we harvest, how does it go?

2 Tell me one stage at a time okay. That's what I would like to know.

3 In the beginning the field --- you look for a place to clear a swidden, and when
you find a place to clear a swidden, how do you clear it of trees, how many days
do you have to spend clearing it, that sort of thing.

4 Cà-bo: The way we Lahu cultivate our swiddens for a living is like this.

5 The first thing is we must clear away the undergrowth from the swidden.

6 After we finish clearing away the undergrowth, we clear away the big trees.

7 As for the time it takes to clear away the big trees, we have to do it about
two months.

8 And after we've cleared away the big trees, we do the first burning of the swidden.

9 Once this first burning is over, we do a second burning.

10 As for the time it takes to do the second burning, it doesn't take very long.

11 It only takes one or two weeks.

12 Then as for the amount of time it takes to go on to plant the fields, it takes
about a month and a half for us to get to plant the rice.

13 After the planting is done, we have to do the weeding.

14 And now for the amount of time it takes for weeding, it's this way.

15 Some years when we cultivate the fields the weeds are very high, so we can't
do the weeding right away.

16 Sometimes it takes about a month to do the weeding.

17 Sometimes it takes about two months to do the weeding.

18 If they're very deep we must keep weeding right up until harvest time.

19 But as for the time it takes to harvest, it doesn't take too long.

20 For those who have large swiddens to cultivate it takes about one month; for
those who don't have a large swidden to cultivate, it just takes two or three weeks
to do the harvesting, to have to harvest.

21 Paul: Well then, what you were saying about clearing away the undergrowth just
now, how do you actually do it?

22 And what you called doing a second burning, how do you do that?

23 I still don't understand. This clearing away undergrowth and doing a second
burning. How do you do that burning, and how do you clear away the undergrowth?

24 Cà-bo: So what we call clearing away the undergrowth is like this.

25 We must clear away the undergrowth in the hot season.

26 When we get rid of these overgrown weeds, the weeds decrease greatly --- when
they become fewer the swidden is not overgrown --- so we Lahu say.

27 For this reason we have to clear the undergrowth as soon as possible.

28 And doing the second burning of the field is like this.

29 After we've done the first burning, we gather together and pile up all the stuff
which didn't get burned and set fire to it --- this is called doing the second
burning.

30 Paul: So what you said was that clearing away the undergrowth still didn't clear
away the big trees, right?

31 Before you've cleared away the big trees, you finish pulling out and getting
rid of the weeds little by little, that's what you do.

32 Clearing away the undergrowth goes that way, you said, right?

33 Cà-bo: Yep. We can say that's the way it goes.

34 You can't just clear away the big trees right away.

35 You have to clear away and kill all the weeds under the trees in the field,
all of them.

36 And it's not just a question of clearing away and killing those weeds.

37 It becomes fertilizer --- as it rots it becomes fertilizer and it's very good
for the rice plants.

38 Paul: So then when you harvest the rice, how do you harvest it, do you help
each other out?

39 Or else does everybody do the harvesting by himself?

40 Cà-bo: When we Lahu harvest the rice, it's not exactly like the way the plains
folk help each other.

41 Sometimes somebody is not able to harvest his own rice since he's unwell or
miserably poor, and when we realize this the whole village goes and manages to
help him out.

42 If he's not sick and ailing since everybody has his own swidden to cultivate
and sometimes he fields are not even in one area, so he doesn't have time to go
off to this person's place and that person's place [to help them out].

43 It's just in times of real need that we go and help them out.

44 Paul: Now I'd just like to know about harvesting the rice.

45 Harvesting the rice, how does the rice have to get to be so you know when the
time has come to harvest it?

46 How does the rice have to look so that you know that it's ripe?

47 How many days does it take to do the harvesting, would you say?

48 When harvesting is done, how do you stack up [the rice]?

49 Tell me how you go about threshing it.

50 Cà-bo: Harvesting the rice is like this.

51 When we look it over this is what we have to know.

52 After the rice has sprouted when we see that its grains have hardened, we know
it's time.

53 When the grains have hardened and become yellow we know it's time to do the
harvesting, so we must harvest it.

54 After the harvesting, we must stack up the rice.

55 This is why stacking up the rice is important:

56 If we don't stack up the rice plants the rice is no good. Besides, it can't
be threshed loose [so that the grains fall off the stalks easily].

57 Paul: When you say ``can't be threshed loose'' what do you mean?

58 Cà-bo: Not come loose --- the meaning of ``not come loose'' is that when you
thresh it the grains don't fall completely off.

59 Paul: Doesn't come falling off.

60 Cà-bo: Yes, some of it is still hanging on. That's what we call it.

61 Paul: How many days do you have to leave it stacked up before you do the threshing?

62 Cà-bo: Properly speaking our custom is that you can do the threshing after
it's been stacked up two or three nights.

63 But if you don't have a lot of time and you have a great deal stacked up, some
years it has to be stacked up for a whole month.

64 Some years it happens that you have to leave it for two or three weeks.

65 When you've left it like this after you've finished stacking it up, even if
it rains, even if it's very hot, there's nothing to worry about. It doesn't get
wet.

66 Paul: So then when you've threshed the rice, when it's all threshed, do you
carry it to store at home? How do you carry it to store?

67 Cà-bo: Threshing the rice goes this way.

68 If the swidden is far away you do it right there in the field.

69 You make a house in the field, you make a huge storage basket, and you do the
threshing right there in the field.

70 But if the field isn't far away, if it's right outside the village, you can
bring it right back home directly.

71 Paul: So then after the rice grains have been carried and put away it's all
over right?

72 So how do you do this? When the threshing is done these --- er, what do you
call it? --- the stalks, how do you call them in Lahu?

73 How do you deal with them?

74 Cà-bo: When the threshing is done we Lahu don't prepare [the stalks] [for other
uses]\footnote{This sentence may be more clearly paraphrased as: yù lɛ yɛ̂ ve tê cə̀} the way the people in the towns do.

75 We just take all those rice stalks that have been threshed away and stack them
up into piles.

76 And if we have to recultivate an old field we carry all those rice stalks and
scatter them all over the field, and set fire to them, and when we plant rice the
next time it will be good for the crop.

77 Paul: Well, so then I have come to understand a lot about how we Lahu clear
our swiddens for cultivation, how we cultivate the rice in the field, haven't I?

78 So then, at the same time you are cultivating the rice, what other things are
you also growing besides?

79 Planting all kinds of things, planting cornfields, whatever fields you plant,
things like that.

80 Cà-bo: It's not that we only cultivate [rice in] our swiddens.

81 We Lahu at the same time we are cultivating [rice in] our swiddens we also plant
all kinds of legumes, various kinds of bananas, kinds of taro, all kinds of things
which are good to eat, all kinds of edible things.

82 Besides, we are also planting cornfields so we have to cultivate the cornfields
in a separate plot of land.

83 If we didn't have cornfields, since we Lahu raise pigs to eat, if we didn't
have anything to feed the pigs with, we'd have no way of earning money.

84 Paul: Well, that's great then.

85 Now the thing to worry about is that the Thai authorities don't want to let
you clear swiddens, this is the most worrisome thing, isn't it?

86 If the Lahu aren't allowed to clear swiddens, they wouldn't have anything to
eat, would they?

87 Even if you don't have any money, it doesn't matter.

88 If you just have rice you can live, wouldn't you say?

89 This matter of not being able to clear [new] swiddens and having to rely on
old ones by putting fertilizer on them, this is very hard to do.

90 So then going to find a new place, as you said already, every single place has
an owner, but even though it's said they have owners, you never see anybody actually
cultivating [those places].

91 We even see places abandoned for ten years, or even twenty years.

92 So then I don't know what can be done in this situation, right?

93 If you all got together, some people who knew their language, if you could go
down there to the government office, to the officials, and have a discussion about
this, it would probably be good.

94 Cà-bo: Yes. It's not just that we have to be familiar with other people's languages
and customs or that we all have to be able to speak their languages.

95 However what we have planned time after time we have not succeeded in, what
we had planned to get by asking for it, to the limit of our ability.

96 We won't know whether we succeed or fail until we try.

97 And as for the authorities, what they tell us is, ``Go find a place in the plains!''.

98 But we can't find such a place, no matter how we look.

99 Because of this the government is not taking pity on us and is not helping us
out, so that for us Lahu who can neither clear swiddens nor have paddy fields,
even though it is difficult now, in the future it will get even more and more difficult.

100 There is probably no way that we can come up in the world.

101 Paul: They tell you to plant coffee, fruit orchards, all kinds of things, so
couldn't you earn a living that way?

102 How could you do that, things like that?

103 Cà-bo: It's this way. If we Lahu succeeded in doing what they tell us, we'd
be able to earn a living.

104 But even if we wanted to plant a fruit orchard, we can't buy the fruit trees.

105 We don't have the money to buy them. We don't have the money to do it.

106 It takes a very long time. It can't be done quickly.

107 For this reason, during the time when we still haven't gotten anything, getting
food to eat will be a very difficult thing for us Lahu.

108 The second thing is the matter of buying the fruit trees, if we don't have
the money we can't buy them.

109 Since you just have to spend money for everything, being able to earn a living
is very hard for us Lahu.

