\setcounter{footnote}{0}

1. Singer: Would you like to get some songs?

2. P: Sure!

3. Singer: This time it's not a song for the girls to sing. It's for the guys. Please
give a listen!

160a.

4. Oh maiden who dwells above the morning star\footnote{This song is appropriate when a girl from the north comes visiting to the south. \textbf{tê} \textbf{mû} \textbf{pə̄-tɔ̂ʔ} \textbf{ɔ̀-na}: lit. ``one-sky-morning.star-above'', i.e. to the north of where the morning star appears, i.e. the northeast.} --- ah!

5. If perchance you have\footnote{An informant translated this particle \textbf{ha} into ordinary Lahu as the comitative particle \textbf{gɛ} `with/for'.} no lover --- ah!

6. Please let me speak a few sweet words to you --- ah!

Singer: That's how he goes at it.

160b.

7. Singer: And let's do it again. So this time let's listen to what the girl says.

8. Oh faithful lover who has dwelled beneath the evening star\footnote{\textbf{tê} \textbf{mû} \textbf{pə̄-qɛ̀} \textbf{ma} \textbf{hɔ́}: lit. ``one-sky-evening.star-below'', i.e. to the south of where the evening star sets, i.e. the southwest.} --- ah!

9. If perchance you have no mistress --- ah!

10. Lift up your body's feet, lift up your body's hands,

11. And come sojourn in the land to the north of the east --- ah!

12. Singer: That's what they say. These are the nice words of the woman. ``Come
to her,'' she's saying. They live in separate places, you see.

160c.

13. Singer: And now the song ``Sometime Soon'':

14. Before too long a time has passed

15. We shall be, shall we not, like the red-headed a-bɛ̂ branch!\footnote{The \textbf{a-bɛ̂} plant is used to strengthen cotton thread after it is spun. When young, the fruits are on twigs growing close together. As the plant matures, the twigs open up, like loves separating.}

16. Each branch going off to dwell in a different place!

17. Singer: This is a North-country dialect.

18. At the time they're singing about, the two of them\footnote{The speaker says ``the two of us'', placing himself in the perspective of the woman in the song.} are both living in the
north, but in the future each of them is going to a different place.

19. They're separating. They won't get to see each other again.

20. This is what the young woman sings. They'll each be living in a separate land.

21. So then, what sort of thing would you still like to learn now? The things you'd
still like to learn?

22. You just have to ask.

23. P: As for what the man's side is ---

24. Singer: I just sang the man's side already. You just heard the man's side a
while ago.

160d.

25. Cà-bî (the newly appointed headman): A Yellow Lahu song. I'll try doing it
now: \direct{laughs}

26. In a land to the north of the east you dwell

27. Maid, if perchance you have no \{suitor/lover/man\}

28. Please try singing to me with words of love.

29. Oh, you Lahu [maiden]!

160e.

30. Cà-bî: In a land to the north of the east you have dwelled!

31. Maid, if perchance you have no suitor

32. Please try singing to me with words of love!

33. Were you not fair this Lahu would not court you.

34. Cà-bî: This is a Yellow Lahu song. \direct{laughs}

35. P: We've heard that one already.\footnote{Indeed, the first three lines are almost identical to 160d.}

36. Cà-bî: You've heard it?

37. T: Well, you've got a very good voice!

38. Cà-bî: I'll try it again. Just a little bit. All the ones I've done.\footnote{I.e., I won't mess up the words.}

39. T: Do it well, one song after the other. Do it as prettily as you can!

40. Now I'll do another one in Lahu --- In Black Lahu.

160f. unusable

160g.

1. Today --- er, that's not right yet! ---

2. Tonight in the best of all nights

3. And today in the best of all days!

4. Let's talk everything over and see!

5. Say whatever you want to say!

6. In the north of the east\footnote{Cf. \textbf{pə̄-tɔ̂ʔ} `morning star'.} a maiden lives

7. Even if I were to talk to you until the morning star turns around backwards

8. I could never say\footnote{\textbf{jɛ̀} \textbf{mâ} \textbf{mi}, lit. `discuss-not-catch'.} to you all I want to say; I could never tell you all I want
to tell.\footnote{Lit. `tell-not-catch'}

9. If I tell you when the moon is shining

10. If the dawn breaks when the moon is shining

11. We shall be, shall we not, like the red-headed a-bɛ̂ branch

12. Each of us will go our separate ways.

13. Is it not so, O Lahu [maid]!

14. [the audience]: Wow!

160h. -j unusable

