\setcounter{footnote}{0}

1. Once upon a time, a blind man and a cripple, leading each other along, went into
the woods, and when the two of them had gotten to a place to spend the night on
their journey, before they settled down for the night they caught two turtles.

2. When they had caught the two turtles, the cripple made the blind man pick him
up and carry him.

3. At this point a tiger came by, and when he came sneaking up to have a look, he
said, ``Ah, there are two people over there, I'll eat them.

4. I'll bite into both of them and eat them all up!''

5. And meanwhile [the two of them] went along together, taking the turtles with
them, and [the tiger] went to have a look and said, ``Oh, this one has heads on
both sides!\footnote{The tiger had noticed the cripple's head in close proximity to that of the blind man.} I can't eat it!''

6. So then the turtles asked the tiger to kindly pick them up and set them down
on top of a tree-stump, so they could also just have a look [at the two-headed
animal].

7. But when [the tiger] picked [the turtles] up in his mouth and set them down,
they fell off, he kept picking them up and setting them down, but they kept falling
off.

8. Finally [the tiger] violently picked them up in his mouth and threw them away.

9. He roughly grabbed them with his mouth and threw him away.

10. Once the turtles had been grabbed and thrown away they went inside the carrying-basket
of the blind man and the cripple.

11. And so as they went on, continuing on their way, going on and on, and ah! they
reached a place to sleep. The two of them got there and lay down, and they slept
the whole night through, they slept and slept, slept deeply, and they woke up in
the morning when the sun was already high, and they got up and went on and on,
entering the dwelling place of the Phi-tan-ngweh.\footnote{\textbf{phí-tan-ŋwe}: a one-eyed monster who lives in the jungle. See DL:889.}

12. And when they got there [the monster] said, ``I'm going to eat you two all up,
so let me take a look at your lice then!"

13. Saying that they [the turtles] were their lice, they took the turtles and showed
them to him.

14. ``Oh, these lice are bigger than I am --- I can't eat them!'', he said.

15. Then (he said), ``Let me see your comb!'', and they took a rake and showed it
to him.

16. ``I probably can't eat them!'' he said there.

17. So then, this time coming back again\footnote{Note the concatenation \textbf{qɔ̀ʔ} \textbf{qɔ̀ʔ} `coming back again', where the 1st element is a pre-head versatile verb, while the 2nd element is the main verb. One could also say \textbf{qɔ̀ʔ} \textbf{qòʔ}, with no difference in meaning.} from the [monster's] place where the
two of them had gotten to, when they reached a certain point on the road, there
was in the distance a bird's nest nestling in a tree-top.\footnote{\textbf{ŋâʔ-phɨ} \textbf{tê} \textbf{phɨ} \textbf{phɨ} \textbf{ā}: This is a cognate object construction, where the 1st occurrence of \textbf{phɨ} is part of a compound noun, while the 2nd is a classifier, and the 3rd is a verb. See GL:68.}

18. So then the hunchback\footnote{Here for the first time called \textbf{cɔ̄-qɔ̀ʔ-pā} ``crooked back one'' instead of \textbf{khɨ-qɔ̀ʔ-pā} ``twisted leg one'', translated `cripple'.} made the blind man climb up [to get it] while he watched.

19. When he climbed up there, [the cripple] said ``Did you reach it yet?'', and
he said ``Not yet''; ``Did you reach it yet?'', ``Not yet.''

20. Then when he had reached it, [the cripple] said ``Grab it!'', and when he grabbed
it, inside the bird's nest there was a snake!

21. When he looked, when this blind man tried to look at it, he saw that ``It's
a snake!''\footnote{His vision had miraculously been restored because of his fright.}, and roughly flung it away, sending it down [from the tree], so that
the cripple jumped up and ran away, and that [crooked] place in his leg had straightened
out.

22. [The cripple's] leg now straightened, and [the blind man's] eyes now seeing,
the two of them went along together again, and things were now very fine, and they
returned home.

