
{\LARGE{}\textbf{The Orphan and the Magic Frog  \#163}}

{\LARGE{}1 Once upon a time, there was a certain orphan. Day after day he would
go to chop firewood. One day while chopping up firewood he got hungry and ate a
meal, when a frog appeared.}

{\LARGE{}2 When the frog came out, he fed her [the frog] day after day.}

{\LARGE{}3 So every day when he would return home, a meal would be cooked for him.}

{\LARGE{}4 Since food had been cooked for him, even though he went and asked other
people, ``Who has cooked food for me in my house?'', they said ``Nobody has been
cooking in your house, that ragweed hut [1], it's too filthy!'' So one day he tried
hiding himself, and --- wow! --- she [the frog] came jumping out from inside a
gourd.}

{\LARGE{}5 When she jumped out, he quickly grabbed at her and caught her.}

{\LARGE{}6 ``Will you be my older sister?'', he said.}

{\LARGE{}7 But she said, ``No!''}

{\LARGE{}8 ``Will you be my younger sister?'', he said.}

{\LARGE{}9 But she said, ``No!''}

{\LARGE{}10 ``Will you be my mother?'', he said.}

{\LARGE{}11 But she said, ``No!''}

{\LARGE{}12 ``Will you be my wife?'', he said.}

{\LARGE{}13 And she said, ``Yes, yes!''}

{\LARGE{}14 So then for a long time afterwards they stayed married.}

{\LARGE{}15 Then she said, ``Build a pigsty, build a cattle shed.''}

{\LARGE{}16 After this, when he had built a cattle shed and a pigsty,}

{\LARGE{}17 there were cattle there, there were pigs there, there were all kinds
of animals, there were horses ---}

{\LARGE{}18 After this happened, later on she got all infected, and he didn't want
to stay married to her, he wanted rather to marry the headman's wife's daughter.}

{\LARGE{}19 So then since he wanted to divorce that wife of his he drove her out
day after day, saying, ``Hey, go and dig a pond at the edge of the village.''}

{\LARGE{}20 When he said this on a certain day when she was going to to dig the
pond, she just left and didn't come home.}

{\LARGE{}21 She left, and entered the pond, and when she got up to her waist she
said, ``You don't really want to separate from me, do you?''}

{\LARGE{}22 But he said, ``Indeed I do!''}

{\LARGE{}23 She went in further and this time when she went in she reached as far
as her neck, and said, ``You don't really want to separate from me, do you?''}

{\LARGE{}24 But he said, ``Indeed I do!''}

{\LARGE{}25 ``But pretty soon I'm going to die,'' she said, and kept on going in
further.}

{\LARGE{}26 After this happened he went home, but nothing was there anymore.}

{\LARGE{}27 All he saw was a ragweed hut.}

{\LARGE{}28 So then when he went to ask for the headman's wife's daughter in marriage,}

{\LARGE{}29 she said, ``Phooey! I don't want to marry someone like you! A miserable
wretch like you! You don't have anything!''}

{\LARGE{}30 Day after day he went there where she had gone into the pond, where
his wife had entered the pond, and wept there.}

{\LARGE{}31 So then on a certain day a frog appeared.}

{\LARGE{}32 When the frog appeared, it said, ``Well, why are you crying, orphan?''}

{\LARGE{}33 ``Because my wife went in here, I am weeping,'' he said.}

{\LARGE{}34 [The frog said,] ``You go and crush some chili peppers in a mortar.}

{\LARGE{}35 If you crush them, I will suck it empty for you,'' so he immediately
crushed them.}

{\LARGE{}36 Having crushed them [the frog] ate [2] them all up, it sucked and sucked
[the mortar] empty, so that its anus became all red, and that guy [the orphan]
laughed.}

{\LARGE{}37 [The frog] nearly split apart.}

{\LARGE{}38 ``Now you only have two chances left.}

{\LARGE{}39 Do not laugh!''; when it said this it sucked it again, but [the orphan]
laughed again.}

{\LARGE{}40 When he had laughed again, all of a sudden water came spurting out
[of the frog's mouth] and went splash!}

{\LARGE{}41 ``Now you only have one chance left.}

{\LARGE{}42 If you laugh again, you will not succeed.}

{\LARGE{}43 You will not see your wife again.''}

{\LARGE{}44 When it [the frog] said this [the orphan] managed to restrain himself
properly, so when [the mortar] was sucked empty he got to see his wife again.}

{\LARGE{}45 That's the end.}

{\LARGE{}\textbf{Notes}}

{\LARGE{}[1] }{\textbf{Á-qhâ-yɛ̀}}{\LARGE{} 'ragweed hut', taken as
the archetype of a miserable place to live. DL:83.}

{\LARGE{}[2] The verb  }{\textbf{lɛ̀ʔ}}{\LARGE{}  , lit. ``lick``,
is also used for snacking for pleasure (e.g., fruit, sweets, spicy things).}


