\setcounter{footnote}{0}

1. Once upon a time there was a man who had two sons.

2. Then later the man died, and the two sons were left.

3. So then the older and younger brothers divided up the property.

4. Dividing up the property, all the older brother gave his younger brother were
just two dogs and a curved

bush-knife.\footnote{\textbf{á-thɔ-qwɛ̀ʔ}: a heavy curved machete used to cut undergrowth before burning a swidden.}

5. As for the older brother, he got two oxen.

6. All the rest of the property he kept for himself too, the story says.\footnote{The phrase \textbf{qôʔ-ve} is used here as a sort of quotative topicalizer, rendered as `it is said (that)'. Its effect is similar to that of the final unrestricted particle \textbf{cê} `quotative'.}

7. Well, one day when the younger brother was off clearing his swidden with his
curved bush-knife, he scooped

up a white mushroom and munched it\footnote{\textbf{khâʔ-khòʔ}: \textbf{khâʔ} `scoop up,' \textbf{khòʔ} `chew noisily'. Evidently the younger brother cut off the hallucino- genic mushroom by accident, then decided to taste it.} and fell asleep.

8. Having scooped up the white mushroom and munched it and fallen asleep, one day
\footnote{The younger brother had clearly been knocked out for several days.} a great troop of

monkeys showed up and said, ``Ah, this person has died! His heart was very pure.
Let us pick him up and carry

him to where we'll perform a service for him.\footnote{\textbf{bo} \textbf{te} \textbf{ve} (OV) `perform a religious ceremony'. An expression used by both animist and Christian Lahu. See DL:942.} Let us sing hymns for him,''
they said, and they picked him

up and laid him down somewhere way up on a rock cliff.

9. Having carried him up there, they arrived at a cave, where they sang hymns and
ministered to him, that orphan.

10. Then they summoned a certain bird and had it sing hymns.

11. That bird having been summoned and made to sing hymns, he [the younger brother]
suddenly stood up

and shouted, so all those monkeys ran away.

12a. Then, when he looked around inside the cave, he found silver and gold, and
he took it all and went back

home, where his elder brother said, ``Younger brother, how did you get all this?''


12b. ``Well, you once gave me a bush-knife. I cut down a mushroom with that knife
and munched it while I

was clearing my field and I fell asleep, and a big troop of monkeys came, and they
said, `Oh, this fellow is

dead! Blowfly eggs are swarming all over him! But his heart is pure', and they
picked me up and carried me,

and we got to a cave, and they sang hymns for me and took care of me, and I shouted
at them and they ran

away, and when I looked around the cave it was full of silver and gold. So I took
it for you and brought it

back,'' he said.\footnote{This stretch of speech (\textbf{12d-e}) is a fine example of the use of the suspensive Punf \textbf{lɛ}, which occurs multiple times, as the younger brother compresses his whole adventure into a single sentence.}

13. ``Oh younger brother, your older brother will go too! Older brother will go
do it too. Let me have your

bush-knife,'' he said, and the older brother set off right away.

14. The elder brother set off, and while clearing his field he scooped up a white
mushroom and munched it

and fell asleep, and that great troop of monkeys came and said, ``Oh, this one's
heart is very evil!" But even

so, although his heart is evil, we'll do something for him, we'll help him,''
so they picked him up and carried

him up onto a rock cliff.

15. As they were carrying him up, when they reached the halfway point, climbing
and climbing up, he opened

his eyes and said, ``Do it right, don't let go of me and let me fall!'' So those
monkeys got frightened and let go

of him so that he fell, and falling down there he landed right inside a tiger's
den.

16. After he landed inside the tiger's den, two or three days went by without his
coming back, so the younger

brother was very worried, and went to look for him, and carried him back.\footnote{We are not told what had happened to him in the tiger's den.}

17. After this, one day the younger brother went off with his two dogs.

18. Taking the two dogs along, on the way to plow his paddy-field,\footnote{Instead of clearing a swidden (\textbf{hɛ} \textbf{phɔ̂} \textbf{ve}), the younger brother is now preparing to plow a wet-rice field (\textbf{ti-mi-tháy} \textbf{ve}). Perhaps the riches acquired in the previous episode helped the brothers come up in the world.} a group
of traders came by.

19. ``These dogs of yours can't plow a paddy-field, we bet,\footnote{``We bet'' translates \textbf{nɛ̀-ɔ̄} (Pnf) `suppositional'.}'' they said. ``They
can \textit{too} plow!'' he said, but they

said they didn't believe it, those merchants.

20. So then the orphan summoned the two dogs, and let fly with a riceball and told
them to go get\footnote{\textbf{cɔ} (V) `follow a trail (as a hunting dog)', \textbf{e} (Pv) `transitive motion'.} it.

21. Then he threw a riceball [in the opposite direction] and told them to go get
it also.

22. ``Look, you guys said my dogs couldn't plow a field, but they can \textit{too}
plow!'' he said, so that orphan won a

horse-drawn cart.

23. Having won the horse-drawn cart, he went back home, and told his elder brother
about it.

24. ``How did you get this, younger brother?'' asked the older brother.

25. ``Those two dogs that you gave me, when a bunch of traders came along they
said, `We bet that these dogs

of yours can't plow a paddy-field', and I said, `Sure they can!' and when I taught
them how to plow the field, they

gave me a horse-cart.''

26. When he had said this, the older brother said he would go too.

27. The older brother set off, but he didn't take any rice with him.

28. He was tossing a white stick when the big group of traders came along.

29. When they came they said, ``If your dogs are able to plow this paddy-field,
we'll give you this horse-cart.''

30. So he threw the white stick to the dogs with all his might, trying to send
them off, but they wouldn't go.

31. He threw it again, and urged them on, but they still wouldn't go.

32. So he just beat them to death.

33. After he had beaten them to death he told his younger brother. When he asked,
``Where did you go with my

dogs?'' he [the older brother] said, ``My dogs, er, your dogs I whacked to death.''

34. ``What happened was, when I sent them to plow the field they wouldn't
plow it for me.

35. ``When I urged them on they wouldn't do it for me, so I just beat them
to death and buried them down there,''

he said, and the younger brother wept.

36. He wept for days and days, and when he went to sob bitterly down where his
dogs had died, a flower had

sprung up.

37. When he shook that flower that had sprung up, silver and gold came falling
down under it, such as no one

had ever seen.\footnote{\textbf{chɔ} \textbf{qhe} \textbf{mâ} \textbf{hêʔ} \textbf{o}: lit. ``not like a person.''}

38. So he picked it up and carried it back, and when he went to talk to his older
brother at home, he said, ``How

did you get this, younger brother?''

39. So he said, ``At that place down there where you beat the dogs to death, on
their grave a flower was

growing, when I looked I saw a flower. When I shook it, silver came falling down,
so I picked it up and carried

it back here,'' he said.

40. So the older brother said, ``I'm going to shake it too,'' and the younger brother
said ``Go for it!''\footnote{\textbf{qay-ʔqay-ʔ}: lit. ``go, go!''} and

when older brother went to shake it, he also saw a flower up above him.

41. When he shook it, what came falling down was caterpillars biting him, and it
went badly for him all over

[his body].

42. So he got angry and slashed down the flower, right on that spot.

43. When it had been slashed down, Younger Brother wept again, day after day, he
kept weeping and weeping, until

a bamboo stalk sprouted.

44. So he chopped down the bamboo stalk and made a bird's nest out of it at home.

45. When he had built the nest, one bird came and laid an egg, then two birds came
and laid their eggs, so

every morning he had eggs to eat.

46. So Older Brother said, ``Younger Brother, how did you get these to eat?''

47. ``At the place where you beat the dogs to death a bamboo stalk came up, and
I made a bird's nest out of it,

and ever so many birds came and laid eggs in it for me. All I'm eating is bird's
eggs,'' he said, and Older

Brother said, ``Oh, Younger Brother, lend it [the nest] to me!'' so he did lend
it to him, and Older Brother took

it away and set it on the edge of the veranda of his house, and one bird came and
shat in it, then two birds

came and shat in it.

48. So Older Brother set fire to it, and Younger Brother wept.

49. Younger Brother wept, he sobbed and sobbed, then he ate up a single sugar-pea
\footnote{\textbf{thù-khá-šī}: called \textit{thùa lantaw} in Thai.} that he had, and his belly

got bloated with gas, so he went down to the town.

50. When he got to the town he farted, but the sound of his fart was so gentle
that the people there said, ``Oh, what

did you eat this morning that you fart like this?\footnote{The expression \textbf{qhɛ̂} \textbf{tɛ̀ʔ} usually occurs with nothing intervening between the elements, but sometimes (as here) they are separated by other material.} Come here, come here!''

51. So he went with them, and when he said, ``This morning I ate a sugar-pea, but
I didn't get to eat any rice, so

my stomach got bloated and I came here,'' so they gave him a bolt of cloth, fed
him well, and sent him back.

52. So Older Brother asked him, ``How did you get this, Younger Brother?'',
and he said, ``Down there in the

place where you burned up my bird's nest I scratched around and searched, and I
ate a sugar-pea, and I went

off and arrived at the town, and I farted, and they said, `Since you fart so beautifully,
come over here! Who

are you?' so I went along, and they fed me, and gave me a bolt of cloth, and I
came back,'' so Older Brother

boiled up a big bowl of sugar-peas to eat, and went off.

53. He boiled them up and ate them and went off [to the town], and he farted mightily,
he did! It was loud, that

older Brother's!

54. Since it was loud, since the sound of his farting was so very loud, they said
``Come here, who are you?'' so

he went [with them] and said, ``It's me, it's me, masters! Since I haven't had
any food to eat, I've come to you.''

55. When he said this, they fed him rice and sewed up his asshole.

56. After they had finished sewing up his asshole, he went home, and staying at
the outskirts of the village,\footnote{Where the villagers would normally go to relieve themselves.}

he called to his wife.

57. ``My wife, my wife, come here!'' he said. ``Why?'' his wife said.

58. ``Fetch a scissors, fetch a scissors!'' he said, and the wife took a scissors
and got back to where he was,

and snipped his asshole right open, and the shit came bursting out on her, and
one of her eyes became

cloudy.\footnote{This ending seems both abrupt and morally unsatisfactory. Why did the older brother's wife have to suffer? And didn't the younger brother ever get a definitive reward for his naïve virtue? Evidently this was part of a longer story where these loose threads were tied up.}

