\setcounter{footnote}{0}

1. Oh ho! today the sun and the moon form a pair, but these two, Mr. Black and Miss
White\footnote{\textbf{Cà-nâʔ} (Mr. Black) and \textbf{Na-phu} (Miss White) are common Lahu personal names. They generally refer to the color of the person's complexion. Mr. Black would be a man of dark skin while Miss White would have a rather light brown complexion.} are not one pair, and so we elders and young people take these two and
make them one pair.

2. The sun and the moon are one pair, these two are not one pair, so here in this
place this evening we of this community, all of us, take Mr. Black and Miss White
and make them one pair.

3. Oh, let them enjoy good and carefree health, and when they have children may
their birth be easy,\footnote{The expression \textbf{yâ ye ve} can mean either `bear children easily' or `bear strong children who do not die'.} may they be as plentiful as the fruit of the táw-ma tree
\footnote{This tree has yet to be identified, although it is proverbial for bearing abundant fruit.}, as plentiful as the fruit of the wild gooseberry; open this blessing on Mr.
Black and Miss White, favor them with these boons!

4. The small bamboo spoon and the large bamboo ladle\footnote{The small bamboo spoon is used for serving relishes; the large ladle is for stirring and serving rice.} are one pair, but these
two are not yet one pair; the sun and the moon are one pair; but these two are
not one pair, so we of this community, we elders and young people all together
take Mr. Black and Miss White and make them one pair.

5. Oh, when they have children, may the children not die; let there be many animals
underneath the house and many people inside the house\footnote{To possess many animals (tethered under the house since Lahu houses are raised on stilts) and many children are signs of \textbf{ɔ̀-bon} `merit; blessing' in a Lahu village community.}; let this blessing be
opened upon them, let them be favored by these boons!

6. Oh ho! Good fortune to Mr. Black and Miss White!

