\setcounter{footnote}{0}

1. Ah! Oh! Oh! This year\footnote{\textbf{chi} \textbf{qhɔ̀ʔ} 'this year' is Red Lahu. In standard Black Lahu these two syllables have been fused into the compound \textbf{cì-qhɔ̀ʔ}.}, our whole group, our whole community have come together.
You who dwell there in the land of the spirits\footnote{Lahu ritual prayers frequently make use of Tai vocabulary in order to enhance the poetic effect. Here the Tai word \textbf{phí }for `spirit' rather than the Lahu \textbf{nê} is used.}, come and eat this new rice,
come and drink this new water, come to eat and drink this year!

2. Let all of your community [of spirits] also come, oh, whoever are elders [in
the land of the spirits], every one of them, everybody, come to this house, come
to eat, come to drink.

3. There is here an abundance of food in this household, so please let there be
no worries, no sadness, this blessing and this boon please once again ordain for
us!

4. Just like the Enduring Pillar of White Rock\footnote{The \textbf{kɔ-mô} or \textbf{kɔ-mô-twɛ̄} are roughly carved wooden posts placed in the ground in front of the village temple. (See Plates \#17, 58, and 59 in Matisoff 1988.) Walker's Red Lahu informants had various opinions on what these posts signify. A common view was that they are earthly replicas of G̈ɨ̀-ša's divine \textbf{kɔ-mô-twɛ̄} in heaven, which is said to be of white stone representing indestructible and undecaying permanence. As Walker has demonstrated, the animist Lahu believe that everything on earth has a divine prototype.} up there in the eternal heavenly
abode (of G'uisha), once again ordain the blessings of eternal life upon the womenfolk
and the menfolk.

5. Oh, if we work for one day in the fields, may the food not be exhausted in ten
days! If we work for one year in the fields, may the food not be exhausted in
two years! With these boons once again bless us; please ordain them for us, upon
each and every person, upon each and every household.

6. If we work for one day in the fields, may the food not be exhausted in ten days!
If we work for one year in the fields, may the food not be exhausted in ten years,
in one hundred years! For this boon we once again wait\footnote{Lit. ``we wait to be established within this boon''.}, please ordain this
blessing upon us.

7. Oh, you with the all-hearing ear, with the all-seeing eye, you who rule over
the eternal heavenly dominions\footnote{This is \textbf{G̈ɨ̀-ša}, the supreme creator divinity of the Lahu.}, you who think eternal efficacious thoughts
\footnote{\textbf{dɔ̂-tìʔ-gâ-tìʔ} 'thoughts of immortal spirits who only need to think of something for it to come to pass'.}, you who have reached the path to the heavenly land of the blessed, please
shower each and every person within this household with your blessings, a blessing
for each one, please ordain that for us!

8. Upon every last person open up the blessing and ordain the boon that when they
work for one day in the fields there will be enough to eat for ten years.

9. Like the eternal heavenly pillar up there, bless us once again with a life like
that of the stone, like the stone pillar that forever resists decay and blight,
so that the husbands and wives and all their descendants may not have to worry
or be distressed, oh, please open up this blessing and ordain it for us, bless
us with this boon and ordain it for us!

