\setcounter{footnote}{0}

1. Haw, O, O! This night is indeed an auspicious one!

The other day this man's soul went away in a direction where it has suffered much
trouble.\footnote{The specialist who recorded this prayer said the soul of his patient had wandered away into the jungle where it had no place to eat or sleep.} Tonight come back along these silver steps, come back along these
golden steps.\footnote{A poetic reference to symbolic steps carved in wood.}

2. Come back again into the four corners of the household, come back to the guest's
side of the fireplace, come back to the household head's side of the fireplace.
\footnote{Guests are seated in the area below the fireplace ( \textbf{qha-cî-tɛ̂-kɨ} ), while the household head sits in the area above the fireplace ( \textbf{qha-cî-tɛ̂-na} ).}

3. Oh do not dwell between the feet of Pū-phā up there, do not dwell between
Pū-phā's hands.\footnote{The gatekeeper \textbf{Pū-phā} lives in a large house at the gate of the land of the dead. At the birth of each human, \textbf{Pū-phā} knows when he or she will die. If a sick person's soul has wandered towards the land of the dead before its time, \textbf{Pū-phā} takes his great staff and beats the erring soul resolutely, sending it fleeing back to its rightful place in the land of the living.}

4. Do not obey the voices of death, do not obey the voices of sickness; do not prepare
yourself for the pathway of death, do not prepare yourself for the pathway of sickness.

5. Listen to the voice of this chicken\footnote{This refers on one level to the chicken held by the specialist during the rite, but it also seems to be viewed as the earthly counterpart of the two divine chickens of G'uisha. This seems to be implied by the poetic word \textbf{pɨ̀-ca}, where the 2nd syllable means `pair'. See note 9.} within the four corners of this house
and come back! Cover your legs up to the knees with this silver jewelry, cover
your hands up to your wrists with this golden jewelry! Within the four corners
of this house there are all kinds of food, every kind of drink.\footnote{Lit., ``there are firmly attached all kinds of food and drink''.}

6. Oh soul, do not jump into the mouth of a yellow bird of prey, a green bird of
prey! Do not jump into the mouth of a yellow tree, a green tree.\footnote{This means ``do not get lost in the jungle''.}

7. Do not think about the pathway of death; they say that over there Pū-phā's
hills are precipitously steep.

8. Pū-phā, you who rule over me, command this soul to be put back between the
feet of this man, order that it be put back between his hands.

9. Up there, G'uisha's Eternal Golden Fowl, Kāy-khâ-tiʔ\footnote{The specialist who recorded this text explained that \textbf{G̈ɨ̀-ša}, the supreme Lahu supernatural, owns two birds: one called ``Kāy-khâ-tìʔ'' (cf. Thai key ``chicken''; kham `gold'), while the other is called ``\textbf{ŋɛ̂ʔ-phu-tìʔ}'' (ŋɛ̂ʔ means simply ``bird'' in Red Lahu; \textbf{phu} `white'). He said that when a person's soul wanders from his body, these two birds call it to return to him. They also cry for the long life of people in general.} , you have cried
one time, do not cease to cry, cry that this man may enjoy a long life! You two
white servants of G'uisha up there\footnote{The ``two white servants'' are G̈ɨ̀-ša's two birds.}, cry that this man may enjoy a long life,
cry that he may have a life like rock and stone, cry that he may come back here
again.

10. Some time ago, oh soul, you went away and suffered much trouble. Oh, tonight
you are wandering between the streams and the rocks, you are wandering between
the hills and the streams. Tonight come back!

11. Oh, if you are ashamed, cover your face with these leaves and come back, come
back, follow underneath this white cotton string, this red cotton string\footnote{In Text II of this article (p.24, sentence 9), the soul is also bidden to return by following a red string: \textbf{khɛ-ni} \textbf{qhâ-šɨ} \textbf{qɔ̀ʔ} \textbf{la} \textbf{ôʔ}.}
and come back, listen to the voice of this chicken within the four corners of the
house and come back!

12. Has the soul come back?

13. It has come back!\footnote{The officiating specialist ends the prayer by calling out this question to the people inside the patient's house. The people inside the house reply loudly in the affirmative.}

