
202 Wandering Soul

1 Haw, O, O! The day before yesterday this man's soul went away to that side and
suffers much trouble;\footnote{The specialist who recorded this prayer said the soul of his patient had wandered away into the jungle where it had no place to eat and sleep, and consequently it suffered.} tonight, this good night, tonight come back along these
silver steps, come back along these golden steps.\footnote{A poetic reference to the symbolic steps carved in wood.}

2 Come back again into the four corners of the household, come back to the guest's
side of the fireplace, come back to the household head's side of the fireplace.\footnote{Lahu Nyi prayer language includes many couplets, used primarily for sound effect. The result may be simple metaphor or wild hyperbole. In this instance, instead of ``come back to the fireside'' the specialist calls the soul to ``come back to the guest's side of the fireplace, come back to the household head's side of the fireplace''.}

3 Oh do not jump between the feet of Pū-phā up there, do not jump between Pū-phā's
hands.\footnote{The gatekeeper of the land of the dead (Pū-phā) lives, according to the Lahu Nyi myth, in a large house at the gate of the land of the dead. At the birth of each human, Pū-phā knows when he or she will die. If a sick person's soul has wandered towards the land of the dead before it is time for that person to die, Pū-phā takes his great staff and beats the erring should resolutely, sending git fleeing back to its rightful place in the land of the living. See also fn 8 below.}

4 Do not obey the voices of death, do not obey the voices of sickness; do not move
on to the pathway of death, do not move on to the pathway of sickness.

5 Listen to the voice of this chicken\footnote{Reference to the chicken held by the specialist during the rite.} within the four corners of this house
and come back; cover your legs up to your knees with this silver jewelry, cover
your arms up to your elbows with this golden jewelry;\footnote{In Lahu ``silver'' and ``gold'' are part of a couplet.} within the four corners
of this house there is every kind of food, every kind of drink.

6 Oh soul, do not jump into the mouth of a yellow bird of prey, a green bird of
prey; do not jump into the mouth of a yellow tree, a green tree.\footnote{This means ``do not get lost in the jungle''.}

7 Do not think about the pathway of death; up there Pū-phā's hills are precipitously
steep.

8 Pū-phā, you who rule over me, order this soul to be put back between the
feet of this man, order that it be put back between his hands.\footnote{``Feet'' and ``hands'' are part of a couplet, not specific references. The meaning is ``order this soul to be returned to the body of this man''.}

9 Up there, G̈ɨ̀-ša's chicken, Kāy-qhâ-tiʔ,\footnote{The specialist who recorded this text explained that G̈ɨ̀-ša, the supreme Lahu supernatural, owns two birds: one called ``KÁy-qhâ-tìʔ'', apparently a chicken (kÁy is ``chicken'' in Thai), and the other called ``Ngɛ̂ʔ-phu-tìʔ'' (ngɛ̂ʔ means simply ``bird'' in Lahu). He said that when a person's soul wanders from his body, these two birds call it to return to the body. They also cry for the long life of the people.} you cry one time, do not
cease to cry, cry that this man may enjoy a long life; you two white servants of
G̈ɨ̀-ša up there,\footnote{The ``two white servants'' are G̈ɨ̀-ša's two birds.} cry that this man may enjoy a long life, a life like
rock and stone; cry that he may come back here again.

10 Some time ago, oh soul, you went away and suffer much trouble; oh, tonight you
are wandering between the streams and the rocks, you are wandering between the
hills and the streams; tonight come back!

11 Oh, if you are ashamed, cover your face with these leaves and come back, come
back, follow underneath this red cotton string and come back, listen to the voice
of this chicken within the four corners of the house and come back!

12 Has the soul come back?

13 It has come back!\footnote{The officiating specialist ends the prayer by calling out this question to the people inside the patient's house. The people inside the house reply loudly in the affirmative.}

