\setcounter{footnote}{0}

1. This evening these two young people are divorcing,\footnote{\textbf{bà-dàʔ-chêʔ-dàʔ}: lit. "mutually cast-aside mutually break".} they say.\footnote{``They say'' translates the quotative particle \textbf{cê} .}

2. Great Ruler, Powerful Ruler here at this place\footnote{Addressed to the guardian spirit of the locality. \textbf{chò} \textbf{kàʔ} \textbf{ve} \textbf{kà} 'here' is poetic language for \textbf{chò} \textbf{kàʔ} \textbf{ve} \textbf{tê} \textbf{kà}, where the classifier \textbf{kà} appears after the numeral \textbf{tê} 'one'.}, I bring for you these beautiful
beeswax candles made by my own hands. By the lighting of these (two) candles\footnote{Walker (1976:179) had never personally witnessed this rite, but was informed that two candles were involved.},
we say that the marriage is dissolved. We pray that neither party suffer misfortune.

3. We people have not broken up this couple, they themselves have broken up.

4. Today we of this community offer you these beautiful beeswax candles made by
our own hands. The village fine of two rupees has been paid.

5. If it is the man who wishes the divorce, the man must pay the fine; if it is
the woman who wishes the divorce, the woman must pay the fine. Once the fine has
been paid, the matter has been resolved.

6. May both sides enjoy good health!

7. Let the whole village witness that these two people are breaking up! Let the
whole community today witness that this couple are divorcing!\footnote{Lit., `` O, all you villagers, today they are divorcing.''}

8. All you elders and youngsters\footnote{\textbf{ú-phu} 'elders' (lit. ``white-haired ones''); \textbf{ú-nâʔ} 'youngsters' (lit. ``black-haired ones'').}, all you adolescents and children, please
let them break up!

9. It is dissolved! It is dissolved!

