\setcounter{footnote}{0}

1. Oh! Oh! This man has departed for Pū-phā-ša-phā's country.\footnote{\textbf{Pū-phā-ša-phā}: the boatsman or gatekeeper at the ``shore'' (\textbf{pū}) of G'uisha's heaven. He decides who may enter. Also called simply \textbf{Pū-phā}. \textit{\textbf{See ``Recalling a wandering soul''.}}}

2. Ah! He is separated from his wife and his children and from all his animals;
these two, husband and wife, are now separated, and he is living in Pū-phā-ša-phā's
country.

3. Ah! From today and hereafter let him, and his wife and children not seek one
another.

4. You\footnote{The first part of the chant is addressed to the surviving villagers, but at this juncture the officiating specialist begins to address the deceased's spirit (\textbf{ɔ̀-ha}) directly.} are not the same as you were before\footnote{Lit., ``your appearance is not you''.}, so what (your spirit) craves
is to stay over there in

Pū-phā-ša-phā's country.

5. Now, from today and hereafter, your wife works in the fields like everyone else,
she is looking for a new husband\footnote{This lie is supposed to encourage the deceased's spirit to leave, since he will now want to find a new wife in the land of the dead.}, so let her live in peace.\footnote{\textbf{chɛ̂} \textbf{phɛ̀ʔ} \textbf{pî}: lit., ``make her able to live''. That is, ``Unlike you, she is not a spirit, so do not try to take her with you to the land of the dead.''}

6. You have become a spirit, so when you go to Pū-phā-ša-phā's country, go
carefully, and there carefully look for and marry a new wife.

7. You are not as you were before, so look for and marry a new wife from over there.

8. Your wife and children are here, so let her marry someone from here. She does
not disrespect you, for you must leave now, and over there in Pū-phā-ša-phā's
country you must look for and marry a new wife.

9. You are not as you were before. From today and hereafter, when you leave, go
carefully! From now on, do not cause trouble for your wife and children, or for
your livestock; let them live here in peace.

10. You go and seek [a new family and new livestock] in Pū-phā-ša-phā's country,
Oh go, go carefully!\footnote{The non-3rd person benefactive particle \textbf{lâ} shows that the spirit's departure is for ``our'' benefit, i.e. the benefit of the deceased's family and of the whole village.}

