\setcounter{footnote}{0}

1. Oh today, at this place, I offer you this silver altar, this golden altar\footnote{\textbf{thī-ŋə̂-thī-khâ}: 'silver and gold altar'. A poetic way of referring to the modest construction of bamboo used in this prayer. [See Plate 38 in Matisoff 1988.] The spirits are conceived of as powerful but rather stupid, easily deceived by hyperbolic language.},
you who watch over the seven thousand male and female pigs on the right-hand side,
you who watch over the seven million male and female pigs on the left-hand side,
you rulers, great princes of this place.\footnote{This prayer is directed mainly to the Hill Spirit (\textbf{qhɔ-nê}), the guardian of the game in a certain area, but in order to avoid any problems, it is also addressed to the local Shan legendary princes or deities.}

2. Oh, I beg for male and female pigs, I beseech you for male and female barking
deer.

3. Today I bring for you these offerings, I place them under your feet and under
your hands.\footnote{Lit., ``they have reached your feet and hands''.}

4. Oh, do not block me up with distress! Make them easy for me to shoot! Bind them
with your chains of iron and brass! And when I am stalking around the summit of
a hill, take and put them at the summit of the hill; when I am following a trail
around the foot of a hill, take and put them at the foot of the hill! Cover [their
eyes] with leaves!

5. Oh, one half [of the meat I obtain] I wish to contribute according to the sharing
customs of the elders, I wish to follow the customs of the menfolk and the womenfolk
\footnote{That is, ``Let me shoot large animals, not small ones like squirrels and birds''. Lahu do not customarily divide up small game.}; one half I desire for my wife and children, and for my kinsmen.

6. I pray to you not to make us suffer!\footnote{\textbf{tâ-hā-tâ-g̈ɔ̂} \textbf{pèʔ} \textbf{lâ}: The verb \textbf{pèʔ} 'give; bestow' here functions as a causative post-head auxiliary, like its more common synonym \textbf{pî}.} Please provide sustenance for us!

7. Today grant that the evening star not set, that the moon not set [before I succeed]!
When I am stalking around the summit of the hill, take and put them at the summit
of the hill; when I am following a trail around the foot of a hill, take and put
them at the foot of the hill!

8. We human beings lack your all-hearing ears and all-seeing eyes!.\footnote{Exaggerated humility is one of the strategies commonly used to try and win the spirits' good will.} Within
your feet and hands I beg and beseech you, that you cause us no suffering! Today,
before the evening star has set, before the moon has set, please provide us sustenance!

9. Oh, when you put them over there at the summit of the hill, cover [their eyes]
with leaves! Bind them in your chains of iron and brass, put them right in front
of me and lead them to my big gun of iron and brass! You who have supernatural
powers, I myself have none, but I put my humble request under your feet, under
your hands.

10. Oh, today I offer you this silver altar, this golden altar; I offer you these
silver pendants, these golden pendants.\footnote{These refer to paper streamers of interlocked rings of paper attached to the ``altar''.} I offer you this splendid food and
drink, and I put my prayer under your feet and under your hands.

11. Please provide us sustenance!

12. One half I desire for my wife and children, with one half I wish to follow the
customs of the elders, I wish to contribute according to the customs of the menfolk
and the womenfolk.

13. I pray to you under your feet and hands, for male and female pigs, for male
and female barking deer, enough to divide up and enjoy [8]! Give us sustenance!
Today, before the setting of the evening star and the moon!

14. Oh, please accept [my offerings and my prayers]!

